### 2019-12-16 ms.rmd
initial commit

---
title: The Processing of American Indian/Alaska Native Children through non-Tribal
  Child Welfare Systems
author: "Frank Edwards"
date: "12/12/2019"
output: html_document
---

```{r include=F}
rm(list=ls()); gc()
library(tidyverse)
source("lifetable.r")

pop<-read_csv("./data/pop_aian_nat.csv")
afcars<-read_csv("./data/afcars_subset.csv",
                 col_types = cols(stfcid="c"))

ncands<-read_csv("./data/ncands_subset.csv")

ncands<-ncands %>% 
  mutate(year = year(rptdt)) %>% 
  filter(year>=2003)%>% 
  filter(chage<18)

### can't use FCID because of separate imputation
```

## OUTLINE

this paper examines how american indian children are processed through us child welfare systems. it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated. I compare these pathways through the foster care system to the pathways experienced by children of other racial and ethnic groups.

## What am i estimating

1. Pr(investigation)
2. Pr(substantiated|investigation)
3. Pr(placement|substantiated), Pr(placement|investigation), Pr(placement)
4. Pr(non-indian FC|placement)
5. Pr(termination|placement)

A portrait of system involvement for AI/AN kids:

Lifetime risk of ever:
being investigated
being substantiated
being placed
being placed in non-native home
being placed in long-term care (>1yr)
being terminated

## $Pr(\textsf{investigation})$

Establish the probability that a child was ever investigated.

```{r ncandstables.pop}
ncands_first<-ncands %>% 
  group_by(chid) %>% 
  filter(rptdt==min(rptdt))

ncands_first_nat<-ncands_first %>% 
  group_by(chage, year) %>% 
  summarise(var = n()) %>% 
  rename(age = chage) %>% 
  left_join(pop) %>% 
  ungroup()

### run lifetable by year

investigation_tables<-list()
years<-unique(ncands_first_nat$year)
for(i in 1:length(years)){
  investigation_tables[[i]]<-make_life_table(ncands_first_nat %>% 
                  filter(year==years[i]))
}

investigation_tables<-bind_rows(investigation_tables)
```

Use chid to identify probability of ever foster care for those ever investigated.

Use CHID - match to AFCARS

Get first FC placement - use ncands_first counts as denominator.

could also do matching based on ID

```{r fctables.pop}
### look at first fc placement
afcars_first <- afcars %>% 
  group_by(stfcid) %>% 
  filter(age == min(age) & year == min(year)) %>% 
  select(stfcid, age, year) %>% 
  distinct()
### this matches length(unique(afcars$stfcid))

afcars_first_nat<-afcars_first %>% 
  group_by(age, year) %>% 
  summarise(var = n()) %>% 
  left_join(pop) %>% 
  ungroup()

### run lifetable by year
fc_tables<-list()
years<-unique(afcars_first_nat$year)
for(i in 1:length(years)){
  fc_tables[[i]]<-make_life_table(afcars_first_nat %>% 
                  filter(year==years[i]))
}

fc_tables_pop<-bind_rows(fc_tables)
```

```{r fctables.ncands}

```

Compare to lifetime risk of ever investigated

### 2019-12-16 lifetable.r

@@ -1,0 +1,44 @@
+
+make_life_table<-function(dat){
+  ### create proportions of pop with each outcome by age/year
+  dat<-dat%>%
+    mutate(m = var / pop)
+  
+  ### convert to probability 
+  ### age_period (n) = 1 for all cases
+  ### a = 0.5 (avg age of TPR for ppl in year, within-period survival)
+  dat<-dat%>%
+    mutate(q = 1 * m / (1 + (1 - 0.5) * m),
+           p = 1 - q)
+  ### make cumulative survival
+  dat<-dat%>%
+    mutate(lx = 1e5 * cumprod(c(1, p))[-nrow(dat)])
+  ### deaths
+  dat<-dat%>%
+    mutate(d = -c(diff(lx),0))
+  ## person-years in each group
+  dat<-dat%>%
+    mutate(L = (lx - d) * 1 + d * 0.5,
+           t = sum(L)- cumsum(L) + L)
+  ## life expectancy (time to TPR)
+  dat<-dat%>%
+    mutate(e = t/lx)
+  ### cum prevalence
+  dat<-dat%>%
+    mutate(c = 1-lx/1e5)
+  
+  dat<-dat %>% 
+    mutate(s = 1 - c) %>% 
+    mutate(var = as.numeric(var),
+           pop = as.numeric(pop)) %>% 
+    mutate(se = s * 
+             sqrt(
+               sum(var / 
+                     (pop * (pop - var)))
+             ))
+  
+  dat<-dat %>% 
+    mutate(se = ifelse(age==0, 0, se)) # set se to 0 when age == 0 for variance on s, c=1-s
+  
+  return(dat)
+}

### 2019-12-16 read.r

@@ -1,0 +1,46 @@
+rm(list=ls()); gc()
+library(tidyverse)
+library(lubridate)
+
+pop<-read_fwf("./data/us.1990_2017.singleages.adjusted.txt",
+              fwf_widths(c(4, 2, 2, 3, 2, 1, 1, 1, 2, 8),
+                         c("year", "state", "st_fips",
+                           "cnty_fips", "reg", "race",
+                           "hisp", "sex", "age", "pop")))
+pop<-pop%>%
+  mutate(pop = as.integer(pop))%>%
+  mutate(race_ethn =
+           case_when(
+             race==1 & hisp ==0 ~ "White",
+             race==2 ~ "Black",
+             race==3 ~ "AI/AN",
+             race==4 ~ "Asian/PI",
+             hisp==1 ~ "Hispanic")) %>%
+  mutate(age = as.integer(age)) %>%
+  filter(year>=2000)
+
+afcars<-read_csv("./data/afcars_imputed_all_cases.csv",
+                 col_types = cols(stfcid = "c")) %>% 
+  filter(year>2003) 
+
+ncands<-read_csv("./data/ncands_imputed.csv")
+ncands_xwalk<-read_csv("./data/ncands_xwalk.csv")
+ncands<-ncands %>% 
+  left_join(ncands_xwalk)
+
+pop_aian<-pop %>% 
+  filter(race_ethn=="AI/AN") %>% 
+  group_by(year, age) %>% 
+  summarise(pop = sum(pop)) %>% 
+  write.csv("./data/pop_aian_nat.csv", row.names=F)
+
+ncands_aian<-ncands %>% 
+  filter(race_ethn=="AI/AN") %>% 
+  filter(.imp==1) %>% 
+  write.csv("./data/ncands_subset.csv", row.names=F)
+
+afcars_aian<-afcars %>% 
+  filter(race_ethn=="AI/AN") %>% 
+  filter(.imp==1) %>% 
+  write.csv("./data/afcars_subset.csv", row.names=F)
+  

### 2019-12-17 ms.rmd
update with proability models

@@
 rm(list=ls()); gc()
+### for libra
+setwd("~/Projects/ai_an_transitions")
 library(tidyverse)
+library(lubridate)
 source("lifetable.r")
@@
 afcars<-read_csv("./data/afcars_subset.csv",
-                 col_types = cols(stfcid="c"))
+                 col_types = cols(stfcid="c")) %>% 
+  filter(age<18)
 
@@
 ### can't use FCID because of separate imputation
+
+afcars %>% group_by(age, year) %>% summarise(n=n())
 ```
@@
 
-## $Pr(\textsf{investigation})$
+## $P(\textsf{investigation})$
 
@@
 
-Use chid to identify probability of ever foster care for those ever investigated.
+## $P(substantiation)$
 
-Use CHID - match to AFCARS
+```{r fctables.ncands}
+ncands_first<-ncands %>% 
+  filter(rptvictim==1) %>% 
+  group_by(chid) %>% 
+  filter(rptdt==min(rptdt))
 
-Get first FC placement - use ncands_first counts as denominator.
+ncands_first_subst_nat<-ncands_first %>% 
+  group_by(chage, year) %>% 
+  summarise(var = n()) %>% 
+  rename(age = chage) %>% 
+  left_join(pop) %>% 
+  ungroup()
 
-could also do matching based on ID
+### run lifetable by year
+
+subst_tables<-list()
+years<-unique(ncands_first_subst_nat$year)
+for(i in 1:length(years)){
+  subst_tables[[i]]<-make_life_table(ncands_first_subst_nat %>% 
+                  filter(year==years[i]))
+}
+
+subst_tables<-bind_rows(subst_tables)
+```
+
+## $P(subst|inv)$
+
+of all investigated, how many are ever substantiated?
+
+from lifetables:
+
+Because $P(inv|subst)=1$, and we've converted population incidence into annual age-adjusted probability, the ratio of events is the conditional probability
+
+\[P(subst|inv) = \frac{d_{subst}}{d_{inv}}\]
+
+```{r}
+
+make_conditionals<-function(num, denom){
+  out<-num %>% 
+  select(age, year, d) %>% 
+  rename(d_num = d) %>% 
+  left_join(denom %>% 
+              select(age, year, d) %>% 
+              rename(d_denom = d)) %>% 
+  group_by(year) %>% 
+  summarise(q = sum(d_num)/sum(d_denom))
+  return(out)
+}
+
+sub_inv<-make_conditionals(subst_tables, investigation_tables)
+
+```
+
+## $P(FC)$ using AFCARS
 
@@
 
-fc_tables_pop<-bind_rows(fc_tables)
+fc_tables<-bind_rows(fc_tables)
 ```
 
-```{r fctables.ncands}
+## $P(FC)$ using NCANDS
 
+```{r}
+### look at first fc placement
+ncands_first_fc <- ncands %>% 
+  filter(!(is.na(afcarsid))) %>% 
+  group_by(chid) %>% 
+  filter(rptdt==min(rptdt)) %>% 
+  select(chid, afcarsid, chage, year) %>% 
+  ungroup() %>% 
+  distinct()
+### this matches length(unique(afcars$stfcid))
+
+ncands_fc_nat<-ncands_first_fc %>% 
+  rename(age = chage) %>% 
+  group_by(age, year) %>% 
+  summarise(var = n()) %>% 
+  left_join(pop) %>% 
+  ungroup()
+
+### run lifetable by year
+fc_tables<-list()
+years<-unique(afcars_first_nat$year)
+for(i in 1:length(years)){
+  fc_tables[[i]]<-make_life_table(afcars_first_nat %>% 
+                  filter(year==years[i]))
+}
+
+fc_tables<-bind_rows(fc_tables)
+```
+
+## $P(FC|inv)$
+
+```{r}
+fc_inv<-make_conditionals(fc_tables, investigation_tables)
+fc_sub<-make_conditionals(fc_tables, subst_tables)
 ```
@@
 Compare to lifetime risk of ever investigated
+
+CONDITIONALS ONLY WORK FROM SAME DATA - CANT ASSUME THAT p(b|a)=1 otherwise
+
+need to go back to original afcars - rely on imputed ncands to get uncertainty, link to afcars by id, check year for investigation incident, maybe +/-1 yr for lags in placement starts
+
+want to use the proportions to obtain information about probability. perhaps use a model to populate uncertainty? or construct CI from data - either way should be ok
+
+CONFIRM THAT THEIR IS ONLY ONE FCID PER CHID
+
+```{r}
+
+temp<-ncands %>% 
+  group_by(chid) %>% 
+  summarise(n_fc=n_distinct(afcarsid))
+
+```

### 2019-12-18 ms.rmd
update with proability models
ms update

@@
 rm(list=ls()); gc()
-### for libra
-setwd("~/Projects/ai_an_transitions")
 library(tidyverse)
-library(lubridate)
 source("lifetable.r")
@@
 afcars<-read_csv("./data/afcars_subset.csv",
-                 col_types = cols(stfcid="c")) %>% 
-  filter(age<18)
+                 col_types = cols(stfcid="c"))
+
+afcars_all<-read_csv("./data/afcars_id_xwalk.csv",
+                     col_types = cols(stfcid="c"))
 
@@
 
-ncands<-ncands %>% 
-  mutate(year = year(rptdt)) %>% 
-  filter(year>=2003)%>% 
-  filter(chage<18)
+xwalk<-ncands_xwalk %>% 
+  left_join(afcars_all) %>% 
+  filter(!(is.na(fy))) %>% 
+  select(stfcid, rptdt, fy) %>% 
+  rename(fc_year = fy)
 
-### can't use FCID because of separate imputation
-
-afcars %>% group_by(age, year) %>% summarise(n=n())
 ```
@@
 
-this paper examines how american indian children are processed through us child welfare systems.
+This report examines how american indian children are processed through us child welfare systems.
 it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
-I compare these pathways through the foster care system to the pathways experienced by children of other racial and ethnic groups.
+I compare these pathways through the foster care system to the pathways experienced by white and African American children.
 
-## What am i estimating
+I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster placement, and termination of parental rights.
+Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.
+
+## What am I estimating
 
@@
 ncands_first<-ncands %>% 
+  select(.imp, chid, chage, rptdt) %>% 
   group_by(chid) %>% 
-  filter(rptdt==min(rptdt))
+  filter(rptdt==min(rptdt)) %>% 
+  distinct() %>% 
+  ungroup() %>% 
+  mutate(year = year(rptdt))
 
@@
 investigation_tables<-bind_rows(investigation_tables)
+
+investigation_tables %>% 
+  filter(age==17) %>% 
+  ggplot(aes(x=year, y = c)) +
+  geom_line()
+
+### way lower rates in 17
+### check total case counts
+### counts are low, probably some non-reporting states, removing for now
 ```
@@
 
+## $P(entered, investigation)$ using ncands
+```{r fctables.ncands}
+ncands_fc_first<-ncands %>% 
+  filter(ever_entered==T) %>% 
+  group_by(stfcid) %>% 
+  filter(chage==min(chage)) %>% 
+  ungroup() %>% 
+  select(stfcid, year, chage) %>% 
+  distinct()
+
+ncands_fc_nat<-ncands_fc_first %>% 
+  group_by(year, chage) %>% 
+  summarise(var = n()) %>% 
+  rename(age = chage) %>% 
+  left_join(pop) %>% 
+  ungroup()
+
+### run lifetable by year
+ncands_fc_tables<-list()
+years<-unique(ncands_fc_nat$year)
+for(i in 1:length(years)){
+  ncands_fc_tables[[i]]<-make_life_table(ncands_fc_nat %>% 
+                  filter(year==years[i]))
+}
+
+ncands_fc_tables<-bind_rows(ncands_fc_tables)
+```
+
+## $P(investigation , entered)$ using afcars
+
+```{r}
+### among all entries, match on investigation
+### question is afcars or ncands on left for join
+### likely assymetry
+### need first inv date in ncands_xwalk
+### to ensure I have inv prior to fc
+### to obtain P(inv,entered) where entered occurs after inv
+
+afcars_inv<-afcars_first %>% 
+  left_join(ncands_xwalk %>% 
+              select(stfcid, rptdt) %>% 
+              mutate(ncands_inv = T))
+
+afcars_inv<-afcars_inv %>% 
+  ungroup() %>% 
+  mutate(rpt_year = year(rptdt)) %>% 
+  filter(!(is.na(ncands_inv)))
+
+## remove when year(rptdt)>year
+
+afcars_inv<-afcars_inv %>% 
+  filter(year>=rpt_year)
+
+```
+
+RECALL THAT
+
+\[P(A|B) = \frac{P(A,B)}{P(B)}\]
+
+ALSO, BAYES THEOREM
+
+\[P(B|A) = \frac{P(A|B)P(B)}{P(A)}\]
+
+for $A=FC$ and $B=investigation$, we wish to obtain $P(A|B)$. We obtain differing measurements of $P(A,B)$ from two datasets. I'll calculate each and compare.
+
 ## $P(FC|inv)$
@@
 fc_sub<-make_conditionals(fc_tables, subst_tables)
+fc_ncands_inv<-make_conditionals(ncands_fc_tables, 
+                                 investigation_tables)
+fc_ncands_subst<-make_conditionals(ncands_fc_tables, 
+                                 subst_tables)
 ```
+
+NEXT: compare these to raw rates,
 
@@
 want to use the proportions to obtain information about probability. perhaps use a model to populate uncertainty? or construct CI from data - either way should be ok
-
-CONFIRM THAT THEIR IS ONLY ONE FCID PER CHID
-
-```{r}
-
-temp<-ncands %>% 
-  group_by(chid) %>% 
-  summarise(n_fc=n_distinct(afcarsid))
-
-```

### 2019-12-18 read.r
ff

@@ -25,8 +25,19 @@ afcars<-read_csv("./data/afcars_imputed_all_cases.csv",
 
 ncands<-read_csv("./data/ncands_imputed.csv")
 ncands_xwalk<-read_csv("./data/ncands_xwalk.csv")
+
+library(maps)
+data(state.fips)
+state.fips<-state.fips %>% 
+  select(fips, abb) %>% 
+  distinct() %>% 
+  rename(staterr = abb)
+
+### attach stfcid
 ncands<-ncands %>% 
   left_join(ncands_xwalk)
+
+afcars_xwalk<-read_csv("./data/afcars_xwalk.csv")
 
 pop_aian<-pop %>% 
   filter(race_ethn=="AI/AN") %>% 

### 2019-12-18 ms.rmd
begin writeup

@@
 
-This report examines how american indian children are processed through us child welfare systems.
-it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
+This report examines how American Indian and Alaska Native children are processed through US child welfare systems.
+It explores the likelihood that American Indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
 I compare these pathways through the foster care system to the pathways experienced by white and African American children.
 
-I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster placement, and termination of parental rights.
+I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster care settings, and termination of parental rights.
 Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.
 
-## What am I estimating
+This report pays particular attention to the probability of particular case processing decisions for AI/AN children, conditional on prior decision points.
+In doing so, we can begin to zoom-in on whether inequalities for AI/AN children emerge at particular decision points.
+The first component of the report provides national estimates of the following probabilities
 
-1. Pr(investigation)
-2. Pr(substantiated|investigation)
-3. Pr(placement|substantiated), Pr(placement|investigation), Pr(placement)
-4. Pr(non-indian FC|placement)
-5. Pr(termination|placement)
+1. $Pr(investigation)$
+2. $Pr(substantiated|investigation)$
+3. $Pr(placement)$
+3. $Pr(placement|investigation)$
+4. $Pr(placement|substantiation)$
+4. $Pr(\textsf{ICWA non-compliant placement}|placement)$
+4. $Pr(\textsf{long-term placement}|placement)$
+5. $Pr(termination|placement)$
 
-A portrait of system involvement for AI/AN kids:
+In doing so, this report provides a portrait of child-welfare system involvement for American Indian/Alaska Native children at the national level.
+Future versions of this report will detail sub-national variation.
+Sub-national estimates can be coupled with relevant system-level or child-level data to help evaluate which features of places are associated with investigation, substantiation, placement, and termination decisions for AI/AN children and families.
 
-Lifetime risk of ever:
-being investigated
-being substantiated
-being placed
-being placed in non-native home
-being placed in long-term care (>1yr)
-being terminated
+In each section, I first explore how risk has shifted over time with an exploration of national trends in lifetime risk between 2003 and 2016.
+Next, I provide detailed estimates of risk for the 2012 - 2016 period with an emphasis on both lifetime and age-specific risks.
 
-## $P(\textsf{investigation})$
+In contrast to per capita event rates, life table estimates allow us to simulate the incidence of an event over the course of a child's life.
+These estimates adjust for variable age-specific risk to provide a more comprehensive single estimate of event incidence in the population.
+
+## Investigations
+
+What proportion of AI/AN children are likely to experience a child welfare system investigation during the course of their childhood?
+I examine the child-level probability of ever experiencing an investigation by examining the proportion of children in each year of the data that experience their first maltreatment investigation, then using these numbers to construct period life tables, first as annual tables, and second as a grouped 2012 - 2016 table (which stabilizes trends driven by small age-specific event counts).
+
+$P(\textsf{investigation})$
 
@@
 ```
+
+VISUALIZE investigation tables
 

### 2019-12-18 ms.rmd
add white data

--- 2019-12-18 ms.rmd
@@
 library(tidyverse)
+library(lubridate)
 source("lifetable.r")
 
-pop<-read_csv("./data/pop_aian_nat.csv")
+pop<-read_csv("./data/pop_nat.csv")
 afcars<-read_csv("./data/afcars_subset.csv",
-                 col_types = cols(stfcid="c"))
-
-afcars_all<-read_csv("./data/afcars_id_xwalk.csv",
                  col_types = cols(stfcid="c"))
@@
 
-xwalk<-ncands_xwalk %>% 
-  left_join(afcars_all) %>% 
-  filter(!(is.na(fy))) %>% 
-  select(stfcid, rptdt, fy) %>% 
-  rename(fc_year = fy)
+xwalk<-read_csv("./data/xwalk_merged.csv")
 
@@
 ```{r ncandstables.pop}
+### identify first report
 ncands_first<-ncands %>% 
-  select(.imp, chid, chage, rptdt) %>% 
-  group_by(chid) %>% 
+  select(.imp, race_ethn, stfcid, chage, rptdt) %>% 
+  group_by(stfcid) %>% 
   filter(rptdt==min(rptdt)) %>% 
+  ungroup() %>% 
   distinct() %>% 
-  ungroup() %>% 
   mutate(year = year(rptdt))
@@
 ncands_first_nat<-ncands_first %>% 
-  group_by(chage, year) %>% 
+  group_by(.imp, race_ethn, chage, year) %>% 
   summarise(var = n()) %>% 
+  ungroup() %>% 
   rename(age = chage) %>% 
   left_join(pop) %>% 
-  ungroup()
+  filter(age < 18)
 
@@
   filter(age==17) %>% 
-  ggplot(aes(x=year, y = c)) +
+  ggplot(aes(x=year, y = c, color = race_ethn)) +
   geom_line()

### 2019-12-18 read.r

@@ -2,6 +2,8 @@ rm(list=ls()); gc()
 library(tidyverse)
 library(lubridate)
 
+setwd("~/Projects/ai_an_transitions/")
+
 pop<-read_fwf("./data/us.1990_2017.singleages.adjusted.txt",
               fwf_widths(c(4, 2, 2, 3, 2, 1, 1, 1, 2, 8),
                          c("year", "state", "st_fips",
@@ -20,11 +22,12 @@ pop<-pop%>%
   filter(year>=2000)
 
 afcars<-read_csv("./data/afcars_imputed_all_cases.csv",
-                 col_types = cols(stfcid = "c")) %>% 
-  filter(year>2003) 
+                 col_types = cols(stfcid = "c")) 
 
 ncands<-read_csv("./data/ncands_imputed.csv")
-ncands_xwalk<-read_csv("./data/ncands_xwalk.csv")
+
+ncands_xwalk<-read_csv("./data/ncands_xwalk.csv",
+                       col_types = cols(stfcid = "c"))
 
 library(maps)
 data(state.fips)
@@ -37,21 +40,39 @@ state.fips<-state.fips %>%
 ncands<-ncands %>% 
   left_join(ncands_xwalk)
 
-afcars_xwalk<-read_csv("./data/afcars_xwalk.csv")
+### index files to crosswalk with varying imputations
+# afcars_all<-read_csv("./data/afcars_id_xwalk.csv",
+#                      col_types = cols(stfcid="c"))
+# ncands_xwalk<-read_csv("./data/ncands_xwalk.csv")
+# xwalk<-ncands_xwalk %>% 
+#   left_join(afcars_all) %>% 
+#   filter(fy<=2003, fy<2017) %>% 
+#   filter(!(is.na(fy))) %>% 
+#   select(stfcid, rptdt, fy) %>% 
+#   rename(fc_year = fy)
+# 
+# write_csv(xwalk, "./data/xwalk_merged.csv")
+
+### remove non-matches (prior to 2003, 2017)
+
+ncands<-ncands %>% 
+  filter(!(is.na(subyr)))
 
 pop_aian<-pop %>% 
-  filter(race_ethn=="AI/AN") %>% 
-  group_by(year, age) %>% 
+  filter(race_ethn=="AI/AN" | race_ethn=="White") %>% 
+  group_by(year, age, race_ethn) %>% 
   summarise(pop = sum(pop)) %>% 
[…]

### 2019-12-19 ms.rmd
ff

--- 2019-12-18 ms.rmd
@@
 library(tidyverse)
+library(lubridate)
 source("lifetable.r")
-
-pop<-read_csv("./data/pop_aian_nat.csv")
-afcars<-read_csv("./data/afcars_subset.csv",
-                 col_types = cols(stfcid="c"))
-
-afcars_all<-read_csv("./data/afcars_id_xwalk.csv",
-                     col_types = cols(stfcid="c"))
-
-ncands<-read_csv("./data/ncands_subset.csv")
-
-xwalk<-ncands_xwalk %>% 
-  left_join(afcars_all) %>% 
-  filter(!(is.na(fy))) %>% 
-  select(stfcid, rptdt, fy) %>% 
-  rename(fc_year = fy)
 
@@
 In contrast to per capita event rates, life table estimates allow us to simulate the incidence of an event over the course of a child's life.
-These estimates adjust for variable age-specific risk to provide a more comprehensive single estimate of event incidence in the population.
+These estimates adjust for variable age-specific risk and population size to provide a more comprehensive single estimate of event incidence in the population.
 
@@
 
-$P(\textsf{investigation})$
-
-Establish the probability that a child was ever investigated.
-
 ```{r ncandstables.pop}
-ncands_first<-ncands %>% 
-  select(.imp, chid, chage, rptdt) %>% 
-  group_by(chid) %>% 
-  filter(rptdt==min(rptdt)) %>% 
-  distinct() %>% 
-  ungroup() %>% 
-  mutate(year = year(rptdt))
-
-ncands_first_nat<-ncands_first %>% 
-  group_by(chage, year) %>% 
-  summarise(var = n()) %>% 
-  rename(age = chage) %>% 
-  left_join(pop) %>% 
-  ungroup()
-
+### read in new first investigation csv
 ### run lifetable by year
@@
 years<-unique(ncands_first_nat$year)
+race_ethn<-unique(ncands_first$race_ethn)
 for(i in 1:length(years)){
+  for(r in 1:length(race_ethn)){
   investigation_tables[[i]]<-make_life_table(ncands_first_nat %>% 
-                  filter(year==years[i]))
+                  filter(year==years[i]) %>% 
+                    filter(race_ethn == race_ethn[r]))
+  }
 }
@@
 
+FIG 1: TS plot of age 17 cumulative investigation risk by year
+
 VISUALIZE investigation tables
@@
 ```{r fctables.ncands}
-ncands_first<-ncands %>% 
-  filter(rptvictim==1) %>% 
-  group_by(chid) %>% 
-  filter(rptdt==min(rptdt))
-
-ncands_first_subst_nat<-ncands_first %>% 
-  group_by(chage, year) %>% 
-  summarise(var = n()) %>% 
-  rename(age = chage) %>% 
-  left_join(pop) %>% 
-  ungroup()
 
@@
 ```{r fctables.pop}
-### look at first fc placement
-afcars_first <- afcars %>% 
-  group_by(stfcid) %>% 
-  filter(age == min(age) & year == min(year)) %>% 
-  select(stfcid, age, year) %>% 
-  distinct()
-### this matches length(unique(afcars$stfcid))
-
-afcars_first_nat<-afcars_first %>% 
-  group_by(age, year) %>% 
-  summarise(var = n()) %>% 
-  left_join(pop) %>% 
-  ungroup()
 
@@
 
-## $P(FC)$ using NCANDS
-
-```{r}
-### look at first fc placement
-ncands_first_fc <- ncands %>% 
-  filter(!(is.na(afcarsid))) %>% 
-  group_by(chid) %>% 
-  filter(rptdt==min(rptdt)) %>% 
-  select(chid, afcarsid, chage, year) %>% 
-  ungroup() %>% 
-  distinct()
-### this matches length(unique(afcars$stfcid))
-
-ncands_fc_nat<-ncands_first_fc %>% 
-  rename(age = chage) %>% 
-  group_by(age, year) %>% 
-  summarise(var = n()) %>% 
-  left_join(pop) %>% 
-  ungroup()
-
-### run lifetable by year
-fc_tables<-list()
-years<-unique(afcars_first_nat$year)
-for(i in 1:length(years)){
-  fc_tables[[i]]<-make_life_table(afcars_first_nat %>% 
-                  filter(year==years[i]))
-}
-
-fc_tables<-bind_rows(fc_tables)
-```
-
-## $P(entered, investigation)$ using ncands
-```{r fctables.ncands}
-ncands_fc_first<-ncands %>% 
-  filter(ever_entered==T) %>% 
-  group_by(stfcid) %>% 
-  filter(chage==min(chage)) %>% 
-  ungroup() %>% 
-  select(stfcid, year, chage) %>% 
-  distinct()
-
-ncands_fc_nat<-ncands_fc_first %>% 
-  group_by(year, chage) %>% 
-  summarise(var = n()) %>% 
-  rename(age = chage) %>% 
-  left_join(pop) %>% 
-  ungroup()
-
-### run lifetable by year
-ncands_fc_tables<-list()
-years<-unique(ncands_fc_nat$year)
-for(i in 1:length(years)){
-  ncands_fc_tables[[i]]<-make_life_table(ncands_fc_nat %>% 
-                  filter(year==years[i]))
-}
-
-ncands_fc_tables<-bind_rows(ncands_fc_tables)
-```
-
 ## $P(investigation , entered)$ using afcars
@@
 
-afcars_inv<-afcars_first %>% 
-  left_join(ncands_xwalk %>% 
-              select(stfcid, rptdt) %>% 
-              mutate(ncands_inv = T))
-
-afcars_inv<-afcars_inv %>% 
-  ungroup() %>% 
-  mutate(rpt_year = year(rptdt)) %>% 
-  filter(!(is.na(ncands_inv)))
-
-## remove when year(rptdt)>year
-
-afcars_inv<-afcars_inv %>% 
-  filter(year>=rpt_year)
+### lifetable for p(inv,entered)
 

### 2019-12-19 read.r

@@ -21,6 +21,10 @@ pop<-pop%>%
   mutate(age = as.integer(age)) %>%
   filter(year>=2000)
 
+pop<-pop %>% 
+  group_by(age, race_ethn) %>% 
+  summarise(pop = sum(pop))
+
 afcars<-read_csv("./data/afcars_imputed_all_cases.csv",
                  col_types = cols(stfcid = "c")) 
 
@@ -28,6 +32,7 @@ ncands<-read_csv("./data/ncands_imputed.csv")
 
 ncands_xwalk<-read_csv("./data/ncands_xwalk.csv",
                        col_types = cols(stfcid = "c"))
+sink("./log.txt")
 
 library(maps)
 data(state.fips)
@@ -40,39 +45,132 @@ state.fips<-state.fips %>%
 ncands<-ncands %>% 
   left_join(ncands_xwalk)
 
-### index files to crosswalk with varying imputations
-# afcars_all<-read_csv("./data/afcars_id_xwalk.csv",
-#                      col_types = cols(stfcid="c"))
-# ncands_xwalk<-read_csv("./data/ncands_xwalk.csv")
-# xwalk<-ncands_xwalk %>% 
-#   left_join(afcars_all) %>% 
-#   filter(fy<=2003, fy<2017) %>% 
-#   filter(!(is.na(fy))) %>% 
-#   select(stfcid, rptdt, fy) %>% 
-#   rename(fc_year = fy)
-# 
-# write_csv(xwalk, "./data/xwalk_merged.csv")
-
-### remove non-matches (prior to 2003, 2017)
+### get needed first events
 
 ncands<-ncands %>% 
   filter(!(is.na(subyr)))
 
-pop_aian<-pop %>% 
-  filter(race_ethn=="AI/AN" | race_ethn=="White") %>% 
-  group_by(year, age, race_ethn) %>% 
-  summarise(pop = sum(pop)) %>% 
-  write.csv("./data/pop_nat.csv", row.names=F)
+ncands_first<-ncands %>% 
+  select(.imp, chid, race_ethn, chage, rptdt) %>% 
+  group_by(chid) %>% 
+  filter(rptdt==min(rptdt)) %>% 
+  distinct() %>% 
+  ungroup() %>% 
+  mutate(year = year(rptdt))
 
-ncands_aian<-ncands %>% 
[…]

### 2019-12-19 read_ncands.r

@@ -1,0 +1,59 @@
+rm(list=ls()); gc()
+library(data.table)
+library(tidyverse)
+library(lubridate)
+library(mice)
+
+ncands_path<-"~/Projects/ndacan_data/ncands/"
+ncands_files<-paste(ncands_path,
+                    list.files(ncands_path),
+                    sep = "")[1:16]
+
+ncands<-lapply(ncands_files, fread)
+
+for(i in 1:length(ncands_files)){
+  ncands[[i]]<-ncands[[i]]%>%
+    rename_all(tolower) %>%
+    mutate(rptdt = ymd(rptdt)) %>%
+    mutate(race_ethn =
+             ifelse(chracbl==1,
+                    "Black",
+                    ifelse(chracai==1, "AI/AN",
+                           ifelse(chracas==1 | chracnh==1,
+                                  "Asian/PI",
+                                  ifelse(cethn==1, "Hispanic",
+                                          ifelse(chracwh == 1, "White",
+                                                 NA)))))) 
+  }
+
+##### make an FCID chid xwalk file
+ncands_xwalk<-list()
+for(i in 1:length(ncands_files)){
+  temp<-ncands[[i]] %>% 
+    select(chid, staterr, 
+           afcarsid, rptdt, subyr) 
+  ncands_xwalk[[i]]<-temp
+}
+
+ncands_xwalk<-bind_rows(ncands_xwalk) 
+
+library(maps)
+data(state.fips)
+state.fips<-state.fips %>% 
+  select(fips, abb) %>% 
+  distinct() %>% 
+  rename(staterr = abb)
+
+ncands_xwalk<-ncands_xwalk %>% 
+  left_join(state.fips) %>% 
+  filter(afcarsid!="",
+         staterr!="XX") %>% 
+  mutate(stfcid = paste(fips, afcarsid, sep="")) 
+
+ncands_xwalk<-ncands_xwalk %>% 
+  select(chid, stfcid, rptdt, subyr)
+
+write_csv(ncands_xwalk, 
+          "~/Projects/ai_an_transitions/data/ncands_xwalk.csv")
+
+q("no")

### 2019-12-19 ms.rmd
update

@@
 ## OUTLINE
+
+This report examines how american indian children are processed through us child welfare systems.
+it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
+I compare these pathways through the foster care system to the pathways experienced by white and African American children.
+
+I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster placement, and termination of parental rights.
+Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.
+
+## What am I estimating
 

### 2019-12-21 ms.rmd
add national data

--- 2019-12-19 ms.rmd
@@
 
+ncands_inv<-read_csv("./data/ncands_first_inv.csv") %>% 
+  filter(year>=2003, year<2017) %>% 
+  filter(age<=18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+ncands_sub<-read_csv("./data/ncands_first_sub.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<=18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_fc<-read_csv("./data/afcars_first_fc.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<=18) %>%   
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_tpr<-read_csv("./data/afcars_tpr.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<=18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_non_icwa<-read_csv("./data/afcars_non_icwa.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<=18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_fc_inv<-read_csv("./data/afcars_fc_post_inv.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<=18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
 ```
@@
 investigation_tables<-list()
-years<-unique(ncands_first_nat$year)
-race_ethn<-unique(ncands_first$race_ethn)
-for(i in 1:length(years)){
+years<-unique(ncands_inv$year)
+race_ethn<-unique(ncands_inv$race_ethn)
+index<-1
+for(i in 1:5){
+  for(j in 1:length(years)){
     for(r in 1:length(race_ethn)){
-  investigation_tables[[i]]<-make_life_table(ncands_first_nat %>% 
-                  filter(year==years[i]) %>% 
-                    filter(race_ethn == race_ethn[r]))
+      temp<-ncands_inv %>% 
+        filter(year==years[j])
+      temp<-temp %>% 
+        filter(race_ethn == race_ethn[r])
+      temp<-temp %>% 
+        filter(.imp==i)
+      investigation_tables[[index]]<-make_life_table(temp)
+      index<-index+1
     }
   }
-
+}
 investigation_tables<-bind_rows(investigation_tables)
 
-investigation_tables %>% 
-  filter(age==17) %>% 
-  ggplot(aes(x=year, y = c, color = race_ethn)) +
-  geom_line()
+investigation_tables_c<-investigation_tables %>% 
+  filter(age==18) %>% 
+  group_by(year, race_ethn) %>% 
+  summarise(c = mean(c), cmax = max(c), cmin = min(c))
+
+investigation_tables_c %>% 
+  ggplot(aes(x=year, y = c, color = race_ethn, group = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_line() +
+  geom_ribbon()
 

### 2019-12-21 ms.rmd
git figures started, check on p(fc, inv) > p(fc) cases

--- 2019-12-19 ms.rmd
@@
 
+ncands_inv<-read_csv("./data/ncands_first_inv.csv") %>% 
+  filter(year>=2003, year<2017) %>% 
+  filter(age<18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+ncands_sub<-read_csv("./data/ncands_first_sub.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_fc<-read_csv("./data/afcars_first_fc.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<18) %>%   
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_tpr<-read_csv("./data/afcars_tpr.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_non_icwa<-read_csv("./data/afcars_non_icwa.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_fc_inv<-read_csv("./data/afcars_fc_post_inv.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
 ```
@@
 investigation_tables<-list()
-years<-unique(ncands_first_nat$year)
-race_ethn<-unique(ncands_first$race_ethn)
-for(i in 1:length(years)){
+years<-unique(ncands_inv$year)
+race_ethn<-unique(ncands_inv$race_ethn)
+index<-1
+for(i in 1:5){
+  for(j in 1:length(years)){
     for(r in 1:length(race_ethn)){
-  investigation_tables[[i]]<-make_life_table(ncands_first_nat %>% 
-                  filter(year==years[i]) %>% 
-                    filter(race_ethn == race_ethn[r]))
+      temp<-ncands_inv %>% 
+        filter(year==years[j])
+      temp<-temp %>% 
+        filter(race_ethn == race_ethn[r])
+      temp<-temp %>% 
+        filter(.imp==i)
+      investigation_tables[[index]]<-make_life_table(temp)
+      index<-index+1
     }
   }
-
+}
 investigation_tables<-bind_rows(investigation_tables)
 
-investigation_tables %>% 
+investigation_tables_c<-investigation_tables %>% 
   filter(age==17) %>% 
-  ggplot(aes(x=year, y = c, color = race_ethn)) +
-  geom_line()
+  group_by(year, race_ethn) %>% 
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
+
+investigation_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             fill = race_ethn,
+             group = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_ribbon(alpha = .5) + 
+  geom_line(aes(color = race_ethn)) + 
+  xlab("Year") + 
+  ylab("Proportion with screened-in report before age 18")
 
@@
 ### run lifetable by year
+subst_tables<-list()
+years<-unique(ncands_sub$year)
+race_ethn<-unique(ncands_sub$race_ethn)
+index<-1
+for(i in 1:5){
+  for(j in 1:length(years)){
+    for(r in 1:length(race_ethn)){
+      temp<-ncands_sub %>% 
+        filter(year==years[j])
+      temp<-temp %>% 
+        filter(race_ethn == race_ethn[r])
+      temp<-temp %>% 
+        filter(.imp==i)
+      subst_tables[[index]]<-make_life_table(temp)
+      index<-index+1
+    }
+  }
+}
+subst_tables<-bind_rows(subst_tables)
 
-subst_tables<-list()
-years<-unique(ncands_first_subst_nat$year)
-for(i in 1:length(years)){
-  subst_tables[[i]]<-make_life_table(ncands_first_subst_nat %>% 
-                  filter(year==years[i]))
-}
+subst_tables_c<-subst_tables %>% 
+  filter(age==17) %>% 
+  group_by(year, race_ethn) %>% 
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
 
-subst_tables<-bind_rows(subst_tables)
+subst_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             fill = race_ethn,
+             group = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_ribbon(alpha = .5) + 
+  geom_line(aes(color = race_ethn)) + 
+  xlab("Year") + 
+  ylab("Proportion with substantiation before age 18")
 ```
@@
 ```{r fctables.pop}
+fc_tables<-list()
+years<-unique(afcars_fc$year)
+race_ethn<-unique(afcars_fc$race_ethn)
+index<-1
+for(i in 1:5){
+  for(j in 1:length(years)){
+    for(r in 1:length(race_ethn)){
+      temp<-afcars_fc %>% 
+        filter(year==years[j])
+      temp<-temp %>% 
+        filter(race_ethn == race_ethn[r])
+      temp<-temp %>% 
+        filter(.imp==i)
+      fc_tables[[index]]<-make_life_table(temp)
+      index<-index+1
+    }
+  }
+}
+fc_tables<-bind_rows(fc_tables)
 
-### run lifetable by year
-fc_tables<-list()
-years<-unique(afcars_first_nat$year)
-for(i in 1:length(years)){
-  fc_tables[[i]]<-make_life_table(afcars_first_nat %>% 
-                  filter(year==years[i]))
-}
+fc_tables_c<-fc_tables %>% 
+  filter(age==17) %>% 
+  group_by(year, race_ethn) %>% 
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
 
-fc_tables<-bind_rows(fc_tables)
+fc_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             fill = race_ethn,
+             group = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_ribbon(alpha = .5) + 
+  geom_line(aes(color = race_ethn)) + 
+  xlab("Year") + 
+  ylab("Proportion ever in foster care before age 18")
 ```
@@
 ```{r}
-### among all entries, match on investigation
-### question is afcars or ncands on left for join
-### likely assymetry
-### need first inv date in ncands_xwalk
-### to ensure I have inv prior to fc
-### to obtain P(inv,entered) where entered occurs after inv
+fc_post_inv_tables<-list()
+years<-unique(afcars_fc_inv$year)
+race_ethn<-unique(afcars_fc_inv$race_ethn)
+index<-1
+for(i in 1:5){
+  for(j in 1:length(years)){
+    for(r in 1:length(race_ethn)){
+      temp<-afcars_fc_inv %>% 
+        filter(year==years[j])
+      temp<-temp %>% 
+        filter(race_ethn == race_ethn[r])
+      temp<-temp %>% 
+        filter(.imp==i)
+      fc_post_inv_tables[[index]]<-make_life_table(temp)
+      index<-index+1
+    }
+  }
+}
+fc_post_inv_tables<-bind_rows(fc_post_inv_tables)
 
-### lifetable for p(inv,entered)
+fc_post_inv_tables_c<-fc_post_inv_tables %>% 
+  filter(age==17) %>% 
+  group_by(year, race_ethn) %>% 
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
+
+fc_post_inv_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             fill = race_ethn,
+             group = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_ribbon(alpha = .5) + 
+  geom_line(aes(color = race_ethn)) + 
+  xlab("Year") + 
+  ylab("Proportion with foster care entry after screened-in report")
 

### 2019-12-22 read.r
ff

@@ -28,11 +28,11 @@ pop<-pop %>%
 afcars<-read_csv("./data/afcars_imputed_all_cases.csv",
                  col_types = cols(stfcid = "c")) 
 
-ncands<-read_csv("./data/ncands_imputed.csv")
+# ncands<-read_csv("./data/ncands_imputed.csv")
 
 ncands_xwalk<-read_csv("./data/ncands_xwalk.csv",
                        col_types = cols(stfcid = "c"))
-
+  
 library(maps)
 data(state.fips)
 state.fips<-state.fips %>% 
@@ -40,52 +40,58 @@ state.fips<-state.fips %>%
   distinct() %>% 
   rename(staterr = abb)
 
-### attach stfcid
-ncands<-ncands %>% 
-  left_join(ncands_xwalk)
-
-### get needed first events
-
-ncands<-ncands %>% 
-  filter(!(is.na(subyr)))
-
-ncands_first<-ncands %>% 
-  select(.imp, chid, race_ethn, chage, rptdt) %>% 
-  group_by(chid) %>% 
-  filter(rptdt==min(rptdt)) %>% 
-  distinct() %>% 
-  ungroup() %>% 
-  mutate(year = year(rptdt))
-
-ncands_first_nat<-ncands_first %>% 
-  group_by(.imp, chage, race_ethn, year) %>% 
-  summarise(var = n()) %>% 
-  rename(age = chage) %>% 
-  left_join(pop) %>% 
-  ungroup() %>% 
-  write_csv("./data/ncands_first_inv.csv")
-
-ncands_first_subst<-ncands %>% 
-  filter(rptvictim==1) %>% 
-  group_by(chid) %>% 
-  filter(rptdt==min(rptdt))%>% 
-  distinct() %>% 
-  ungroup() %>% 
-  mutate(year = year(rptdt))
-
-ncands_first_subst_nat<-ncands_first_subst %>% 
-  group_by(.imp, chage, race_ethn, year) %>% 
-  summarise(var = n()) %>% 
-  rename(age = chage) %>% 
-  left_join(pop) %>% 
-  ungroup() %>% 
-  write_csv("./data/ncands_first_sub.csv")
+# ### attach stfcid
+# ncands<-ncands %>% 
+#   left_join(ncands_xwalk)
+# 
+# ### get needed first events
+# 
+# ncands<-ncands %>% 
[…]

### 2019-12-22 read.r
fix diagnostic logging

@@ -88,9 +88,9 @@ afcars_first <- afcars %>%
 
 sink("log.txt")
 print("afcars_first length")
-nrow(afcars_first)
+print(nrow(afcars_first))
 print("unique fcid")
-length(unique(afcars$stfcid))
+print(length(unique(afcars$stfcid)))
 sink()
 
 afcars_first_nat<-afcars_first %>% 
@@ -110,9 +110,9 @@ afcars_inv<-afcars_first %>%
 
 sink("log2.txt")
 print("afcars_inv post join length")
-nrow(afcars_inv)
+print(nrow(afcars_inv))
 print("unique ids")
-length(unique(afcars_inv$stfcid))
+print(length(unique(afcars_inv$stfcid)))
 sink()
 
 afcars_inv<-afcars_inv %>% 

### 2019-12-22 read.r
ff

@@ -28,7 +28,7 @@ pop<-pop %>%
 afcars<-read_csv("./data/afcars_imputed_all_cases.csv",
                  col_types = cols(stfcid = "c")) 
 
-# ncands<-read_csv("./data/ncands_imputed.csv")
+ncands<-read_csv("./data/ncands_imputed.csv")
 
 ncands_xwalk<-read_csv("./data/ncands_xwalk.csv",
                        col_types = cols(stfcid = "c"))
@@ -40,51 +40,55 @@ state.fips<-state.fips %>%
   distinct() %>% 
   rename(staterr = abb)
 
-# ### attach stfcid
-# ncands<-ncands %>% 
-#   left_join(ncands_xwalk)
-# 
-# ### get needed first events
-# 
-# ncands<-ncands %>% 
-#   filter(!(is.na(subyr)))
-# 
-# ncands_first<-ncands %>% 
-#   select(.imp, chid, race_ethn, chage, rptdt) %>% 
-#   group_by(chid) %>% 
-#   filter(rptdt==min(rptdt)) %>% 
-#   distinct() %>% 
-#   ungroup() %>% 
-#   mutate(year = year(rptdt))
-# 
-# ncands_first_nat<-ncands_first %>% 
-#   group_by(.imp, chage, race_ethn, year) %>% 
-#   summarise(var = n()) %>% 
-#   rename(age = chage) %>% 
-#   left_join(pop) %>% 
-#   ungroup() %>% 
-#   write_csv("./data/ncands_first_inv.csv")
-# 
-# ncands_first_subst<-ncands %>% 
-#   filter(rptvictim==1) %>% 
-#   group_by(chid) %>% 
-#   filter(rptdt==min(rptdt))%>% 
-#   distinct() %>% 
-#   ungroup() %>% 
-#   mutate(year = year(rptdt))
-# 
-# ncands_first_subst_nat<-ncands_first_subst %>% 
-#   group_by(.imp, chage, race_ethn, year) %>% 
-#   summarise(var = n()) %>% 
-#   rename(age = chage) %>% 
-#   left_join(pop) %>% 
-#   ungroup() %>% 
-#   write_csv("./data/ncands_first_sub.csv")
+zz <- file("all.Rout", open = "wt")
+sink(zz)
+sink(zz, type = "message")
+### attach stfcid
+ncands<-ncands %>%
[…]

### 2019-12-30 ms.rmd
update natl files

@@
 
-```{r fctables.ncands}
+```{r}
 
@@
 
-## $P(subst|inv)$
-
-of all investigated, how many are ever substantiated?
-
-from lifetables:
-
-Because $P(inv|subst)=1$, and we've converted population incidence into annual age-adjusted probability, the ratio of events is the conditional probability
-
-\[P(subst|inv) = \frac{d_{subst}}{d_{inv}}\]
-
-```{r}
-
-make_conditionals<-function(num, denom){
-  out<-num %>% 
-  select(age, year, d) %>% 
-  rename(d_num = d) %>% 
-  left_join(denom %>% 
-              select(age, year, d) %>% 
-              rename(d_denom = d)) %>% 
-  group_by(year) %>% 
-  summarise(q = sum(d_num)/sum(d_denom))
-  return(out)
-}
-
-sub_inv<-make_conditionals(subst_tables, investigation_tables)
-
-```
-
 ## $P(FC)$ using AFCARS
@@
 index<-1
-for(i in 1:5){
+for(i in 1:length(afcars_fc$.imp)){
   for(j in 1:length(years)){
@@
 index<-1
-for(i in 1:5){
+for(i in 1:length(afcars_fc_inv$.imp)){
   for(j in 1:length(years)){
@@
 
+## p(tpr)
+
+```{r}
+tpr_tables<-list()
+years<-unique(afcars_tpr$year)
+race_ethn<-unique(afcars_tpr$race_ethn)
+index<-1
+for(i in 1:length(afcars_tpr$.imp)){
+  for(j in 1:length(years)){
+    for(r in 1:length(race_ethn)){
+      temp<-afcars_tpr %>% 
+        filter(year==years[j])
+      temp<-temp %>% 
+        filter(race_ethn == race_ethn[r])
+      temp<-temp %>% 
+        filter(.imp==i)
+      tpr_tables[[index]]<-make_life_table(temp)
+      index<-index+1
+    }
+  }
+}
+tpr_tables<-bind_rows(tpr_tables)
+
+tpr_tables_c<-tpr_tables %>% 
+  filter(age==17) %>% 
+  group_by(year, race_ethn) %>% 
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
+
+tpr_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             fill = race_ethn,
+             group = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_ribbon(alpha = .5) + 
+  geom_line(aes(color = race_ethn)) + 
+  xlab("Year") + 
+  ylab("Proportion with parental rights terminated")
+```
+
+## P(sub|inv)
+
+```{r}
+
+## p(sub|inv) = p(inv|sub) * p(sub) / p(inv) =
+## p(sub) / p(inv)
+ncands_sub_inv<-subst_tables %>% 
+  mutate(sub_inv = q / investigation_tables$q)
+
+ggplot(ncands_sub_inv %>% 
+         filter(year==2016),
+         aes(x = age, y = sub_inv, color = race_ethn)) + 
+  geom_line()
+```
+
 ## $P(FC|inv)$
 
+\[p(fc|inv) = \frac{p(fc, inv)}{p(inv)}\]
+
 ```{r}
-fc_inv<-make_conditionals(fc_tables, investigation_tables)
-fc_sub<-make_conditionals(fc_tables, subst_tables)
-fc_ncands_inv<-make_conditionals(ncands_fc_tables, 
-                                 investigation_tables)
-fc_ncands_subst<-make_conditionals(ncands_fc_tables, 
-                                 subst_tables)
+fc_cond_inv<-fc_post_inv_tables %>% 
+  mutate(fc_inv = q / investigation_tables$q)
+
+ggplot(fc_cond_inv %>% 
+         filter(year==2016),
+         aes(x = age, y = fc_inv, color = race_ethn)) + 
+  geom_line()
+```
+
+## P(FC|sub)
+```{r}
+
+fc_cond_sub<-fc_post_inv_tables %>% 
+  mutate(fc_sub = q / subst_tables$q)
+
+ggplot(fc_cond_sub %>% 
+         filter(year==2016),
+         aes(x = age, y = fc_sub, color = race_ethn)) + 
+  geom_line()
+```
+
+## P(FC|tpr)
+```{r}
+### use bayes' rule because p(fc|tpr)=1
+# \[P(B|A) = \frac{P(A|B)P(B)}{P(A)}\]
+## P(tpr|fc) = p(fc|tpr)p(tpr) / p(fc)
+## p(tpr|fc) = p(tpr)/p(fc)
+fc_tpr_fc<-tpr_tables %>% 
+  mutate(tpr_fc = q/ fc_tables$q)
+
+ggplot(fc_tpr_fc %>% 
+         filter(year==2016),
+         aes(x = age, y = tpr_fc, color = race_ethn)) + 
+  geom_line()
+
 ```

### 2019-12-30 ms.rmd
report draft in progress, add p(afcars,subst)

--- 2019-12-21 ms.rmd
@@
 ---
-title: The Processing of American Indian/Alaska Native Children through non-Tribal
-  Child Welfare Systems
+title: The prevalence of child welfare system contact among American Indian / Alaska Native children, 2003 - 2016
+subtitle: Report prepared for Casey Family Programs
 author: "Frank Edwards"
-date: "12/12/2019"
-output: html_document
+date: "Rutgers University - Newark"
+output: 
+  bookdown::pdf_document2
+bibliography: library.bib
 ---
@@
 rm(list=ls()); gc()
+knitr::opts_chunk$set(echo = F, warning = F, message = F)
 library(tidyverse)
 library(lubridate)
+library(gridExtra)
 source("lifetable.r")
+theme_set(theme_bw())
 
@@
 
-## OUTLINE
+# Executive summary {-#executive-summary}
 
-This report examines how american indian children are processed through us child welfare systems.
+This report examines how american indian children are processed through
+us child welfare systems.
 it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
-I compare these pathways through the foster care system to the pathways experienced by white and African American children.
+I compare these pathways through the foster care system to the pathways experienced by white and children.
 
-I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster placement, and termination of parental rights.
-Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.
+**FINDINGS HERE**
 
-## What am I estimating
+# Introduction
 
-This report examines how American Indian and Alaska Native children are processed through US child welfare systems.
-It explores the likelihood that American Indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
-I compare these pathways through the foster care system to the pathways experienced by white and African American children.
+American Indian / Alaska Native (AI/AN) children and families are far more likely to experience contact with the child welfare system than are their white peers *(CITES)*. *STATEMENT ON SIGNIFICANCE*
+
+This report provides national estimates of the prevalence of a series of child welfare system contact outcomes for AI/AN children, in addition to estimating the likelihood that children transition to more serious child welfare system outcomes conditional on lower-level contacts, such as investigations.
 
@@
 
-This report pays particular attention to the probability of particular case processing decisions for AI/AN children, conditional on prior decision points.
-In doing so, we can begin to zoom-in on whether inequalities for AI/AN children emerge at particular decision points.
-The first component of the report provides national estimates of the following probabilities
+For the national population of both AI/AN and white children, I estimate the following probabilities of event occurence by age, and as a cumulative likelihood of event occurrence before a child's 18th birthday:
 
-1. $Pr(investigation)$
-2. $Pr(substantiated|investigation)$
-3. $Pr(placement)$
-3. $Pr(placement|investigation)$
-4. $Pr(placement|substantiation)$
-4. $Pr(\textsf{ICWA non-compliant placement}|placement)$
-4. $Pr(\textsf{long-term placement}|placement)$
-5. $Pr(termination|placement)$
+1. Screened-in maltreatment report
+2. Confirmed or substantiated maltreatment report
+3. Foster care placement
+3. Placement in foster home that is not compliant with ICWA guidelines (for AI/AN children)
+4. Termination of both parent's rights while in foster care
 
-In doing so, this report provides a portrait of child-welfare system involvement for American Indian/Alaska Native children at the national level.
-Future versions of this report will detail sub-national variation.
-Sub-national estimates can be coupled with relevant system-level or child-level data to help evaluate which features of places are associated with investigation, substantiation, placement, and termination decisions for AI/AN children and families.
+Inequalities in system outcomes for AI/AN children could emerge at multiple stages of case processing.
+To evaluate how each stage of case processing may contribute to inequality, I explore the probability of upstream system decisions, conditional on the occurrence of earlier decisions.
+I estimate the following conditional probabilities by child age for both AI/AN and white children:
 
-In each section, I first explore how risk has shifted over time with an exploration of national trends in lifetime risk between 2003 and 2016.
-Next, I provide detailed estimates of risk for the 2012 - 2016 period with an emphasis on both lifetime and age-specific risks.
+1. Substantiation, conditional on screened-in report
+2. Foster care placement, conditional on screened-in report
+3. Foster care placement, conditional on substantiated report
+4. ICWA non-compliant placement, conditional on any foster care placement
+5. Parental rights termination, conditional on any foster care placmeent
 
-In contrast to per capita event rates, life table estimates allow us to simulate the incidence of an event over the course of a child's life.
-These estimates adjust for variable age-specific risk and population size to provide a more comprehensive single estimate of event incidence in the population.
+In doing so, this report provides a portrait of child-welfare system involvement for American Indian/Alaska Native children at the national level, and draws attention to the stages of case processing during which inequalities emerge.
 
-## Investigations
+Because federal data systems do not currently include information from tribal child welfare systems, estimates for AI/AN should be interpreted as conservative.
+All estimates presented below are for contact with non-tribal state and local child welfare agencies.
+Also, note that this analysis does not adjust for case-level characteristics.
+It is likely that child-level, family-level, caseworker-level, agency-level, and state-level features all affect these probabilities.
+
+# Data
+
+This analysis relies on data from the National Child Abuse and Neglect Data System (NCANDS) and the Adoption and Foster Care Analysis and Reporting System (AFCARS), using data files from 2003 - 2017.
+I adjust all data to the calendar year beginning January 1 and ending December 31.
+Because NCANDS and AFCARS reporting years span October 1 through September 30 of each year, this report does not provide estimates for 2017 probabilities, as data from October, November and December 2017 had not yet been released as of the date of this report.
+
+NCANDS provides data on each screened-in report of child abuse and neglect.
+I use a unique child identifier (harmonized across NCANDS and AFCARS) to identify the first time a child was the subject of a maltreatment investigation, and the first time a child was the subject of a substantiated or confirmed maltreatment investigation across the 15 years of maltreatment data included in this analysis.
+Some states did not begin reporting data to NCANDS until later in this time period.
+All population figures are adjusted to account for the variable participation of state in NCANDS data reporting.
+
+All children placed by non-Tribal child welfare agencies into out-of-home foster care are documented annually in AFCARS.
+Using a unique child identifier, I record the first year in which a child a) entered foster care, b) was in an ICWA non-compliant foster placement, and c) had the rights of both parents terminated while in foster care.
+Because AFCARS only records the latest placement during a reporting year, and a child may be subject to multiple placements during a reporting year, estimates for ICWA non-compliant placements may be negatively biased.
+
+I match children across NCANDS and AFCARS to identify children in foster care that ever experienced a maltreatment report, then compare first maltreatment date to first foster care entry date to restrict matches to those cases where a maltreatment report preceded a child's first foster care entry.
+
+Data on child and caretaker race/ethnicity are occasionally missing in AFCARS and NCANDS.
+I construct multiple imputation models to provide reasonable uncertainty bounds on estimates driven by missing data.
+
+# Methods
+
+## Period life tables
+
+The prevalence of an event can be estimated at a point-in-time using annual data on events, but these estimates may be misleading in understanding event prevalence in a population if individuals experience multiple events in a year, multiple events over their life course, or if there are significant levels of mortality during the period of potential exposure.
+For example, a child may be the subject of multiple maltreatment events during a year, or multiple foster care entries across years.
+Infant and child mortality rates may also differ across groups in ways that structure underlying risk of exposure.
+
+Life table methods adjust for the variable risk of an event and the varying size of a population over the life course by following a cohort of children over their life course to observe the incidence of an event.
+Given shifts in risk over time, and the lack of longitudinal data on children over a sufficientily long period of time, I use a period life table approach, that creates a synthetic cohort of 100,000 children, and simulates the risk of experiencing each event based on the observed incidence in the population.
+This method relies on two key assumptions: 1) that event risk remains stable at observed levels for each age-group as the cohort ages and 2) that mortality rates for each age-group remain stable at observed levels as the cohort ages.
+
+As shown in the figures below, risk for each child welfare system event is variable over time, calling into question the assumption of stable event risk.
+I describe these trends by estimating single-year life tables, then provide smoothed estimates by pooling 5 years of data from the most recent observations (2012-2016).
+
+## Marginal and conditional probabilities
+
+Marginal probabilities can be interpreted as either a) the likelihood that an individual will experience an event or b) the proportion of the population that will experience an event.
+In sections **N THROUGH P** I estimate the marginal probability of a series of child welfare events for the complete population of AI/AN and white children.
+
+Conditional probabilities describe the likelihood of an event, conditional on knowing that another event occurred.
+A probability of event $A$ conditional on event $B$, $Pr(A|B)$ can be interpreted as the likelihood that an individual who experiences event B will experience event A.
+Alternatively, we can interprent this probability as the proportion of individuals who experienced B that also experienced A.
+Formally, the conditional probability of a event A conditional on event B is defined as the ratio of the joint probability of A and B divided by the marginal probability of event B:
+
+\[P(A|B) = \frac{P(A,B)}{P(B)}\]
+
+Some events, like substantiation of a maltreatment report, are defined by their relationship to a prior event.
+All children who were substantiated must have a maltreatment report, such that the joint probability of having a substantiated report and having a report is equal to the marginal probability of having a report $Pr(\textrm{substantiation}, \textrm{report}) = Pr(\textrm{substantiation})$.
+For these variables, the conditional probability is simply the ratio of the marginal probabilities: $Pr(A|B) = \frac{Pr(A)}{Pr(B)}$.
+
+For each of these calculations, the probabilities are age-specific event probabilities estimated from period life tables.
+
+# Investigations
 
@@
 
-```{r ncandstables.pop}
+```{r inv, fig.pos="h", out.extra = '', fig.cap = "(A): Risk of maltreatment investigation by age 18, 2003 - 2016 annual period life table.
+(B): Age-specific risk of first maltreatment investigation, 2012 - 2016 pooled period life table"}
 ### read in new first investigation csv
@@
 years<-unique(ncands_inv$year)
-race_ethn<-unique(ncands_inv$race_ethn)
+race_id<-unique(ncands_inv$race_ethn)
 index<-1
@@
   for(j in 1:length(years)){
-    for(r in 1:length(race_ethn)){
+    for(r in 1:length(race_id)){
       temp<-ncands_inv %>% 
@@
       temp<-temp %>% 
-        filter(race_ethn == race_ethn[r])
+        filter(race_ethn == race_id[r])
       temp<-temp %>% 
@@
 
-investigation_tables_c %>% 
+investigation_tables_pooled<-list()
+race_id<-unique(ncands_inv$race_ethn)
+imps<-unique(ncands_inv$.imp)
+index<-1
+for(i in 1:imps){
+  for(r in 1:length(race_id)){
+    temp<-ncands_inv %>% 
+      filter(year>2011) 
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==i)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    investigation_tables_pooled[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+}
+investigation_tables_pooled<-bind_rows(investigation_tables_pooled)
+
+ts<-investigation_tables_c %>% 
   ggplot(aes(x=year, y = c_mn, 
-             fill = race_ethn,
-             group = race_ethn,
+             color = race_ethn,
              ymin = cmin, ymax = cmax)) +
-  geom_ribbon(alpha = .5) + 
-  geom_line(aes(color = race_ethn)) + 
+  geom_line() + 
+  geom_point() + 
   xlab("Year") + 
-  ylab("Proportion with screened-in report before age 18")
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(A)") + 
+  theme(plot.title = element_text(hjust = 0.5))
 
-### way lower rates in 17
-### check total case counts
-### counts are low, probably some non-reporting states, removing for now
+age<-investigation_tables_pooled %>% 
+  ggplot(aes(x=age, y = q, 
+             color = race_ethn)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(B)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+grid.arrange(ts, age)
 ```
 
-FIG 1: TS plot of age 17 cumulative investigation risk by year
+Changes in risk of ever-experiencing a maltreatment investigation over time are displayed in Figure \@ref(fig:inv), panel A.
+Risk for AI/AN children is indicated with the pink line, and white children with a blue line.
+Underlying risk of maltreatment investigations has shifted for both groups over between 2003 and 2016.
+For AI/AN children, risk has declined over time from a peak of 0.38 in 2003 to 0.34 in 2016.
+During this period, risk was at its lowest in 2010 (0.32).
+Inequality in risk between AI/AN and white children narrowed substantially over this period.
+In 2003, AI/AN children were 1.37 times more likely than white children to have a screened-in report.
+In 2016, were 1.05 times more likely than white children to have a screened-in report.
 
-VISUALIZE investigation tables
+Figure \@ref(fig:inv) panel B displays the age-specific risk of a first screened-in maltreatment report for AI/AN and white children using pooled 2012 - 2016 data.
+Risk is highest in the first years of life, and risk for AI/AN children is markedly higher than risk for white children during infancy.
+AI/AN infants are 1.44 times more likely than white infants to recieve a screened-in maltreatment report.
+Between 2012 and 2016, 7.7 percent of AI/AN infants recieve a report, compared to 5.3 percent of white infants.
+AI/AN children are at elevated risk until age 6.
+At later ages, risk profiles for AI/AN and white children converge.
 
-## $P(substantiation)$
+# Substantiation
 
-```{r fctables.ncands}
+Next, I estimate the marginal probability of ever experiencing a maltreatment investigation before age 18, and the probability of experiencing a first investigation at each age between birth and 17.
+Panel A in Figure \@ref(fig:sub) shows how the lifetime risk of substantiation has shifted over time.
+Unlike investigations, substantiation risk levels have not clearly increased or decreased across this time period.
+AI/AN risk levels fluctuated around the period-average level of 0.14, whereas white risk levels centered around 0.1.
+In 2016, AI/AN children's risk of substantiation before age 18 was 0.16, and white chilren's risk was 0.1.
 
+```{r sub,  fig.pos="h", out.extra = '', fig.cap = "(A): Risk of substantiated maltreatment investigation by age 18, 2003 - 2016 annual period life table.
+(B): Age-specific risk of first substantiated maltreatment investigation, 2012 - 2016 pooled period life table"}
 ### run lifetable by year
@@
 years<-unique(ncands_sub$year)
-race_ethn<-unique(ncands_sub$race_ethn)
+race_id<-unique(ncands_inv$race_ethn)
 index<-1
@@
   for(j in 1:length(years)){
-    for(r in 1:length(race_ethn)){
+    for(r in 1:length(race_id)){
       temp<-ncands_sub %>%
@@
       temp<-temp %>%
-        filter(race_ethn == race_ethn[r])
+        filter(race_ethn == race_id[r])
       temp<-temp %>%
@@
 
-subst_tables_c %>% 
+sub_tables_pooled<-list()
+race_id<-unique(ncands_sub$race_ethn)
+imps<-unique(ncands_sub$.imp)
+index<-1
+for(i in 1:imps){
+  for(r in 1:length(race_id)){
+    temp<-ncands_sub %>% 
+      filter(year>2011) 
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==i)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    sub_tables_pooled[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+}
+sub_tables_pooled<-bind_rows(sub_tables_pooled)
+
+ts<-subst_tables_c %>% 
   ggplot(aes(x=year, y = c_mn, 
-             fill = race_ethn,
-             group = race_ethn,
+             color = race_ethn,
              ymin = cmin, ymax = cmax)) +
-  geom_ribbon(alpha = .5) + 
-  geom_line(aes(color = race_ethn)) + 
+  geom_line() + 
+  geom_point() + 
   xlab("Year") + 
-  ylab("Proportion with substantiation before age 18")
-```
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(A)") + 
+  theme(plot.title = element_text(hjust = 0.5))
 
-## $P(subst|inv)$
+age<-sub_tables_pooled %>% 
+  ggplot(aes(x=age, y = q, 
+             color = race_ethn)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(B)") + 
+  theme(plot.title = element_text(hjust = 0.5))
 
-of all investigated, how many are ever substantiated?
-
-from lifetables:
-
-Because $P(inv|subst)=1$, and we've converted population incidence into annual age-adjusted probability, the ratio of events is the conditional probability
-
-\[P(subst|inv) = \frac{d_{subst}}{d_{inv}}\]
-
-```{r}
-
-make_conditionals<-function(num, denom){
-  out<-num %>% 
-  select(age, year, d) %>% 
-  rename(d_num = d) %>% 
-  left_join(denom %>% 
-              select(age, year, d) %>% 
-              rename(d_denom = d)) %>% 
-  group_by(year) %>% 
-  summarise(q = sum(d_num)/sum(d_denom))
-  return(out)
-}
-
-sub_inv<-make_conditionals(subst_tables, investigation_tables)
+grid.arrange(ts, age)
 
@@
 
-## $P(FC)$ using AFCARS
+The shape of the age-specific substantiation risk curve, shown in Figure \@ref(fig:sub) panel B is similar to the shape of the age-specific risk curve (Figure \@ref(fig:inv)).
+This is unsurprising, as an investigation is a necessary condition for a substantiation.
+Unlike with investigations, however, AI/AN children are at higher risk of substantiation at all ages.
+During their first year, AI/AN children are 1.7 times more likely than white children to experience a substantiated report.
+At age 17, AI/AN children are 1.26 times more likely than white children to experience a first substantiated report.
 
-```{r fctables.pop}
+```{r subCond, fig.cap="Age-specific risk of first substantiation, conditional on prior maltreatment report, 2012 - 2016 pooled period life tables"}
+
+## p(sub|inv) = p(inv, sub)/p(inv)
+## p(sub) / p(inv)
+ncands_sub_inv<-sub_tables_pooled %>%
+  mutate(sub_inv = q / investigation_tables_pooled$q)
+
+ggplot(ncands_sub_inv,
+         aes(x = age, y = sub_inv, color = race_ethn)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event")
+```
+
+Figure \@ref(fig:subCond) shows the age-specific probability of a first maltreatment report substantiation, conditional on having a prior maltreatment report (inclusive of a first report).
+AI/AN children are substantially more likely than white children to have a substantiated maltreatment report than white children.
+Among AI/AN infants that have been investigated by a child welfare agency, about 41 percent will be substantiated.
+For white infants, about 36 percent that receive an investigation will be subtantiated.
+The gap is largest at age 9, when AI/AN children with a report are 1.6 times more likely than white children with a report to experience a substantiated report.
+
+# Foster Care
+
+Next, I describe the lifetime and age-specific risk of foster care placement, along with trends in risk over time.
+Figure \@ref(fig:fc) panel A shows changes in lifetime risk of foster care placement for AI/AN and white children between 2003 and 2016.
+Like substantiation risk, both groups saw an increase between 2003 and 2005, a sharp decline between 2005 and 2009, and a gradual increase between 2009 and 2016.
+AI/AN children's lifetime foster care risk was at its highest in 2005, at 0.14, and at its lowest in 2009, at 0.89.
+In 2016, AI/AN lifetime foster care risk was 0.12.
+In 2016, AI/AN children's lifetime risk of ever entering foster care was 2.30 times higher than white children's risk.
+
+```{r fc,  fig.pos="h", out.extra = '', fig.cap = "(A): Risk of ever entering foster care by age 18, 2003 - 2016 annual period life table.
+(B): Age-specific risk of first foster care entry, 2012 - 2016 pooled period life table"}
 fc_tables<-list()
 years<-unique(afcars_fc$year)
-race_ethn<-unique(afcars_fc$race_ethn)
+race_id<-unique(afcars_fc$race_ethn)
 index<-1
-for(i in 1:5){
+for(i in 1:length(unique(afcars_fc$.imp))){
   for(j in 1:length(years)){
-    for(r in 1:length(race_ethn)){
+    for(r in 1:length(race_id)){
       temp<-afcars_fc %>%
@@
 
-fc_tables_c %>% 
+fc_tables_pooled<-list()
+race_id<-unique(afcars_fc$race_ethn)
+imps<-unique(afcars_fc$.imp)
+index<-1
+for(i in 1:imps){
+  for(r in 1:length(race_id)){
+    temp<-afcars_fc %>% 
+      filter(year>2011) 
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==i)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    fc_tables_pooled[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+}
+fc_tables_pooled<-bind_rows(fc_tables_pooled)
+
+ts<-fc_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             color = race_ethn,
+             ymin = cmin, ymax = cmax)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Year") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(A)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+age<-fc_tables_pooled %>% 
+  ggplot(aes(x=age, y = q, 
+             color = race_ethn)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(B)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+grid.arrange(ts, age)
+
+```
+
+Figure \@ref(fig:fc) panel B shows the age-specific risk of a fist foster care entry using 2012 - 2016 pooled AFCARS data.
+AI/AN children are at markedly higher risk of entering foster care than are white children throughout childhood.
+In infancy, about 2.7 percent of AI/AN children enter foster care, compared to about 1 percent of white children.
+At age 17, inequalities persist despite relatively low event risk: AI/AN children are 1.7 times more likely to experience a first foster care entry at 17 than are their white peers.
+
+```{r}
+## P(FC,inv)
+fc_post_inv_tables<-list()
+years<-unique(afcars_fc_inv$year)
+race_id<-unique(afcars_fc_inv$race_ethn)
+index<-1
+for(i in 1:length(unique(afcars_fc_inv$.imp))){
+  for(r in 1:length(race_id)){
+    temp<-afcars_fc_inv %>%
+      filter(year>2011)
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==i)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    fc_post_inv_tables[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+}
+
+# P(A,B)
+fc_post_inv_tables<-bind_rows(fc_post_inv_tables)
+
+# P(A|B) = P(A,B)/P(B)
+fc_cond_inv<-fc_post_inv_tables %>%
+  mutate(fc_inv = q / investigation_tables_pooled$q)
+
+ggplot(fc_cond_inv,
+       aes(x = age, y = fc_inv, 
+           color = race_ethn)) + 
+  geom_line()
+```
+
+# Placement outside of kin, native foster homes
+
+The Indian Child Welfare Act, among other protections including tribal jurisdiction over AI/AN children, guarantees native children in foster care the right to be in a placement that preserves their heritage *CITE MEASURING COMPLIANCE*.
+Preferences are guaranteed for kin, and non-kin members of the child's nation.
+Otherwise, AI/AN foster families from outside of the child's tribe or nation are to be preferred.
+Placement into a non-native non-kin placement is to be considered a last resort.
+
+Figure \@ref(fig:icwa) panel A shows the lifetime risk of ever being placed into a non-kin non-native foster placement for AI/AN children over time.
+Note that AFCARS only records the final placement setting for each child in a reporting year, so these estimates are likely negatively biased.
+AI/AN children's lifetime risk of ever being placed in a non-kin non-native foster placement has declined over time, from a peak of 0.12 in 2005 to a low of 0.08 in 2016.
+Panel B displays the age-specific risk of non-kin, non-native foster placement for AI/AN children.
+As with other foster placements, risk is highest in infancy, and declines as children age.
+**CALL OUT CONDITIONAL PROB FIG**
+
+```{r icwa,  fig.pos="h", out.extra = '', fig.cap = "(A): AI/AN risk of ever being placed in non-kin non-AI/AN caretaker foster care by age 18, 2003 - 2016 annual period life table.
+(B): AI/AN Age-specific risk of being placed in non-kin non-AI/AN caretaker foster care, 2012 - 2016 pooled period life table"}
+icwa_tables<-list()
+years<-unique(afcars_non_icwa$year)
+race_id<-unique(afcars_non_icwa$race_ethn)
+index<-1
+for(i in 1:length(unique(afcars_non_icwa$.imp))){
+  for(j in 1:length(years)){
+    for(r in 1:length(race_id)){
+      temp<-afcars_non_icwa %>%
+        filter(year==years[j])
+      temp<-temp %>%
+        filter(race_ethn == race_id[r])
+      temp<-temp %>%
+        filter(.imp==i)
+      icwa_tables[[index]]<-make_life_table(temp)
+      index<-index+1
+    }
+  }
+}
+icwa_tables<-bind_rows(icwa_tables)
+
+icwa_tables_pooled<-list()
+race_id<-unique(afcars_non_icwa$race_ethn)
+imps<-unique(afcars_non_icwa$.imp)
+index<-1
+for(i in 1:imps){
+  for(r in 1:length(race_id)){
+    temp<-afcars_non_icwa %>% 
+      filter(year>2011) 
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==i)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    icwa_tables_pooled[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+}
+icwa_tables_pooled<-bind_rows(icwa_tables_pooled)
+
+icwa_tables_c<-icwa_tables %>%
+  filter(age==17) %>%
+  group_by(year, race_ethn) %>%
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
+
+ts<-icwa_tables_c %>% 
+  ggplot(aes(x=year, y = c_mn, 
+             ymin = cmin, ymax = cmax)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Year") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(A)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+age<-icwa_tables_pooled %>% 
+  ggplot(aes(x=age, y = q)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(B)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+grid.arrange(ts, age)
+```
+
+# Termination of parental rights
+
+```{r tprTS,  fig.pos="h", out.extra = '', fig.cap="Risk of termination of parental rights while in foster care before age 18, 2003 - 2016 period life table estimates"}
+tpr_tables<-list()
+years<-unique(afcars_tpr$year)
+race_id<-unique(afcars_tpr$race_ethn)
+index<-1
+for(i in 1:length(unique(afcars_tpr$.imp))){
+  for(j in 1:length(years)){
+    for(r in 1:length(race_id)){
+      temp<-afcars_tpr %>%
+        filter(year==years[j])
+      temp<-temp %>%
+        filter(race_ethn == race_id[r])
+      temp<-temp %>%
+        filter(.imp==i)
+      tpr_tables[[index]]<-make_life_table(temp)
+      index<-index+1
+    }
+  }
+}
+tpr_tables<-bind_rows(tpr_tables)
+
+tpr_tables_c<-tpr_tables %>%
+  filter(age==17) %>%
+  group_by(year, race_ethn) %>%
+  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
+
+tpr_tables_c %>%
   ggplot(aes(x=year, y = c_mn,
@@
   xlab("Year") +
-  ylab("Proportion ever in foster care before age 18")
+  ylab("Risk of event") + 
+  labs(fill = "", color = "")
 ```
 
-## $P(investigation , entered)$ using afcars
+## P(FC|sub)
+```{r}
 
-```{r}
-fc_post_inv_tables<-list()
-years<-unique(afcars_fc_inv$year)
-race_ethn<-unique(afcars_fc_inv$race_ethn)
-index<-1
-for(i in 1:5){
-  for(j in 1:length(years)){
-    for(r in 1:length(race_ethn)){
-      temp<-afcars_fc_inv %>% 
-        filter(year==years[j])
-      temp<-temp %>% 
-        filter(race_ethn == race_ethn[r])
-      temp<-temp %>% 
-        filter(.imp==i)
-      fc_post_inv_tables[[index]]<-make_life_table(temp)
-      index<-index+1
-    }
-  }
-}
-fc_post_inv_tables<-bind_rows(fc_post_inv_tables)
-
-fc_post_inv_tables_c<-fc_post_inv_tables %>% 
-  filter(age==17) %>% 
-  group_by(year, race_ethn) %>% 
-  summarise(c_mn = mean(c), cmax = max(c), cmin = min(c))
-
-fc_post_inv_tables_c %>% 
-  ggplot(aes(x=year, y = c_mn, 
-             fill = race_ethn,
-             group = race_ethn,
-             ymin = cmin, ymax = cmax)) +
-  geom_ribbon(alpha = .5) + 
-  geom_line(aes(color = race_ethn)) + 
-  xlab("Year") + 
-  ylab("Proportion with foster care entry after screened-in report")
-
+# fc_cond_sub<-fc_post_inv_tables %>%
+#   mutate(fc_sub = q / subst_tables$q)
+# 
+# ggplot(fc_cond_sub %>%
+#          filter(year==2016),
+#          aes(x = age, y = fc_sub, color = race_ethn)) +
+#   geom_line()
 ```
 
-RECALL THAT
+## P(FC|tpr)
+```{r}
+### use bayes' rule because p(fc|tpr)=1
+# \[P(B|A) = \frac{P(A|B)P(B)}{P(A)}\]
+## P(tpr|fc) = p(fc|tpr)p(tpr) / p(fc)
+## p(tpr|fc) = p(tpr)/p(fc)
+# fc_tpr_fc<-tpr_tables %>%
+#   mutate(tpr_fc = q/ fc_tables$q)
+# 
+# ggplot(fc_tpr_fc %>%
+#          filter(year==2016),
+#          aes(x = age, y = tpr_fc, color = race_ethn)) +
+#   geom_line()
 
-\[P(A|B) = \frac{P(A,B)}{P(B)}\]
-
-ALSO, BAYES THEOREM
-
-\[P(B|A) = \frac{P(A|B)P(B)}{P(A)}\]
-
-for $A=FC$ and $B=investigation$, we wish to obtain $P(A|B)$. We obtain differing measurements of $P(A,B)$ from two datasets. I'll calculate each and compare.
-
-## $P(FC|inv)$
-
-```{r}
-fc_inv<-make_conditionals(fc_tables, investigation_tables)
-fc_sub<-make_conditionals(fc_tables, subst_tables)
-fc_ncands_inv<-make_conditionals(ncands_fc_tables, 
-                                 investigation_tables)
-fc_ncands_subst<-make_conditionals(ncands_fc_tables, 
-                                 subst_tables)
 ```

### 2019-12-30 read.r

@@ -53,7 +53,6 @@ ncands<-ncands %>%
   filter(!(is.na(subyr)))
 
 ncands_first<-ncands %>%
-  select(.imp, chid, race_ethn, chage, rptdt) %>%
   arrange(.imp, rptdt) %>% 
   group_by(.imp, stfcid) %>%
   slice(1)
@@ -125,7 +124,7 @@ afcars_inv<-afcars_inv %>%
 # remove when year(rptdt)>year
 
 afcars_inv<-afcars_inv %>%
-  filter(year<=rpt_year)
+  filter(year>=rpt_year)
 
 afcars_inv_nat<-afcars_inv %>%
   group_by(.imp, age, race_ethn, year) %>%
@@ -134,6 +133,20 @@ afcars_inv_nat<-afcars_inv %>%
   ungroup() %>%
   write_csv("./data/afcars_fc_post_inv.csv")
 
+afcars_sub<-afars_first %>% 
+  left_join(ncands_first_subst %>% 
+              select(.imp, stfcid, rptdt)) %>%
+  mutate(rptyr=year(rptdt)) %>% 
+  filter(!(is.na(rptyr))) %>% 
+  filter(year>=rptyr)
+  
+afcars_sub_nat<-afcars_sub %>% 
+  group_by(.imp, age, race_ethn, year) %>%
+  summarise(var = n()) %>%
+  left_join(pop) %>%
+  ungroup() %>%
+  write_csv("./data/afcars_fc_post_sub.csv")
+  
 afcars_tpr<-afcars %>%
   filter(istpr==1) %>%
   arrange(.imp, year) %>% 

### 2019-12-31 read.r
add joint fc,sub data

@@ -40,9 +40,6 @@ state.fips<-state.fips %>%
   distinct() %>% 
   rename(staterr = abb)
 
-zz <- file("all.Rout", open = "wt")
-sink(zz)
-sink(zz, type = "message")
 ### attach stfcid
 ncands<-ncands %>%
   left_join(ncands_xwalk)
@@ -55,7 +52,9 @@ ncands<-ncands %>%
 ncands_first<-ncands %>%
   arrange(.imp, rptdt) %>% 
   group_by(.imp, stfcid) %>%
-  slice(1)
+  slice(1) 
+
+ncands_first<-ncands_first %>% 
   ungroup() %>%
   mutate(year = year(rptdt))
 
@@ -89,13 +88,6 @@ afcars_first <- afcars %>%
   slice(1) %>% 
   ungroup()
 
-sink("log.txt")
-print("afcars_first length")
-print(nrow(afcars_first))
-print("unique fcid")
-print(length(unique(afcars$stfcid)))
-sink()
-
 afcars_first_nat<-afcars_first %>% 
   group_by(.imp, age, race_ethn, year) %>% 
   summarise(var = n()) %>% 
@@ -107,15 +99,9 @@ afcars_inv<-afcars_first %>%
   left_join(ncands_xwalk %>% 
               select(stfcid, rptdt) %>% 
               arrange(rptdt) %>% 
+              group_by(stfcid) %>% 
               slice(1) %>% 
               mutate(ncands_inv = T))
-
-sink("log2.txt")
-print("afcars_inv post join length")
-print(nrow(afcars_inv))
-print("unique ids")
-print(length(unique(afcars_inv$stfcid)))
-sink()
 
 afcars_inv<-afcars_inv %>% 
   mutate(rpt_year = year(rptdt)) %>% 
@@ -133,7 +119,7 @@ afcars_inv_nat<-afcars_inv %>%
   ungroup() %>%
   write_csv("./data/afcars_fc_post_inv.csv")
 
-afcars_sub<-afars_first %>% 
+afcars_sub<-afcars_first %>% 
   left_join(ncands_first_subst %>% 
               select(.imp, stfcid, rptdt)) %>%
   mutate(rptyr=year(rptdt)) %>% 
@@ -181,23 +167,3 @@ afcars_icwa_nat<-afcars_icwa %>%
   ungroup() %>%
   write_csv("./data/afcars_non_icwa.csv")
 
[…]

### 2020-01-01 ms.rmd
final update before 12/31 report

@@
   filter(race_ethn == "AI/AN" | race_ethn == "White")
+afcars_fc_sub<-read_csv("./data/afcars_fc_post_sub.csv")%>% 
+  filter(year>=2003, year<2017)%>% 
+  filter(age<18) %>% 
+  filter(.imp!=0) %>% 
+  filter(race_ethn == "AI/AN" | race_ethn == "White")
 ```
@@
 
-This report examines how american indian children are processed through
-us child welfare systems.
-it explores the likelihood that american indian children eventually end up in foster care, end up in foster care with non-indian foster parents, or see the rights of their parents terminated.
-I compare these pathways through the foster care system to the pathways experienced by white and children.
+This report examines how American Indian / Alaska Native children are processed through state and local child welfare systems.
+It establishes age-specific and lifetime prevalence of a series of child welfare system events, then evaluates how likely cases are to move "up" the chain of more serious outcomes, conditional on experiencing a lower-level outcome.
+In doing so, it shows that inequalities in child welfare system outcomes for AI/AN children emerge at distinct stages of case processing.
+These national data provide a high-level overview of system inequalities, but may mask important sub-national variation.
 
-**FINDINGS HERE**
+- *Investigation*: Nationally, AI/AN children only experience inequality in investigations at young ages: between birth and age 5.
+As they age, risk of investigation converges to the risk levels for white children.
+- *Substantiation*: However, AI/AN children face a higher risk of substantiation than white children at all ages.
+Conditional on receiving an investigation, AI/AN children are at much higher risk of substantiation than are white children.
+- *Foster care*: AI/AN children face much higher risk of foster care than white children at all ages, and these gaps persist conditional on maltreatment investigation or maltreatment substantiation. 
+- *Placement in non-kin, non-native foster settings*: An overwhelming majority of AI/AN children who ever enter foster care experience placement in a setting that does not meet the standards for appropriate placements for AI/AN children specified in the Indian Child Welfare Act
+- *Termination of parental rights*: AI/AN children are more likely than white children to have their parents' rights terminated through the child welfare system.
+However, this inequality appears to be a function of higher levels of foster care placement.
+Conditional on ever being placed in foster care, AI/AN risks are similar to white risks.
 
@@
 
-American Indian / Alaska Native (AI/AN) children and families are far more likely to experience contact with the child welfare system than are their white peers *(CITES)*. *STATEMENT ON SIGNIFICANCE*
-
+American Indian / Alaska Native (AI/AN) children and families are far more likely to experience contact with the child welfare system than are their white peers [@haightScopingStudyIndigenous2018].
 This report provides national estimates of the prevalence of a series of child welfare system contact outcomes for AI/AN children, in addition to estimating the likelihood that children transition to more serious child welfare system outcomes conditional on lower-level contacts, such as investigations.
 
-I use the 2003 - 2016 AFCARS and NCANDS to calculate period lifetables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster care settings, and termination of parental rights.
+I use the 2003 - 2016 AFCARS and NCANDS to calculate period life tables for lifetime risk of investigation, substantiation, placement, placement into a non-ICWA compliant foster care settings, and termination of parental rights.
 Each of these tables provides an estimate of the cumulative risk of experiencing the event before age 18 if risk-levels observed during the reported period remain constant.
 
-For the national population of both AI/AN and white children, I estimate the following probabilities of event occurence by age, and as a cumulative likelihood of event occurrence before a child's 18th birthday:
+For the national population of both AI/AN and white children, I estimate the following probabilities of event occurrence by age, and as a cumulative likelihood of event occurrence before a child's 18th birthday:
 
@@
 3. Foster care placement
-3. Placement in foster home that is not compliant with ICWA guidelines (for AI/AN children)
+3. Placement in a non-kin, non-AI/AN foster home 
 4. Termination of both parent's rights while in foster care
@@
 3. Foster care placement, conditional on substantiated report
-4. ICWA non-compliant placement, conditional on any foster care placement
-5. Parental rights termination, conditional on any foster care placmeent
+4. Placement in a non-kin, non-AI/AN foster home, conditional on any foster care placement
+5. Parental rights termination, conditional on any foster care placement
 
@@
 Life table methods adjust for the variable risk of an event and the varying size of a population over the life course by following a cohort of children over their life course to observe the incidence of an event.
-Given shifts in risk over time, and the lack of longitudinal data on children over a sufficientily long period of time, I use a period life table approach, that creates a synthetic cohort of 100,000 children, and simulates the risk of experiencing each event based on the observed incidence in the population.
+Given shifts in risk over time, and the lack of longitudinal data on children over a sufficiently long period of time, I use a period life table approach, that creates a synthetic cohort of 100,000 children, and simulates the risk of experiencing each event based on the observed incidence in the population.
 This method relies on two key assumptions: 1) that event risk remains stable at observed levels for each age-group as the cohort ages and 2) that mortality rates for each age-group remain stable at observed levels as the cohort ages.
@@
 Marginal probabilities can be interpreted as either a) the likelihood that an individual will experience an event or b) the proportion of the population that will experience an event.
-In sections **N THROUGH P** I estimate the marginal probability of a series of child welfare events for the complete population of AI/AN and white children.
+I estimate the marginal probability of a series of child welfare events for the complete population of AI/AN and white children.
 
@@
 A probability of event $A$ conditional on event $B$, $Pr(A|B)$ can be interpreted as the likelihood that an individual who experiences event B will experience event A.
-Alternatively, we can interprent this probability as the proportion of individuals who experienced B that also experienced A.
+Alternatively, we can interpret this probability as the proportion of individuals who experienced B that also experienced A.
 Formally, the conditional probability of a event A conditional on event B is defined as the ratio of the joint probability of A and B divided by the marginal probability of event B:
@@
 Risk is highest in the first years of life, and risk for AI/AN children is markedly higher than risk for white children during infancy.
-AI/AN infants are 1.44 times more likely than white infants to recieve a screened-in maltreatment report.
-Between 2012 and 2016, 7.7 percent of AI/AN infants recieve a report, compared to 5.3 percent of white infants.
+AI/AN infants are 1.44 times more likely than white infants to receive a screened-in maltreatment report.
+Between 2012 and 2016, 7.7 percent of AI/AN infants receive a report, compared to 5.3 percent of white infants.
 AI/AN children are at elevated risk until age 6.
@@
 AI/AN risk levels fluctuated around the period-average level of 0.14, whereas white risk levels centered around 0.1.
-In 2016, AI/AN children's risk of substantiation before age 18 was 0.16, and white chilren's risk was 0.1.
+In 2016, AI/AN children's risk of substantiation before age 18 was 0.16, and white children's risk was 0.1.
 
@@
   xlab("Age") + 
-  ylab("Risk of event")
+  ylab("Risk of event")+ 
+  labs(fill = "", color = "")
 ```
+
+## Substantiation, conditional on investigation
 
@@
 Among AI/AN infants that have been investigated by a child welfare agency, about 41 percent will be substantiated.
-For white infants, about 36 percent that receive an investigation will be subtantiated.
+For white infants, about 36 percent that receive an investigation will be substantiated.
 The gap is largest at age 9, when AI/AN children with a report are 1.6 times more likely than white children with a report to experience a substantiated report.
@@
 
-```{r}
+## Foster care, conditional on investigation and substantiation
+
+AI/AN children are more likely to face more serious consequences of maltreatment investigations and confirmed maltreatment reports than their white peers.
+Conditional on an investigation, or conditional on a substantiation, AI/AN children are substantially more likely than white children to ever enter foster care.
+Figure \@ref(fig:fcCond) shows the conditional probabilities of ever entering foster care following a first maltreatment investigation (panel A) and a first confirmed maltreatment investigation (panel B) for each age using life tables estimated from 2012 - 2016 pooled data.
+About 26 percent of AI/AN infants who are investigated will later be placed into foster care, compared to about 15 percent of white infants.
+About 54 percent of AI/AN infants with a substantiated report will ever be placed into foster care, compared to 38 percent of white infants.
+
+```{r fcCond,  fig.pos="h", out.extra = '', fig.cap = "(A): Age-specific risk of first foster care entry conditional on screened-in maltreatment report, 2012 - 2016 pooled period life table.
+(B):  Age-specific risk of first foster care entry conditional on substantiated maltreatment report, 2012 - 2016 pooled period life table"}
 ## P(FC,inv)
@@
 index<-1
-for(i in 1:length(unique(afcars_fc_inv$.imp))){
+#for(i in 1:length(unique(afcars_fc_inv$.imp))){
   for(r in 1:length(race_id)){
@@
     temp<-temp %>% 
-      filter(.imp==i)
+      filter(.imp==1)
     temp<-temp %>% 
@@
   }
-}
-
+#}
 # P(A,B)
@@
 
-ggplot(fc_cond_inv,
+##P(FC,sub)
+fc_sub_tables<-list()
+years<-unique(afcars_fc_sub$year)
+race_id<-unique(afcars_fc_sub$race_ethn)
+index<-1
+#for(i in 1:length(unique(afcars_fc_sub$.imp))){
+  for(r in 1:length(race_id)){
+    temp<-afcars_fc_sub %>%
+      filter(year>2011)
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==1)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    fc_sub_tables[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+#}
+# P(A,B)
+fc_sub_tables<-bind_rows(fc_sub_tables)
+# P(A|B) = P(A,B)/P(B)
+fc_sub_cond<-fc_sub_tables %>%
+  mutate(fc_sub = q / sub_tables_pooled$q)
+
+inv<-ggplot(fc_cond_inv,
        aes(x = age, y = fc_inv,
            color = race_ethn)) +
-  geom_line()
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(A)")
+
+sub<-ggplot(fc_sub_cond,
+            aes(x = age, y = fc_sub,
+                color = race_ethn)) + 
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(B)")
+
+grid.arrange(inv, sub)
+
+### NEXT DO p(fc,subst)
 ```
@@
 
-The Indian Child Welfare Act, among other protections including tribal jurisdiction over AI/AN children, guarantees native children in foster care the right to be in a placement that preserves their heritage *CITE MEASURING COMPLIANCE*.
+The Indian Child Welfare Act, among other protections including tribal jurisdiction over AI/AN children, guarantees native children in foster care the right to be in a placement that preserves their heritage [@williamsIndianChildWelfare2015].
 Preferences are guaranteed for kin, and non-kin members of the child's nation.
@@
 As with other foster placements, risk is highest in infancy, and declines as children age.
-**CALL OUT CONDITIONAL PROB FIG**
+Note that these are population risks; risks only for those children in foster care are presented in Figure \@ref(fig:icwaCond).
 
@@
 
+## Placement outside of kin and/or native foster homes conditional on foster care entry
+
+Conditional on a foster care entry, an overwhelming majority of AI/AN children experience a foster care placement in a non-kin setting where the primary caretaker is not American Indian / Alaska Native.
+Using a 2012 - 2016 pooled life table, Figure \@ref(fig:icwaCond) shows the probability of being placed in a non-kin foster setting with neither primary caretaker being reported as AI/AN, conditional on a first foster care placement at each age.
+At no age is this risk less than 0.7.
+Young children have the highest probability of being placed in placement settings specified as preferred under ICWA.
+When children experience their first placement as teenagers, their risk of being placed in a non-kin non-AI/AN setting increases dramatically, to a peak of 0.95 at age 15.
+
+```{r icwaCond,  fig.pos="h", out.extra = '', fig.cap = "AI/AN Age-specific risk of being placed in non-kin non-AI/AN caretaker foster care conditional on foster care entry, 2012 - 2016 pooled period life table"}
+## P(non-icwa|placement)
+## P(non-icwa, placement)/P(placement)
+fc_tables_pooled_aian<-fc_tables_pooled %>% 
+  filter(race_ethn=="AI/AN")
+
+icwa_cond<-icwa_tables_pooled %>% 
+  mutate(icwa_cond = q / fc_tables_pooled_aian$q)
+
+ggplot(icwa_cond,
+         aes(x = age, y = icwa_cond)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") 
+```
+
 # Termination of parental rights
 
-```{r tprTS,  fig.pos="h", out.extra = '', fig.cap="Risk of termination of parental rights while in foster care before age 18, 2003 - 2016 period life table estimates"}
+AI/AN children are more likely than white children to experience the termination of both parents' rights while in foster care than are white children.
+Figure \@ref(fig:tpr) shows shifts in cumulative risk over time in panel A, and age-specific risks in recent years in panel B.
+Note the decline in probability in 2015 and 2016 is in part, a function of children entering foster care in those years having a shorter duration of exposure to risk; additional years of data collection may reveal upward revisions in those estimates.
+For all years included in this analysis, AI/AN children were at higher risk of having their parents' rights terminated.
+In the average year during this period (excluding 2015 and 2016), about 3 percent of all AI/AN children had their parents' rights terminated through the foster care system, compared to a risk of about 1 percent for white children.
+
+Panel B shows the age-specific probability of termination of parental rights while in foster care for the full child population using 2012 - 2016 pooled data for life table estimates.
+As with other foster care outcomes, infants are at highest risk.
+AI/AN children are at higher risk than white children.
+From birth to age 1, AI/AN infants face a risk 2.3 times higher than white infant risk of parental rights termination.
+
+```{r tpr,   fig.pos="h", out.extra = '', fig.cap = "(A): Risk of termination of parental rights while in foster care by age 18, 2003 - 2016 annual period life table.
+(B): Age-specific risk of termination of parental rights while in foster care, 2012 - 2016 pooled period life table"}
 tpr_tables<-list()
@@
 
+tpr_tables_pooled<-list()
+race_id<-unique(afcars_tpr$race_ethn)
+imps<-unique(afcars_tpr$.imp)
+index<-1
+for(i in 1:imps){
+  for(r in 1:length(race_id)){
+    temp<-afcars_tpr %>% 
+      filter(year>2011) 
+    temp<-temp %>% 
+      filter(race_ethn == race_id[r])
+    temp<-temp %>% 
+      filter(.imp==i)
+    temp<-temp %>% 
+      group_by(.imp, age, race_ethn) %>% 
+      summarise(var = sum(var), pop = sum(pop)) %>% 
+      ungroup()
+    tpr_tables_pooled[[index]]<-make_life_table(temp)
+    index<-index+1
+  }
+}
+tpr_tables_pooled<-bind_rows(tpr_tables_pooled)
+
 tpr_tables_c<-tpr_tables %>%
@@
 
-tpr_tables_c %>%
+ts<-tpr_tables_c %>%
   ggplot(aes(x=year, y = c_mn,
-             fill = race_ethn,
+             color = race_ethn,
              group = race_ethn,
              ymin = cmin, ymax = cmax)) +
-  geom_ribbon(alpha = .5) +
-  geom_line(aes(color = race_ethn)) +
+  geom_line() +
+  geom_point() +
   xlab("Year") +
+  ylab("Risk of event") +
+  labs(fill = "", color = "",
+       title = "(A)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+age<-tpr_tables_pooled %>% 
+  ggplot(aes(x=age, y = q,
+             color = race_ethn)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
+  ylab("Risk of event") + 
+  labs(fill = "", color = "",
+       title = "(B)") + 
+  theme(plot.title = element_text(hjust = 0.5))
+
+grid.arrange(ts, age)
+```
+
+Figure \@ref(fig:tprCond) shows the probability of parental rights termination, conditional on ever entering foster care.
+Unlike other child welfare system outcomes examined here, AI/AN children's risk of having parental rights termination, conditional on ever entering foster care is similar to or slightly lower than white children's risk at all ages.
+About 36 percent of AI/AN children who entered foster care during infancy saw their parents' rights later terminated, compared to about 40 percent of white children who entered foster care during infancy.
+
+```{r tprCond,  fig.pos="h", out.extra = '', fig.cap = "Age-specific risk of termination of parental rights while in foster care conditional on foster care entry, 2012 - 2016 pooled period life table"}
+## conditionals
+tpr_cond<-tpr_tables_pooled %>% 
+  mutate(tpr_cond = q / fc_tables_pooled$q)
+
+ggplot(tpr_cond,
+         aes(x = age, y = tpr_cond,
+             color = race_ethn)) +
+  geom_line() + 
+  geom_point() + 
+  xlab("Age") + 
   ylab("Risk of event") + 
@@
 
-## P(FC|sub)
-```{r}
+# Conclusion
 
-# fc_cond_sub<-fc_post_inv_tables %>%
-#   mutate(fc_sub = q / subst_tables$q)
-# 
-# ggplot(fc_cond_sub %>%
-#          filter(year==2016),
-#          aes(x = age, y = fc_sub, color = race_ethn)) +
-#   geom_line()
-```
+This report provides clear evidence that AI/AN children face unequal outcomes throughout non-tribal child welfare systems at a national level.
+It shows that the investigation stage (for young children), the substantiation decision stage, and the placement decision stage all contribute to inequalities.
+At the national level, parental rights termination decisions for children in foster care are similar for white and AI/AN children.
 
-## P(FC|tpr)
-```{r}
-### use bayes' rule because p(fc|tpr)=1
-# \[P(B|A) = \frac{P(A|B)P(B)}{P(A)}\]
-## P(tpr|fc) = p(fc|tpr)p(tpr) / p(fc)
-## p(tpr|fc) = p(tpr)/p(fc)
-# fc_tpr_fc<-tpr_tables %>%
-#   mutate(tpr_fc = q/ fc_tables$q)
-# 
-# ggplot(fc_tpr_fc %>%
-#          filter(year==2016),
-#          aes(x = age, y = tpr_fc, color = race_ethn)) +
-#   geom_line()
+These descriptive findings cannot attribute causation to any stage of child welfare decision making, however.
+It is likely that incident-level, child-level, family-level, community-level, agency-level, and state-level characteristics all impact the likelihood that a child becomes subject to a report, and experiences particular trajectories as they are processed through a child welfare system.
+Future research should be closely attentive to how characteristics of the reported maltreatment episode and contextual characteristics, such as bias, law, and resources impact case outcomes.
 
-```
+These findings must be interpreted in the context of the legacy of the child welfare system in AI/AN communities.
+Historically, family separation and fostering have been used as a tool of forced assimilation and cultural genocide [@jacobsGenerationRemovedFostering2014; @crossAmericanIndianEducation1999; @haightScopingStudyIndigenous2018].
+These findings show that despite landmark federal legislation and substantial efforts to build autonomous institutions for tribal governments to independently operate child welfare systems [@tropeSurveyAnalysisSelect2014], family separation through child welfare systems is still a common experience for AI/AN families.
 
-NEXT: compare these to raw rates,
-
-Compare to lifetime risk of ever investigated
-
-CONDITIONALS ONLY WORK FROM SAME DATA - CANT ASSUME THAT p(b|a)=1 otherwise
-
-need to go back to original afcars - rely on imputed ncands to get uncertainty, link to afcars by id, check year for investigation incident, maybe +/-1 yr for lags in placement starts
-
-want to use the proportions to obtain information about probability. perhaps use a model to populate uncertainty? or construct CI from data - either way should be ok
+# References {-#references}

### 2020-05-18 ms.rmd
update data

--- 2019-12-30 ms.rmd
@@
 library(gridExtra)
+library(bookdown)
 source("lifetable.r")
@@
 ncands_inv<-read_csv("./data/ncands_first_inv.csv") %>% 
-  filter(year>=2003, year<2017) %>% 
+  filter(year>=2011, year<2017) %>% 
   filter(age<18) %>% 
@@
 ncands_sub<-read_csv("./data/ncands_first_sub.csv")%>% 
-  filter(year>=2003, year<2017)%>% 
+  filter(year>=2011, year<2017)%>% 
   filter(age<18) %>% 
@@
 afcars_fc<-read_csv("./data/afcars_first_fc.csv")%>% 
-  filter(year>=2003, year<2017)%>% 
+  filter(year>=2011, year<2017)%>% 
   filter(age<18) %>%   
@@
 afcars_tpr<-read_csv("./data/afcars_tpr.csv")%>% 
-  filter(year>=2003, year<2017)%>% 
+  filter(year>=2011, year<2017)%>% 
   filter(age<18) %>% 
@@
 afcars_non_icwa<-read_csv("./data/afcars_non_icwa.csv")%>% 
-  filter(year>=2003, year<2017)%>% 
+  filter(year>=2011, year<2017)%>% 
   filter(age<18) %>% 
@@
 afcars_fc_inv<-read_csv("./data/afcars_fc_post_inv.csv")%>% 
-  filter(year>=2003, year<2017)%>% 
+  filter(year>=2011, year<2017)%>% 
   filter(age<18) %>% 
@@
 grid.arrange(ts, age)
+p_out<-arrangeGrob(ts, age)
+ggsave("./vis/aian_inv.pdf", p_out)
 ```
@@
 grid.arrange(ts, age)
-
+p_out<-arrangeGrob(ts, age)
+ggsave("./vis/aian_sub.pdf", p_out)
 ```
@@
 
-ggplot(ncands_sub_inv,
+p<-ggplot(ncands_sub_inv,
          aes(x = age, y = sub_inv, color = race_ethn)) +
@@
   ylab("Risk of event")
+
+p
+
+ggsave("./vis/aian_sub_inv.pdf", p)
 ```
@@
 grid.arrange(ts, age)
-
+p_out<-arrangeGrob(ts, age)
+ggsave("./vis/aian_fc.pdf", p_out)
 ```
@@
 
-ggplot(fc_cond_inv,
-       aes(x = age, y = fc_inv, 
-           color = race_ethn)) + 
-  geom_line()
+# ggplot(fc_cond_inv,
+#        aes(x = age, y = fc_inv, 
+#            color = race_ethn)) + 
+#   geom_line()
 ```
@@
 grid.arrange(ts, age)
+p_out<-arrangeGrob(ts, age)
+ggsave("./vis/aian_icwa.pdf", p_out)
 ```

### 2020-05-18 read_ncands.r

@@ -47,11 +47,13 @@ state.fips<-state.fips %>%
 ncands_xwalk<-ncands_xwalk %>% 
   left_join(state.fips) %>% 
   filter(afcarsid!="",
-         staterr!="XX") %>% 
+         staterr!="XX") 
+
+ncands_xwalk<-ncands_xwalk%>% 
   mutate(stfcid = paste(fips, afcarsid, sep="")) 
 
 ncands_xwalk<-ncands_xwalk %>% 
-  select(chid, stfcid, rptdt, subyr)
+  select(staterr, chid, stfcid, rptdt, subyr)
 
 write_csv(ncands_xwalk, 
           "~/Projects/ai_an_transitions/data/ncands_xwalk.csv")

### 2020-12-08 visuals_cjlr.r
update prior to starting CFP report

@@ -190,10 +190,19 @@ ggplot(icwa_delta_pct %>%
   ggsave("./vis/cjlr/1_3.png", width = 8, height = 4)
     
   
-  
+icwa_nat<-icwa_delta %>% 
+  filter(!state%in%index$state) %>% 
+  group_by(period, race_ethn) %>% 
+  summarise(fc = sum(fc),
+            adopted = sum(adopted))
 
 
-
+### plus boarding
+total_inst<-icwa_delta %>% 
+  group_by(period, race_ethn) %>% 
+  summarise(fc = sum(fc),
+            adopted = sum(adopted, na.rm=T)) %>% 
+  mutate(total = fc + adopted)
 
 
 ##############################################################

### 2020-12-08 cfp_final_report.Rmd

---
title: "Mechanisms of AIAN Inequality in Child Welfare Across U.S. States."
author: "Frank Edwards and Theresa Rocha Beardall"
date: "12/2020"
abstract: "American Indian and Alaska Native children are separated from their families by state child welfare agencies at exceptionally high rates. This study connects contemporary trends in Native family separation to histories of Indian child removal, and locates places and institutional sites where inequality emerges."
output: bookdown::html_document2
---

```{r setup, include=FALSE}
### load packages and suppress warnings
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(tidyverse, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)

library(bookdown)
library(geofacet)
library(RColorBrewer)
library(mapproj)
library(usmap)
library(gridExtra)

```

```{r read-make}
#### make data for easy load
# source("make_state_tables.R")
# source("make_national_tables.r")
# source("read_adopt.r")
# 
# save.image("./cfp_final_report.RData")

load("./cfp_final_report.RData")
```

# Introduction

clear statement on ongoing crisis of native family separation. aim of report is to 1) evaluate magnitude of change since ICWA, 2) ID geographic variation in inequality, 3) ID sites and places where inequality is accelerated

# lit review

1. historical importance of family separation
2. ICWA and policy
3. prior empirics on AIAN inequality

# Data

1. AAIA data
2. NCANDS
3. AFCARS FC
4. AFCARS Adoption
5. Population data

# Methods

0. age distribution from SEER onto pep
1. point-in-time comps
2. period lifetables (age-spec / lifetime)
3. conditional probs
4. imputation

# Findings

## Change since ICWA

```{r}
####################################################################################
############## ICWA figures
####################################################################################

### ICWA AAIA data STATES

icwa_fc<-icwa_fc %>% 
  filter(!is.na(fc))

### ICWA DATA NAT

icwa_nat<-icwa %>% 
  filter(!(is.na(fc_aian))) %>% 
  summarize(fc_aian = sum(fc_aian),
            fc_non = sum(fc_non),
            pop_aian21 = sum(pop_aian21),
            pop_non = sum(pop_nonAIAN21))

icwa_st<-icwa %>% 
  filter(!(is.na(fc_aian)))

##

afcars_caseload19<-read_csv("./data/afcars19_caseload.csv")
afcars_caseload19<-afcars_caseload19 %>% 
  mutate(race_ethn = ifelse(AMIAKN==1,
                            "AIAN",
                            "other")) %>% 
  rename(state = St,
         st_fips = STATE,
         fc = caseload) %>% 
  filter(st_fips!=72) %>% 
  select(st_fips, race_ethn, fc) %>% 
  filter(!is.na(race_ethn))

# afcars_caseload<-read_csv("./data/state_first_fc.csv")%>%
#   filter(.imp==1) %>% 
#   rename(year = fy) %>%
#   filter(state!=72) %>%
#   rename(var = first_entry) %>%
#   rename(st_fips = state) %>% 
#   filter(year == 2018) %>% 
#   mutate(race_ethn = ifelse(race_ethn=="AI/AN",
#                             "AIAN",
#                             "other")) %>% 
#   group_by(st_fips, race_ethn) %>% 
#   summarise(fc = sum(var)) %>% 
#   ungroup() %>% 
#   left_join(pop_seer_nat %>% 
#               filter(age<=21) %>% 
#               select(state, st_fips, age, race_ethn, pop) %>% 
#               group_by(state, st_fips, race_ethn) %>% 
#               summarize(pop = sum(pop)) %>% 
#               ungroup()) %>% 
#   filter(state%in%icwa_fc$state) %>% 
#   ungroup()

### add adoption data

### join with afcars adoption data
afcars_caseload<-afcars_caseload19 %>% 
  left_join(afcars_adopt_caseload_19 %>% 
  mutate(race_ethn = ifelse(race_ethn=="AIAN", 
                            race_ethn,
                            "other")) %>% 
  mutate(n_non_preferred = ifelse(race_ethn=="AIAN",
                                  n - n_kin_or_aian,
                                  n - n_kin),
         adopt_non_preferred = n_non_preferred / n) %>% 
    rename(st_fips = STATE,
           state = St,
           adopted = n) %>% 
  select(state, race_ethn, adopted, adopt_non_preferred)) %>% 
  mutate(state,  race_ethn, fc, adopted) %>% 
  mutate(period = "2019")

### state comparisons
icwa_delta<-icwa_fc %>% 
  select(-St) %>% 
  mutate(race_ethn = ifelse(race_ethn=="White",
                            "other",
                            race_ethn)) %>% 
  bind_rows(afcars_caseload %>% 
              select(state, period, race_ethn, fc, adopted)) %>% 
  filter(state%in%icwa_fc$state)

### make comparable national totals
### can't compare per cap rates
### pop denominators aren't comparable
```

```{r icwa-fc-counts, fig.cap = "Children in foster care on reporting date (caseload) by state and period. Data from AAIA surveys and AFCARS"}
ggplot(icwa_delta %>% 
         mutate(race_ethn = 
                  ifelse(race_ethn=="other",
                         "Non-AIAN",
                         race_ethn)),
       aes(y = state, x = fc, fill = period)) + 
  geom_col(position = position_dodge2(reverse = T)) + 
  facet_wrap(~race_ethn, scales = "free") + 
  scale_fill_brewer(palette = "Dark2") + 
  labs(y = "",
       x = "Children in foster care",
       fill = "") + 
  ggsave("./vis/cjlr/1.png", width = 8, height = 4)
```

```{r icwa-adopt-counts, fig.cap = "Children 21 and under in state-involved adoptions on reporting date by state and period. Data from AAIA surveys and AFCARS"}
  ggplot(icwa_delta %>% 
           mutate(race_ethn = 
                    ifelse(race_ethn=="other",
                           "Non-AIAN",
                           race_ethn)),
         aes(y = state, x = adopted, fill = period)) + 
    geom_col(position = position_dodge2(reverse = T)) + 
    facet_wrap(~race_ethn, scales = "free") + 
    scale_fill_brewer(palette = "Dark2") + 
    labs(y = "",
         x = "Children in adoption",
         fill = "") + 
  ggsave("./vis/cjlr/1_2.png", width = 8, height = 4)

#### make state-level proportional change over time 
### for both outcomes
```

```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}

icwa_delta_pct<-
  icwa_delta %>% 
  pivot_wider(id_cols = c(state, race_ethn),
              names_from = period, 
              values_from = c(fc, adopted)) %>% 
  mutate(delta_fc = (fc_2019 / fc_1976 -1) * 100,
         delta_adopt = (adopted_2019/adopted_1976 -1)*100) %>% 
  select(state, race_ethn, delta_fc, delta_adopt) %>% 
  pivot_longer(cols = delta_fc:delta_adopt) %>% 
  mutate(name = ifelse(
    name=="delta_fc",
    "Foster care",
    "Adoption"
  ))

### remove adoption states where we don't have AIAN ICWA numbers
index<-icwa_delta_pct %>% 
  filter(is.na(value))

icwa_delta_pct <- icwa_delta_pct %>% 
  mutate(value = ifelse(state%in%index$state & name == "Adoption",
         NA,
         value))

ggplot(icwa_delta_pct %>% 
         mutate(race_ethn =
                  ifelse(race_ethn == "AIAN",
                         race_ethn,
                         "Non-AIAN")),
       aes(x = value,
           y = state,
           fill = name)) + 
  geom_col(position = position_dodge2(reverse = T)) + 
  labs(y = "") +
  geom_vline(xintercept = 0, lty = 2) +
  scale_fill_brewer(palette = "Dark2") + 
  facet_wrap(~race_ethn, scales = "free") + 
  labs(fill = "", 
       x = "Percent increase, 1976 to 2019") +
  ggsave("./vis/cjlr/1_3.png", width = 8, height = 4)

icwa_nat<-icwa_delta %>% 
  filter(!state%in%index$state) %>% 
  group_by(period, race_ethn) %>% 
  summarise(fc = sum(fc),
            adopted = sum(adopted))

### plus boarding
total_inst<-icwa_delta %>% 
  group_by(period, race_ethn) %>% 
  summarise(fc = sum(fc),
            adopted = sum(adopted, na.rm=T)) %>% 
  mutate(total = fc + adopted)
```

1. National table from CJLR
2. State counts from CJLR
3. Percent change from CJLR

# National and state-level marginal probabilities

pull from data leaders slides

# Conditional probabilities

pull from data leaders slides

# Discussion

# Conclusion

### 2020-12-08 cfp_final_report.Rmd

@@
 This study connects contemporary trends in Native family separation to histories of Indian child removal, and locates places and institutional sites where inequality emerges."
-output: bookdown::html_document2
+output: bookdown::pdf_document2
 ---
@@
 
-```{r}
+```{r icwa-setup}
 ####################################################################################
@@
        x = "Children in foster care",
-       fill = "") + 
-  ggsave("./vis/cjlr/1.png", width = 8, height = 4)
+       fill = "") 
 ```
@@
          x = "Children in adoption",
-         fill = "") + 
-  ggsave("./vis/cjlr/1_2.png", width = 8, height = 4)
+         fill = "")
 
@@
   labs(fill = "", 
-       x = "Percent increase, 1976 to 2019") +
-  ggsave("./vis/cjlr/1_3.png", width = 8, height = 4)
+       x = "Percent increase, 1976 to 2019")
 

### 2020-12-09 cfp_final_report.Rmd
update icwa comp section

@@
 library(gridExtra)
+theme_set(theme_bw())
 
@@
 1. AAIA data
+
+As landmark data. Explain methods. Explain strengths, explain limitations. states included and selction criteria
+
 2. NCANDS
@@
 1. point-in-time comps
+
+AAIA compiled point-in-time estimates of the numbers of AIAN children 21 years of age and under in adoption or foster care in each state for which they were able to collect complete data through surveys of state agencies and private providers.
+We compare these numbers to counts of children in foster care at year-end and to total numbers of children in state-sponsored adoptions.
+Annual foster care counts are straightforward to compute with AFCARS, by restricting caseload counts to
+
+EXPLAIN VARIABLE YEAR OF DATA COLLECTION FOR AAIA
+
 2. period lifetables (age-spec / lifetime)
@@
 ### pop denominators aren't comparable
-```
-
-```{r icwa-fc-counts, fig.cap = "Children in foster care on reporting date (caseload) by state and period. Data from AAIA surveys and AFCARS"}
-ggplot(icwa_delta %>% 
-         mutate(race_ethn = 
-                  ifelse(race_ethn=="other",
-                         "Non-AIAN",
-                         race_ethn)),
-       aes(y = state, x = fc, fill = period)) + 
-  geom_col(position = position_dodge2(reverse = T)) + 
-  facet_wrap(~race_ethn, scales = "free") + 
-  scale_fill_brewer(palette = "Dark2") + 
-  labs(y = "",
-       x = "Children in foster care",
-       fill = "") 
-```
-
-```{r icwa-adopt-counts, fig.cap = "Children 21 and under in state-involved adoptions on reporting date by state and period. Data from AAIA surveys and AFCARS"}
-  ggplot(icwa_delta %>% 
-           mutate(race_ethn = 
-                    ifelse(race_ethn=="other",
-                           "Non-AIAN",
-                           race_ethn)),
-         aes(y = state, x = adopted, fill = period)) + 
-    geom_col(position = position_dodge2(reverse = T)) + 
-    facet_wrap(~race_ethn, scales = "free") + 
-    scale_fill_brewer(palette = "Dark2") + 
-    labs(y = "",
-         x = "Children in adoption",
-         fill = "")
-
-#### make state-level proportional change over time 
-### for both outcomes
-```
-
-```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}
 
@@
 
+icwa_nat<-icwa_delta %>% 
+  filter(!state%in%index$state) %>% 
+  group_by(period, race_ethn) %>% 
+  summarise(fc = sum(fc),
+            adopted = sum(adopted))
+
+### plus boarding
+total_inst<-icwa_delta %>% 
+  group_by(period, race_ethn) %>% 
+  summarise(fc = sum(fc),
+            adopted = sum(adopted, na.rm=T)) %>% 
+  mutate(total = fc + adopted)
+
+icwa_nat_na<-icwa_delta %>% 
+  group_by(period, race_ethn) %>% 
+  summarise(fc = sum(fc, na.rm=T),
+            adopted = sum(adopted, na.rm=T))
+```
+
+We compare data collected by the Association on American Indian Affairs between 1973 and 1976 on American Indian and Alaska Native children in foster care and adoption to contemporary data on AIAN children in foster care and adoption collected through the Adoption and Foster Care Analysis and Reporting System between 2010 and 2019.
+Figure \@ref(fig:icwa-fc-counts) shows the total numbers of children in foster care as reported by AAIA in the mid-1970s and in AFCARS in 2019.
+There were `r icwa_nat_na$fc[1]` Native children in the 19 states for which AAIA collected data in the mid-1970s.
+In 2019, there were `r icwa_nat_na$fc[3]` Native children in foster care, an increase of `r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,1)` AIAN children in foster care over this period.
+For non-Native children in these 19 states, caseloads increased by `r round(((icwa_nat_na$fc[4] / icwa_nat_na$fc[2]) - 1) * 100,1)` percent over the same period.
+
+```{r icwa-fc-counts, fig.cap = "Children in foster care on reporting date (caseload) by state and period. Data from AAIA surveys and AFCARS"}
+ggplot(icwa_delta %>% 
+         mutate(race_ethn = 
+                  ifelse(race_ethn=="other",
+                         "Non-AIAN",
+                         race_ethn)),
+       aes(y = state, x = fc, fill = period)) + 
+  geom_col(position = position_dodge2(reverse = T)) + 
+  facet_wrap(~race_ethn, scales = "free") + 
+  scale_fill_brewer(palette = "Dark2") + 
+  labs(y = "",
+       x = "Children in foster care",
+       fill = "") 
+```
+
+Figure \@ref(fig:icwa-adopt-counts) displays counts of children 21 years of age or younger in state-involved adoptions for both the mid-1970s and 2019.
+Note that 6 of the 19 included states lack complete data on AIAN children in adoption in the AAIA data (Idaho, Maine, Nevada, New Mexico, New York, and Wyoming).
+For these 13 states with complete adoption data, AAIA estimated that there were `r icwa_nat$adopted[1]` Native children in state-involved adoptions in the mid-1970s.
+In these same 13 states, there were `r icwa_nat$adopted[3]` AIAN children in adoptions in 2019, an increase of `r round(((icwa_nat_na$adopted[3] / icwa_nat_na$adopted[1]) - 1) * 100,1)` percent.
+For non-native children in these states, there was a `r round(abs(((icwa_nat_na$adopted[2] / icwa_nat_na$adopted[4]) - 1) * 100),1)` percent reduction in the numbers of adopted children over the same period.
+
+`r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,1)`
+
+```{r icwa-adopt-counts, fig.cap = "Children 21 and under in state-involved adoptions on reporting date by state and period. Data from AAIA surveys and AFCARS"}
+  ggplot(icwa_delta %>% 
+           mutate(race_ethn = 
+                    ifelse(race_ethn=="other",
+                           "Non-AIAN",
+                           race_ethn)),
+         aes(y = state, x = adopted, fill = period)) + 
+    geom_col(position = position_dodge2(reverse = T)) + 
+    facet_wrap(~race_ethn, scales = "free") + 
+    scale_fill_brewer(palette = "Dark2") + 
+    labs(y = "",
+         x = "Children in adoption",
+         fill = "")
+
+#### make state-level proportional change over time 
+### for both outcomes
+```
+
+Figure \@ref(fig:icwa-pct) shows the relative change in adoptive and foster care caseloads between the mid-1970s and 2019 for these large Native population states.
+Of the 19 states with complete data, 14 saw growth in the numbers of Native children in foster care since the passage of ICWA.
+Of the 13 states with complete adoption data, 8 had more Native children in foster care in 2019 than in the mid-1970s prior to the passage of ICWA.
+For non-Native children, foster care caseloads were higher in 16 of 19 states and adoption caseloads were higher in 6 of the 13 states with complete data.
+Oklahoma saw the largest absolute growth in Native foster care caseloads.
+In the mid-1970s, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 1976) %>% select(fc) ` Native chlidren in foster care in Oklahoma.
+In 2019, here were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 2019) %>% select(fc) ` Native children in foster care in Oklahoma, a growth of nearly `r round(icwa_delta_pct %>% filter(state=="OK", name == "Foster care", race_ethn == "AIAN") %>% select(value), 0)` percent.
+
+```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}
+
 ggplot(icwa_delta_pct %>% 
@@
 
-icwa_nat<-icwa_delta %>% 
-  filter(!state%in%index$state) %>% 
-  group_by(period, race_ethn) %>% 
-  summarise(fc = sum(fc),
-            adopted = sum(adopted))
-
-### plus boarding
-total_inst<-icwa_delta %>% 
-  group_by(period, race_ethn) %>% 
-  summarise(fc = sum(fc),
-            adopted = sum(adopted, na.rm=T)) %>% 
-  mutate(total = fc + adopted)
 ```
 
-1. National table from CJLR
-2. State counts from CJLR
-3. Percent change from CJLR
+```{r}
+#COMPUTE GROWTH IN AIAN - GROWTH IN NONAIAN FOR WHICH GREW FASTER.
+icwa_delta_aian<-icwa_delta_pct %>% 
+  filter(race_ethn=="AIAN") %>% 
+  select(-race_ethn) %>% 
+  rename(aian_value = value)
+icwa_delta_non<-icwa_delta_pct %>% 
+  filter(race_ethn!="AIAN") %>% 
+  select(-race_ethn) %>% 
+  rename(non_aian_value = value)
+
+icwa_pct_diff<-icwa_delta_aian %>% 
+  left_join(icwa_delta_non) %>% 
+  mutate(diff = aian_value - non_aian_value)
+
+tab_higher<-icwa_pct_diff %>% 
+  group_by(name) %>% 
+  summarise(aian_higher = sum(diff>0, na.rm=T))
+```
+
+Foster care caseloads grew in many states for both non-AIAN and AIAN children.
+Of the 19 states included here, 12 saw higher growth in AIAN foster care caseloads than they saw in non-AIAN fooster care caseloads.
+Adoption caseloads also grew faster for AIAN children in these states than they did for non-AIAN children.
+In 7 of 13 states with complete data, AIAN adoption caseloads grew faster between the mid-1970s and 2019 than did non-AIAN adoption caseloads.
+
+Of course, these data do not provide a complete comparative picture of Indian family separation over time.
+Until the late 20th century, American Indian and Alaska Native children were routinely removed from their families and sent to residential boarding schools.
+Approximately 25000 Native children were in Bureau of Indian Affairs boarding schools in 1973.
+This total is larger than the total number of Native children in foster care or adoption recorded by AAIA in 1976.
+There were more AIAN children in boarding schools in 1973 than were in foster care in 2019.
+While the scale of separation of Native children from their families through foster care and adoption has increased over time in these states, the absolute scale of the separation of Native children from their families has likely declined nationally with the end of the boarding school era.
 

### 2020-12-15 cfp_final_report.Rmd
start of day update

@@
 
-# National and state-level marginal probabilities
+## National and state-level marginal probabilities
+
+Figure \@ref(nat-marginal) shows the marginal probabilities of experiencing a series of child welfare system events for both Native and non-Native children.
+This plot shows national age-specific event probabilities at 2014-2018 risk levels for AIAN and non-AIAN children for experiencing 1) a screened-in investigation, 2) a confirmed maltreatment case, 3) a foster care placement, and 4) termination of parental rights through the child welfare system.
+DISCUSSION OF THIS.
+
+```{r nat-marginal, fig.cap = "Marginal probability of child welfare event incidence for AIAN children by age 18, 2014 - 2018 risk levels, US totals"}
+##############################################################
+###### National age-spec risk
+##############################################################
+
+# make 2 x 2 age-spec plots for each outcome
+
+## stack tables for faceting
+plot_dat<-inv_nat_tab %>% 
+  group_by(race_ethn, age) %>% 
+  summarise(qmax = max(q) , qmin=min(q) , q = mean(q),
+            c = mean(c), cmax = max(c), cmin = min(c)) %>% 
+  ungroup() %>% 
+  mutate(var = "Investigation") %>% 
+  bind_rows(
+    sub_nat_tab %>% 
+      group_by(race_ethn, age) %>% 
+      summarise(qmax = max(q) , qmin=min(q) , q = mean(q),
+                c = mean(c), cmax = max(c), cmin = min(c)) %>% 
+      ungroup() %>% 
+      mutate(var = "Substantiation")
+  ) %>% 
+  bind_rows(
+    fc_nat_tab %>% 
+      group_by(race_ethn, age) %>% 
+      summarise(qmax = max(q) , qmin=min(q) , q = mean(q),
+                c = mean(c), cmax = max(c), cmin = min(c)) %>% 
+      ungroup() %>% 
+      mutate(var = "Foster Care")
+  ) %>% 
+  bind_rows(
+    tpr_nat_tab %>% 
+      group_by(race_ethn, age) %>% 
+      summarise(qmax = max(q) , qmin=min(q) , q = mean(q),
+                c = mean(c), cmax = max(c), cmin = min(c)) %>% 
+      ungroup() %>% 
+      mutate(var = "Termination")
+  ) %>% 
+  mutate(var = factor(var,
+                      levels = c("Investigation",
+                                 "Substantiation",
+                                 "Foster Care",
+                                 "Termination")))
+
+ggplot(plot_dat,
+       aes(x = age, y = q, ymin = qmin, ymax = qmax,
+           color = race_ethn, fill = race_ethn)) + 
+  geom_line() + 
+  geom_ribbon(alpha = 0.5, color = NA) +
+  labs(color = "", fill = "") + 
+  labs(x = "Age", y = "Probability") + 
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2") + 
+  facet_wrap(~var)
+
+```
+
+THEN AGE-SPEC GEO FACET
+
+FOR INVESTIGATIONS
+
+```{r nat-inv-age, fig.cap = "Age specific risk of experiencing first CPS investigation by state, 2014 - 2018 risk levels"}
+ggplot(investigation_tables %>% 
+         group_by(state, race_ethn, age) %>% 
+         summarise(qmax = max(q_c), qmin=min(q_c), q = mean(q_c)),
+       aes(x = age, y = q, ymin = qmin, ymax = qmax,
+           color = race_ethn, fill = race_ethn)) + 
+  geom_line() + 
+  geom_ribbon(alpha = 0.5, color = NA) + 
+  facet_geo(~state) + 
+  labs(color = "", fill = "") + 
+  theme_minimal() + 
+  labs(x = "Age", y = "Proportion of children") + 
+  theme(strip.text.x = element_text(size = 7),
+        axis.text.x = element_text(size = 6),
+        axis.text.y = element_text(size = 4)) + 
+  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+
+```
+
+FOR FOSTER CARE
+
+```{r nat-fc-age, fig.cap = "Age specific probability of entering foster care, 2014 - 2018 risk levels"}
+ggplot(fc_tables %>% 
+         group_by(state, race_ethn, age) %>% 
+         summarise(qmax = max(q_c), qmin=min(q_c), q = mean(q_c)),
+       aes(x = age, y = q, ymin = qmin, ymax = qmax,
+           color = race_ethn, fill = race_ethn)) + 
+  geom_line() + 
+  geom_ribbon(alpha = 0.5, color = NA) + 
+  facet_geo(~state) + 
+  labs(color = "", fill = "") + 
+  theme_minimal() + 
+  labs(x = "Age", y = "Proportion of children") + 
+  theme(strip.text.x = element_text(size = 7),
+        axis.text.x = element_text(size = 6),
+        axis.text.y = element_text(size = 4)) + 
+  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) + 
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+```
+
+THEN HIGH RISK STATES FOR EACH OUTCOME FOR AGESPEC
+
+THEN CUMULATIVE MAP
+
+```{r cumul-map, fig.cap = "Cumulative risk of child welfare system event for AIAN children by age 18 at 2014 - 2018 risk levels"}
+#################################################################
+############# MAPS
+#################################################################
+### Lifetime risk
+
+map_dat<-investigation_tables_c %>% 
+  filter(race_ethn=="AIAN") %>% 
+  group_by(state) %>% 
+  summarize(c = mean(c_c))
+
+### xwalk state maps onto abbreviations
+us_map<-us_map()
+
+us_map<-us_map %>% 
+  left_join(map_dat %>% 
+              rename(abbr = state))
+
+inv_map<-ggplot(us_map %>% 
+         filter(!(is.na(var))),
+       aes(x = x, y = y, group = group,
+           fill = c)) + 
+  geom_polygon(color = "black") + 
+  coord_fixed() +
+  theme_void() + 
+  scale_fill_gradient2() +
+  theme(legend.position = "bottom") +
+  labs(fill = "Investigation") +
+  guides(fill = guide_colourbar(title.position = "top",
+                             label.position = "bottom",
+                             title.hjust = 0.5))
+
+map_dat<-subst_tables_c %>% 
+  filter(race_ethn=="AIAN") %>% 
+  group_by(state) %>% 
+  summarize(c = mean(c_c))
+
+us_map<-us_map()
+
+us_map<-us_map %>% 
+  left_join(map_dat %>% 
+              rename(abbr = state))
+
+sub_map<-ggplot(us_map %>% 
+                  filter(!(is.na(var))),
+                aes(x = x, y = y, group = group,
+                    fill = c)) + 
+  geom_polygon(color = "black") + 
+  coord_fixed() +
+  theme_void() + 
+  scale_fill_gradient2() +
+  theme(legend.position = "bottom") +
+  labs(fill = "Substantation") +
+  guides(fill = guide_colourbar(title.position = "top",
+                                label.position = "bottom",
+                                title.hjust = 0.5))
+
+map_dat<-fc_tables_c %>% 
+  filter(race_ethn=="AIAN") %>% 
+  group_by(state) %>% 
+  summarize(c = mean(c_c))
+
+us_map<-us_map()
+
+us_map<-us_map %>% 
+  left_join(map_dat %>% 
+              rename(abbr = state))
+
+fc_map<-ggplot(us_map %>% 
+                  filter(!(is.na(var))),
+                aes(x = x, y = y, group = group,
+                    fill = c)) + 
+  geom_polygon(color = "black") + 
+  coord_fixed() +
+  theme_void() + 
+  scale_fill_gradient2() +
+  theme(legend.position = "bottom") +
+  labs(fill = "Foster Care") +
+  guides(fill = guide_colourbar(title.position = "top",
+                                label.position = "bottom",
+                                title.hjust = 0.5))
+
+map_dat<-tpr_tables_c %>% 
+  filter(race_ethn=="AIAN") %>% 
+  group_by(state) %>% 
+  summarize(c = mean(c_c))
+
+us_map<-us_map()
+
+us_map<-us_map %>% 
+  left_join(map_dat %>% 
+              rename(abbr = state))
+
+tpr_map<-ggplot(us_map %>% 
+                 filter(!(is.na(var))),
+               aes(x = x, y = y, group = group,
+                   fill = c)) + 
+  geom_polygon(color = "black") + 
+  coord_fixed() +
+  theme_void() + 
+  scale_fill_gradient2() +
+  theme(legend.position = "bottom") +
+  labs(fill = "TPR") +
+  guides(fill = guide_colourbar(title.position = "top",
+                                label.position = "bottom",
+                                title.hjust = 0.5))
+
+c_map<-grid.arrange(inv_map, sub_map,
+                    fc_map, tpr_map)
+
+```
+
+THEN INEQ MAP
+
+```{r}
+##################
+#### #### #### #### 
+#### rate ratio for inequality
+#### #### #### #### 
+### xwalk state maps onto abbreviations
+us_map<-us_map()
+
+us_map<-us_map %>% 
+  left_join(map_dat %>% 
+              rename(abbr = state))
+
+#### DO OTHER MAPS HERE
+
+map_dat<-investigation_tables_c %>% 
+  group_by(state, race_ethn) %>% 
+  filter(c == min(c)) %>% 
+  mutate(var = "Investigation") %>% 
+  bind_rows(subst_tables_c %>% 
+              group_by(state, race_ethn) %>% 
+              filter(c == min(c)) %>% 
+              mutate(var = "Substantiation")) %>% 
+  bind_rows(fc_tables_c %>% 
+              group_by(state, race_ethn) %>% 
+              filter(c == min(c)) %>% 
+              mutate(var = "Foster Care")) %>% 
+  bind_rows(tpr_tables_c %>% 
+              group_by(state, race_ethn) %>% 
+              filter(c == min(c)) %>% 
+              mutate(var = "TPR")) %>% 
+  select(-.imp) %>% 
+  distinct()
+
+map_dat_wht<-map_dat %>% 
+  ungroup() %>% 
+  filter(race_ethn == "White") %>% 
+  rename(c_white = c_c) %>% 
+  select(-race_ethn, -c, -race_ethn)
+
+map_dat<-map_dat %>% 
+  filter(race_ethn=="AIAN") %>% 
+  left_join(map_dat_wht) %>% 
+  mutate(disp = c_c / c_white)
+
+### slice into quantiles
+### xwalk state maps onto abbreviations
+us_map<-us_map()
+
+us_map<-us_map %>% 
+  left_join(map_dat %>% 
+              rename(abbr = state)) %>% 
+  mutate(var = factor(var,
+                      levels = c(
+                        "Investigation", 
+                        "Substantiation",
+                        "Foster Care",
+                        "TPR"
+                      )))
+
+### cut into an ordered categorical
+
+us_map<-us_map %>% 
+  mutate(rate_ratio = 
+           case_when(
+             disp<0.5 ~ "0-0.5",
+             disp<1 ~ "0.5-1",
+             disp<1.5 ~ "1-1.5",
+             disp<2 ~ "1.5-2",
+             disp<3 ~ "2-3",
+             disp<4 ~ "3-4",
+             disp>=4 ~ "4+",
+           ),
+         rate_ratio = factor(rate_ratio,
+                             levels = c("0-0.5",
+                                        "0.5-1",
+                                        "1-1.5",
+                                        "1.5-2",
+                                        "2-3",
+                                        "3-4",
+                                        "4+")))
+
+ggplot(us_map,
+       aes(x = x, y = y, group = group,
+           fill = rate_ratio)) + 
+  geom_polygon(color = "black") + 
+  coord_fixed() +
+  facet_wrap(~var) + 
+  theme_void() + 
+  scale_fill_brewer(palette = "Purples") +
+  labs(fill = "AIAN / White\nRate Ratio")
+```
 
@@
 # Conditional probabilities
+
+```{r}
+##############################################################
+###### National conditionals
+##############################################################
+
+# make 2 x 2 age-spec plots for each conditional
+
+plot_dat<-cond_sub_inv %>% 
+  select(.imp, age, race_ethn, cond, var) %>% 
+  bind_rows(
+    cond_fc_inv %>% 
+      select(.imp, age, race_ethn, cond, var)) %>% 
+  bind_rows(cond_fc_sub %>% 
+              select(.imp, age, race_ethn, cond, var)) %>% 
+  bind_rows(
+    cond_tpr_fc %>% 
+      select(.imp, age, race_ethn, cond, var)
+  ) %>% 
+  group_by(age, race_ethn, var) %>% 
+  summarise(q = mean(cond),
+            qmin = min(cond),
+            qmax = max(cond))
+
+plot_dat<-plot_dat %>% 
+  mutate(var = case_when(
+    var == "P(FC|inv)" ~ "2. Foster care after investigation",
+    var == "P(FC|sub)" ~ "3. Foster care after substantiation",
+    var == "P(sub|inv)" ~ "1. Substantiation after investigation",
+    var == "P(TPR|FC)"  ~ "4. Termination after foster care"
+  ))
+
+ggplot(plot_dat,
+       aes(x = age, y = q, ymin = qmin, ymax = qmax,
+           fill = race_ethn, color = race_ethn)) + 
+  geom_line() + 
+  geom_ribbon(alpha = 0.5, color = NA) + 
+  facet_wrap(~factor(var)) + 
+  labs(color = "", fill = "", x = "Age", y = "Conditional probability") + 
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2") 
+```
 

### 2020-12-16 main_state.R
revise for state single year age data

@@ -1,614 +1,0 @@
-### make state tables
-
-source("lifetable.r")
-
-### CHECKING THE CENSUS PEP ALONE/COMB DATA FOR THIS PERIOD
-pop_new<-read_csv("./data/cc-est2019-alldata.csv")
-### code age groups
-pop_new<-pop_new %>% 
-  mutate(age = case_when(
-    AGEGRP==0 ~ "Total",
-    AGEGRP==1 ~ "0-4",
-    AGEGRP==2 ~ "5-9",
-    AGEGRP==3 ~ "10-14",
-    AGEGRP==4 ~ "15-19",
-    AGEGRP>4 ~ "20+"
-  ))
-
-pop_new<-pop_new %>% 
-  mutate(aian = NHIAC_MALE + NHIAC_FEMALE + 
-           HIAC_MALE + HIAC_FEMALE)
-
-pop_new<-pop_new %>% 
-  select(STATE, COUNTY, STNAME, CTYNAME, 
-         YEAR, AGEGRP, TOT_POP, AGEGRP,
-         age, aian) %>% 
-  filter(YEAR>2) %>% 
-  filter(age!="Total") %>% 
-  mutate(year = YEAR + 2007) %>% 
-  select(-YEAR, -AGEGRP) %>% 
-  group_by(STATE, year, age) %>% 
-  summarise(pop = sum(aian)) %>% 
-  mutate(race_ethn = "AIAN")
-
-pop<-read_fwf("~/Projects/cps_lifetables/data/us.1990_2018.singleages.adjusted.txt",
-              fwf_widths(c(4, 2, 2, 3, 2, 1, 1, 1, 2, 8),
-                         c("year", "state", "st_fips",
-                           "cnty_fips", "reg", "race",
-                           "hisp", "sex", "age", "pop")),
-              col_types = "iccciiiiii")
-
-pop_seer_nat<-pop %>%
-  filter(year==2018)%>% 
-  mutate(pop = as.integer(pop),
-         age = as.integer(age),
-         race_ethn =
-           case_when(
-             race==3 ~ "AIAN"),
-         race_ethn = ifelse(is.na(race_ethn),
-                            "other",
-                            race_ethn))%>%
-  group_by(year,state, st_fips, age, race_ethn) %>%
-  summarise(pop = sum(pop)) %>%
-  ungroup() %>%
-  mutate(st_fips = as.numeric(st_fips))
-
-
-pop<-pop %>%
[…]

### 2020-12-16 cfp_final_report.Rmd
fast forward

@@
 library(tidyverse, warn.conflicts = FALSE)
-options(dplyr.summarise.inform = FALSE)
+options(dplyr.summarise.inform = FALSE,
+        knitr.kable.NA = " ")
 
@@
 
-1. AAIA data
+This study relies on a number of data sources.
+First, we compile and digitize statistics collected by the Association for American Indian Affairs during the late 1960s and early to mid 1970s in an effort to document the breadth and depth of the Indian family separation crisis.
+To capture the contemporary exposure of AIAN children to the child welfare system, we rely on federal data systems collected by the US Children's Bureau.
+We use the 2014 - 2018 National Child Abuse and Neglect Data System (NCANDS) child files (CITE), the 2014 - 2018 Adoption and Foster Care Analysis and Reporting System (AFCARS) foster care files (CITE) and the 2010 - 2018 AFCARS adoption files (CITE).
+We use two data sources for contemporary population estimates: the US Census Bureau's Population Estimates Program (PEP) (CITE), and the National Institute of Healths' SEER population estimates (CITE).
 
-As landmark data. Explain methods. Explain strengths, explain limitations. states included and selction criteria
+## Historical child welfare system data
 
-2. NCANDS
-3. AFCARS FC
-4. AFCARS Adoption
+The Association on American Indian Affairs submitted a report titled "Indian Child Welfare Statistical Summary, July 1976" in support of the Indian Child Welfare Act (1978) during congressional hearings, included as Appendix G in hearings convened by the Select Committee on Indian Affairs on August 4, 1977.
+The results presented by AAIA were the result of multiple waves of surveys sent by AAIA to federal, state, local, and private service providers in 19 states with large AIAN populations (Alaska, Arizona, California, Idaho, Maine, Michigan, Minnesota, Montana, Nevada, New Mexico, New York, North Dakota, Oklahoma, Oregon, South Dakota, Utah, Washington, Wisconsin, and Wyoming).
+AAIA compiled statistics on the characteristics of both Native and non-Native children in foster care, in adoption, or in boarding programs for each of these states.
+
+These data consistently provide point-in-time estimates of the numbers of Native and non-Native children in foster care, adoptions, or in Bureau of Indian Affairs boarding schools.
+During this period, some states did not collect information on race/ethnicity for children in foster care.
+When this was the case, AAIA imputed foster care caseload totals for Native and non-Native children by multiplying the total volume of children in foster care by the proportion of the population that was Native and non-Native respectively.
+This approach (as AAIA) acknowledges, likely produced conservative estimates of AIAN foster care caseloads in these states, as Native children were more likely than non-Native children to enter foster care in these states during this period.
+
+These landmark data provided a first close-to-comprehensive national picture of the scale of Indian family separation by combining federal, state, and local data sources.
+These data provided much of the quantitative empirical foundation for the landmark reforms implemented in the Indian Child Welfare Act.
+While AAIA's imputation methods and geographic scope do not match contemporary data collection procedures, these data provide the most comprehensive historical data available to examine the scope of Indian family separation in the period immediately prior to the passage of the Indian Child Welfare Act.
+
+## Contemporary child welfare system data
+
+The federal government maintains a series of data sets documenting the operations of contemporary state and local child welfare systems.
+The National Child Abuse and Neglect Data System is a voluntary reporting system that documents all cases of alleged child maltreatment reported to a state or local child welfare hotline that are screened-in and receive an agency response (typically an investigation from a caseworker).
+Since the early 2010s, all 50 states have submitted these report-level data annually to the US Children's Bureau.
+Here, we use the NCANDS data to measure two events for AIAN (defined as American Indian or Alaska Native alone or in combination with any other racial or ethnic identification) children and white children (defined as non-hispanic white alone).
+Using within-state unique child identifiers, we document the first time a child was the subject of a screened-in CPS report, and the first time a child was the subject of a confirmed or substantiated CPS report over a five-year period (2014 - 2018).
+
+To measure the frequency of family separation into the foster care system across states, we use the Adoption and Foster Care Analysis and Reporting System foster care files for 2014 - 2018.
+All states are required to submit foster care and adoption data for the AFCARS.
+The AFCARS provides a single row of data for each child for each year that child was in foster care for a partial or complete year.
+Details on placement settings are only recorded for the last placement a child was in during a reporting period.
+We use unique child identifiers in the AFCARS foster care data (that match unique identifiers in the NCANDS data) to identify the first time a child entered the foster care system, and to track characteristics of placements that child was in over time.
+These placement variables are limited by the scope of AFCARS.
+If a child was in multiple placement settings during the course of a year, only information about the final placement setting is recorded in the AFCARS foster care file.
+Thus, statistics on the prevalence of particular placement settings should be interpreted as conservative.
+
+The AFCARS adoption file provides detailed information on all children newly adopted with state or local child welfare agency involvement in the US each year.
+To provide point-in-time estimates of total numbers of children in adoptive homes, we aggregate data from 2010 - 2019, then evaluate the number of both Native and non-Native children in adoptive homes who would have been 21 years old or younger in 2019.
+We use the 21-year-old threshold for comparability with data collected by AAIA.
+
+## Population data
+
+To obtain valid estimates of the AIAN child population by age for US states between 2014 and 2018, we rely on two data sources.
+Our definition of American Indian and Alaska Native includes any individual who identifies as American Indian or Alaska Native alone or in combination with any other group.
+The US Census Bureau's Population and Housing Unit Estimates program uses baseline data from the 2010 Census along with various sources of demographic data to estimate age and race specific population totals for all counties in the US annually.
+It
+
+SEER POP DATA
+
+ALONE OR IN COMBINATION PROBLEMS WITH BRIDGED ESTIMATES. CITE DEMOGRAPHY ON SHIFTS IN BASE POPULATION.
+
+PEP POP DATA
+
 5. Population data
@@
 
-# Findings
-
-## Change since ICWA
+# Findings:Change since ICWA
 
 ```{r icwa-setup}
-####################################################################################
-############## ICWA figures
-####################################################################################
 
@@
 
+\newpage
+
 ```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}
@@
        x = "Percent increase, 1976 to 2019")
-
 ```
 
-```{r}
+\newpage
+
+```{r icwa-growth}
 #COMPUTE GROWTH IN AIAN - GROWTH IN NONAIAN FOR WHICH GREW FASTER.
@@
 
-## National and state-level marginal probabilities
+\newpage
+
+# Findings:Contemporary risks of child welfare system contact
+
+INTRO WITH MOTIVATION - INCREASING SERIOUSNESS OF CONTACT, POSSIBILITY OF ESTIMATING POINTS OF INEQUALITY ACCELERATION
+
+## Age-specific risks
 
@@
 ```{r nat-marginal, fig.cap = "Marginal probability of child welfare event incidence for AIAN children by age 18, 2014 - 2018 risk levels, US totals"}
-##############################################################
-###### National age-spec risk
-##############################################################
-
-# make 2 x 2 age-spec plots for each outcome
 
@@
 
-THEN AGE-SPEC GEO FACET
-
-FOR INVESTIGATIONS
+Of course, there is substantial variation in risk of child welfare system contact across US states.
+EXPLAIN WHY THIS IS.
+Figure\@ref(fig:nat-inv-age) shows the age-specific risk of experiencing a first screened-in investigation by US state for both AIAN and white children at 2014 - 2018 levels of risk.
+DESCRIBE FINDINGS
 
@@
 
-FOR FOSTER CARE
+There are also dramatic differences in risk of experiencing foster care across states.
+Figure \@ref(fig:nat-fc-age) shows the distribution of age-specific risks of first foster care entries for Native and white children by US state at 2014 - 2018 levels of risk.
+DESCRIBE FINDINGS
 
@@
 
-THEN CUMULATIVE MAP
+ADD PLOTS HERE
+
+## Lifetime event risks and inequalities in lifetime risk
+
+Figure \@ref(fig:cumul-map) displays the AIAN childrens' risk of ever experiencing each included child welfare system event by their age 18 at 2014 - 2018 levels of risk. DESCRIBE FINDINGS HERE
 
 ```{r cumul-map, fig.cap = "Cumulative risk of child welfare system event for AIAN children by age 18 at 2014 - 2018 risk levels"}
-#################################################################
-############# MAPS
-#################################################################
-### Lifetime risk
 
@@
 
-THEN INEQ MAP
+Figure \@ref(fig:ineq-map) shows the ratio of AIAN lifetime event risk to white lifetime event risk for each state. DISCUSS FINDINGS HERE.
 
-```{r}
-##################
-#### #### #### #### 
-#### rate ratio for inequality
-#### #### #### #### 
-### xwalk state maps onto abbreviations
-us_map<-us_map()
-
-us_map<-us_map %>% 
-  left_join(map_dat %>% 
-              rename(abbr = state))
+```{r ineq-map, fig.cap = "Inequality in cumulative risk of child welfare system event for AIAN children by age 18 at 2014 - 2018 risk levels"}
 
@@
   select(-.imp) %>% 
-  distinct()
+  distinct() %>% 
+  ungroup() %>% 
+  complete(state, race_ethn, var)
 
@@
   theme_void() + 
-  scale_fill_brewer(palette = "Purples") +
+  scale_fill_brewer(palette = "Purples", na.value = "grey45") +
   labs(fill = "AIAN / White\nRate Ratio")
@@
 
-pull from data leaders slides
+SUMMARY GRAF ON MARGINAL PROBABILITIES (AGE-SPEC AND CUMULATIVE)
 
-# Conditional probabilities
+\newpage
 
-```{r}
-##############################################################
-###### National conditionals
-##############################################################
+# Findings: conditional probability of subsequent child welfare events
+
+Table \@ref(tab:mn-demo) show some cool things. EXPLAIN WHAT THESE MEASURE AND HOW THEY ARE COMPUTING
+
+```{r mn-demo, results = "asis"}
+
+mn<-investigation_tables %>% 
+  filter(.imp==1, age == 0, state == "MN") %>% 
+  select(var, race_ethn, pop_max, q_c) %>% 
+  mutate(name = "Investigations") %>% 
+  bind_rows(
+  subst_tables %>% 
+  filter(.imp==1, age == 0,  state == "MN") %>% 
+  select(var, race_ethn, pop_max, q_c) %>% 
+  mutate(name = "Confirmed maltreatment"))
+
+#### make descriptive table
+mn<-mn %>% 
+  filter(race_ethn=="AIAN") %>% 
+  mutate("Number of children" = round(var/5,0),
+         "Child population" = round(pop_max/5,0),
+         "Event rate" = round(`Number of children`/ `Child population`,2)) %>% 
+  select(name, `Number of children`, `Child population`, `Event rate`)
+
+mn<-mn%>% 
+  mutate("Transition rate" = c(NA, round(mn$`Event rate`[2]/mn$`Event rate`[1],2))) %>% 
+  rename(" " = name)
+
+knitr::kable(mn,
+             caption = "How cases move upstream in Minnesota: AIAN infants",
+             )
+```
+
+To identify sites of institutional decision making where inequalities in AIAN child welfare system case processing emerge or are accelerated, we compute conditional event probabilities for 1) case substantiation, conditional on investigation; 2) foster care placement, conditional on investigation; 3) foster care placement, conditional on substantiation; and 4) termination of parental rights after foster care placement for both AIAN and white children.
+Figure \@ref(fig:nat-cond) displays the national risk levels for each event.
+
+DISCUSS FINDINGS
+
+```{r nat-cond, fig.cap = "Risk of child welfare system event, conditional on prior system event at 2014 - 2018 levels of risk"}
 
@@
 
-pull from data leaders slides
+STATE LEVEL FINDINGS
+
+```{r sub-inv}
+ncands_sub_inv_tab<-subst_tables %>%
+  rename(sub = q_c) %>% 
+  select(.imp , age, race_ethn, state, sub) %>% 
+  left_join(investigation_tables %>% 
+              rename(inv = q_c) %>% 
+              select(.imp, age, race_ethn, state, inv)) %>% 
+  mutate(sub_inv = sub / inv) %>% 
+  filter(sub_inv<1,
+         !(is.na(sub_inv)))
+
+ggplot(ncands_sub_inv_tab %>% 
+  group_by(age, race_ethn, state) %>% 
+    summarise(qmin = min(sub_inv),
+              qmax = max(sub_inv),
+              q = mean(sub_inv)),
+  aes(x = age, y = q, color = race_ethn,
+      ymin = qmin, ymax = qmax, fill = race_ethn)) +
+  geom_line() + 
+  geom_ribbon(alpha = 0.5,
+              color = NA) + 
+  labs(fill = "", color = "",
+       y = "Probability", x = "Age") + 
+  facet_geo(~state) +
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+```
+
+```{r fc-inv}
+fc_cond_inv_tabs<-fc_post_inv_tables %>%
+  rename(fc_inv = q_c) %>% 
+  select(.imp,age,  race_ethn, state, fc_inv) %>% 
+  left_join(investigation_tables %>% 
+              rename(inv = q_c) %>% 
+              select(.imp, age, race_ethn, state, inv)) %>% 
+  mutate(fc_cond_inv = fc_inv / inv)
+
+ggplot(fc_cond_inv_tabs %>% 
+         group_by(age, race_ethn, state) %>% 
+         summarise(qmin = min(fc_cond_inv),
+                   qmax = max(fc_cond_inv),
+                   q = mean(fc_cond_inv)) %>% 
+         filter(!(is.na(q)),
+                q<1),
+       aes(x = age, y = q, color = race_ethn,
+           ymin = qmin, ymax = qmax, fill = race_ethn)) +
+  geom_line() + 
+  geom_ribbon(alpha = 0.5,
+              color = NA) + 
+  labs(fill = "", color = "",
+       y = "Probability", x = "Age") + 
+  facet_geo(~state) +
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+```
+
+```{r fc-sub}
+
+fc_cond_sub_tabs<-fc_sub_tables %>%
+  rename(fc_sub = q_c) %>% 
+  select(.imp,age,  race_ethn, state, fc_sub) %>% 
+  left_join(subst_tables %>% 
+              rename(sub = q_c) %>% 
+              select(.imp, age, race_ethn, state, sub)) %>% 
+  mutate(fc_cond_sub = fc_sub / sub)
+
+ggplot(fc_cond_sub_tabs %>% 
+         group_by(age, race_ethn, state) %>% 
+         summarise(qmin = min(fc_cond_sub),
+                   qmax = max(fc_cond_sub),
+                   q = mean(fc_cond_sub)) %>% 
+         filter(!(is.na(q)),
+                q<1),
+       aes(x = age, y = q, color = race_ethn,
+           ymin = qmin, ymax = qmax, fill = race_ethn)) +
+  geom_line() + 
+  geom_ribbon(alpha = 0.5,
+              color = NA) + 
+  labs(fill = "", color = "",
+       y = "Probability", x = "Age") + 
+  facet_geo(~state) +
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+```
+
+```{r non-icwa}
+non_icwa_cond_fc<-non_icwa_tables %>%
+  rename(non_icwa = q_c) %>% 
+  select(.imp, age, race_ethn, state, non_icwa) %>% 
+  left_join(fc_tables%>% 
+              rename(fc = q_c) %>% 
+              select(.imp, age, race_ethn, state, fc)) %>% 
+  mutate(non_icwa_fc = non_icwa / fc)
+
+ggplot(non_icwa_cond_fc %>% 
+         filter(race_ethn=="AIAN") %>% 
+         group_by(age, state) %>% 
+         summarise(qmin = min(non_icwa_fc),
+                   qmax = max(non_icwa_fc),
+                   q = mean(non_icwa_fc)) %>% 
+         filter(!(is.na(q)),
+                q<=1, qmax<=1),
+       aes(x = age, y = q,
+           ymin = qmin, ymax = qmax)) + 
+  geom_line() +
+  geom_ribbon(alpha = 0.5, color = NA) + 
+  facet_geo(~state) + 
+  labs(x = "Age", y = "Proportion") 
+```
+
+\newpage
 

### 2020-12-16 make_state_tables.R

@@ -3,117 +3,71 @@
 source("lifetable.r")
 
 ### CHECKING THE CENSUS PEP ALONE/COMB DATA FOR THIS PERIOD
-pop_new<-read_csv("./data/cc-est2019-alldata.csv")
-### code age groups
-pop_new<-pop_new %>% 
-  mutate(age = case_when(
-    AGEGRP==0 ~ "Total",
-    AGEGRP==1 ~ "0-4",
-    AGEGRP==2 ~ "5-9",
-    AGEGRP==3 ~ "10-14",
-    AGEGRP==4 ~ "15-19",
-    AGEGRP>4 ~ "20+"
-  ))
+### previously used county with age group data "cc-est2019-alldata.csv"
+### 12/16, updating to state file with single ages: "sc-est2019-alldata5.csv"
+### for alone or in combination
+### and alldata6 for alone (white non-hispanic)
+pop_new<-read_csv("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/asrh/sc-est2019-alldata5.csv")
 
-pop_new<-pop_new %>% 
-  mutate(aian = NHIAC_MALE + NHIAC_FEMALE + 
-           HIAC_MALE + HIAC_FEMALE)
-
-pop_new<-pop_new %>% 
-  select(STATE, COUNTY, STNAME, CTYNAME, 
-         YEAR, AGEGRP, TOT_POP, AGEGRP,
-         age, aian) %>% 
-  filter(YEAR>2) %>% 
-  filter(age!="Total") %>% 
-  mutate(year = YEAR + 2007) %>% 
-  select(-YEAR, -AGEGRP) %>% 
-  group_by(STATE, year, age) %>% 
-  summarise(pop = sum(aian)) %>% 
+### obtain AIAN alone or in combination
+pop_aian<-pop_new %>% 
+  filter(RACE==3 & ORIGIN == 0 & SEX == 0) %>% 
+  select(-SUMLEV, -REGION, -DIVISION, -CENSUS2010POP, -ESTIMATESBASE2010) %>% 
+  pivot_longer(cols=POPESTIMATE2010:POPESTIMATE2019,
+               names_to = "YEAR"
+               ) %>% 
+  mutate(YEAR = str_sub(YEAR, -4, -1)) %>% 
+  group_by(STATE, NAME, AGE, YEAR) %>% 
+  summarise(pop = sum(value)) %>% 
[…]

### 2020-12-16 cfp_final_report.Rmd
fix icwa with boarding

@@
 
-# afcars_caseload<-read_csv("./data/state_first_fc.csv")%>%
-#   filter(.imp==1) %>% 
-#   rename(year = fy) %>%
-#   filter(state!=72) %>%
-#   rename(var = first_entry) %>%
-#   rename(st_fips = state) %>% 
-#   filter(year == 2018) %>% 
-#   mutate(race_ethn = ifelse(race_ethn=="AI/AN",
-#                             "AIAN",
-#                             "other")) %>% 
-#   group_by(st_fips, race_ethn) %>% 
-#   summarise(fc = sum(var)) %>% 
-#   ungroup() %>% 
-#   left_join(pop_seer_nat %>% 
-#               filter(age<=21) %>% 
-#               select(state, st_fips, age, race_ethn, pop) %>% 
-#               group_by(state, st_fips, race_ethn) %>% 
-#               summarize(pop = sum(pop)) %>% 
-#               ungroup()) %>% 
-#   filter(state%in%icwa_fc$state) %>% 
-#   ungroup()
-
-### add adoption data
-
 ### join with afcars adoption data
@@
 icwa_delta<-icwa_fc %>% 
-  select(-St) %>% 
   mutate(race_ethn = ifelse(race_ethn=="White",

### 2020-12-16 make_national_tables.r

@@ -20,11 +20,9 @@ for(i in 1:length(imps)){
     temp<-temp %>% 
       group_by(.imp, age, race_ethn) %>% 
       summarise(var = sum(var), 
-                pop = sum(pop),
-                pop_max = sum(pop_max)) %>% 
+                pop = sum(pop)) %>% 
       ungroup()
-    inv_nat_tab[[index]]<-make_life_table(temp %>% 
-                                            mutate(pop = pop_max))
+    inv_nat_tab[[index]]<-make_life_table(temp)
     index<-index+1
   }
 }
@@ -46,11 +44,9 @@ for(i in 1:length(imps)){
     temp<-temp %>% 
       group_by(.imp, age, race_ethn) %>% 
       summarise(var = sum(var), 
-                pop = sum(pop),
-                pop_max = sum(pop_max)) %>% 
+                pop = sum(pop)) %>% 
       ungroup()
-    sub_nat_tab[[index]]<-make_life_table(temp %>% 
-                                            mutate(pop = pop_max))
+    sub_nat_tab[[index]]<-make_life_table(temp)
     index<-index+1
   }
 }
@@ -72,11 +68,9 @@ for(i in 1:length(imps)){
     temp<-temp %>% 
       group_by(.imp, age, race_ethn) %>% 
       summarise(var = sum(var), 
-                pop = sum(pop),
-                pop_max = sum(pop_max)) %>% 
+                pop = sum(pop)) %>% 
       ungroup()
-    fc_nat_tab[[index]]<-make_life_table(temp %>% 
-                                           mutate(pop = pop_max))
+    fc_nat_tab[[index]]<-make_life_table(temp)
     index<-index+1
   }
 }
@@ -98,11 +92,9 @@ for(i in 1:length(imps)){
     temp<-temp %>% 
       group_by(.imp, age, race_ethn) %>% 
       summarise(var = sum(var), 
-                pop = sum(pop),
-                pop_max = sum(pop_max)) %>% 
+                pop = sum(pop)) %>% 
       ungroup()
-    tpr_nat_tab[[index]]<-make_life_table(temp %>% 
-                                            mutate(pop = pop_max))
+    tpr_nat_tab[[index]]<-make_life_table(temp)
     index<-index+1
   }
 }
@@ -144,11 +136,9 @@ for(i in 1:length(imps)){
     temp<-temp %>% 
       group_by(.imp, age, race_ethn) %>% 
       summarise(var = sum(var), 
[…]

### 2020-12-16 make_state_tables.R

@@ -436,21 +436,47 @@ tpr_cond<-tpr_tables_c%>%
 ## AAIA ICWA data
 
 icwa<-read_csv("./data/icwa_data.csv") 
+
 ### add state abbrev
 xwalk<-data.frame(State = as.character(state.name), 
-                  state.abb = as.character(state.abb))
-icwa<-icwa %>% 
-  left_join(xwalk)
-
-icwa<-icwa %>% 
+                  state.abb = as.character(state.abb)) %>% 
   mutate(AIAN = fc_aian,
          White = fc_non)
 
+icwa_adopt<-icwa %>% 
+  left_join(xwalk) %>% 
+  mutate(AIAN = adopted_aian,
+         White = adopted_non)
+
+icwa_boarding<-icwa %>% 
+  left_join(xwalk) %>% 
+  mutate(AIAN = bia_boarding74,
+         White = NA)
+
 icwa_fc<-icwa %>% 
-  select(state.abb, AIAN, White) %>% 
-  rename(state = state.abb) %>% 
+  left_join(xwalk) %>% 
+  ungroup() %>% 
+  mutate(AIAN = fc_aian,
+         White = fc_non) %>% 
+  select(State, state.abb, AIAN, White) %>% 
+  rename(state = state.abb,
+         St = State) %>% 
   pivot_longer(col = AIAN:White, 
                names_to = "race_ethn",
                values_to = "fc") %>% 
-  mutate(period = "1976") 
+  mutate(period = "1976") %>% 
+  left_join(icwa_adopt %>% 
+              select(State, state.abb, AIAN, White) %>% 
+              rename(state = state.abb,
+                     St = State) %>% 
+              pivot_longer(col = AIAN:White, 
+                           names_to = "race_ethn",
+                           values_to = "adopted")) %>% 
+  left_join(icwa_boarding %>% 
+              select(State, state.abb, AIAN, White) %>% 
+              rename(state = state.abb,
+                     St = State) %>% 
+              pivot_longer(col = AIAN:White, 
+                           names_to = "race_ethn",
+                           values_to = "boarding"))
 

### 2020-12-16 cfp_final_report.Rmd
update with rmd changes

@@
 ```{r read-make}
-#### make data for easy load
+# #### make data for easy load
 # source("make_state_tables.R")
@@
 
-### ICWA AAIA data STATES
-
-icwa_fc<-icwa_fc %>% 
-  filter(!is.na(fc))
-
 ### ICWA DATA NAT
 
-icwa_nat<-icwa %>% 
-  filter(!(is.na(fc_aian))) %>% 
-  summarize(fc_aian = sum(fc_aian),
-            fc_non = sum(fc_non),
-            pop_aian21 = sum(pop_aian21),
-            pop_non = sum(pop_nonAIAN21))
-
-icwa_st<-icwa %>% 
-  filter(!(is.na(fc_aian)))
+icwa_nat<-icwa_fc %>%
+  group_by(race_ethn) %>% 
+  summarize(fc = sum(fc, na.rm=T),
+            adopted = sum(adopted, na.rm=T),
+            boarding = sum(boarding, na.rm=T))
 
@@
          group_by(state, race_ethn, age) %>% 
-         summarise(qmax = max(q_c), qmin=min(q_c), q = mean(q_c)),
+         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
        aes(x = age, y = q, ymin = qmin, ymax = qmax,
@@
   group_by(state) %>% 
-  summarize(c = mean(c_c))
+  summarize(c = mean(c))
 
@@
   filter(race_ethn == "White") %>% 
-  rename(c_white = c_c) %>% 
-  select(-race_ethn, -c, -race_ethn)
+  rename(c_white = c) %>% 
+  select(-race_ethn)
 
@@
   left_join(map_dat_wht) %>% 
-  mutate(disp = c_c / c_white)
+  mutate(disp = c / c_white)
 
@@
   filter(.imp==1, age == 0, state == "MN") %>% 
-  select(var, race_ethn, pop_max, q_c) %>% 
+  select(var, race_ethn, pop, q) %>% 
   mutate(name = "Investigations") %>% 
@@
   filter(.imp==1, age == 0,  state == "MN") %>% 
-  select(var, race_ethn, pop_max, q_c) %>% 
+  select(var, race_ethn, pop, q) %>% 
   mutate(name = "Confirmed maltreatment"))
@@
   mutate("Number of children" = round(var/5,0),
-         "Child population" = round(pop_max/5,0),
+         "Child population" = round(pop/5,0),
          "Event rate" = round(`Number of children`/ `Child population`,2)) %>% 
@@
 ncands_sub_inv_tab<-subst_tables %>%
-  rename(sub = q_c) %>% 
+  rename(sub = q) %>% 
   select(.imp , age, race_ethn, state, sub) %>% 
   left_join(investigation_tables %>% 
-              rename(inv = q_c) %>% 
+              rename(inv = q) %>% 
               select(.imp, age, race_ethn, state, inv)) %>% 
@@
 fc_cond_inv_tabs<-fc_post_inv_tables %>%
-  rename(fc_inv = q_c) %>% 
+  rename(fc_inv = q) %>% 
   select(.imp,age,  race_ethn, state, fc_inv) %>% 
   left_join(investigation_tables %>% 
-              rename(inv = q_c) %>% 
+              rename(inv = q) %>% 
               select(.imp, age, race_ethn, state, inv)) %>% 
@@
 fc_cond_sub_tabs<-fc_sub_tables %>%
-  rename(fc_sub = q_c) %>% 
+  rename(fc_sub = q) %>% 
   select(.imp,age,  race_ethn, state, fc_sub) %>% 
   left_join(subst_tables %>% 
-              rename(sub = q_c) %>% 
+              rename(sub = q) %>% 
               select(.imp, age, race_ethn, state, sub)) %>% 
@@
 non_icwa_cond_fc<-non_icwa_tables %>%
-  rename(non_icwa = q_c) %>% 
+  rename(non_icwa = q) %>% 
   select(.imp, age, race_ethn, state, non_icwa) %>% 
   left_join(fc_tables%>% 
-              rename(fc = q_c) %>% 
+              rename(fc = q) %>% 
               select(.imp, age, race_ethn, state, fc)) %>% 

### 2020-12-16 make_state_tables.R

@@ -439,9 +439,7 @@ icwa<-read_csv("./data/icwa_data.csv")
 
 ### add state abbrev
 xwalk<-data.frame(State = as.character(state.name), 
-                  state.abb = as.character(state.abb)) %>% 
-  mutate(AIAN = fc_aian,
-         White = fc_non)
+                  state.abb = as.character(state.abb)) 
 
 icwa_adopt<-icwa %>% 
   left_join(xwalk) %>% 
@@ -454,6 +452,8 @@ icwa_boarding<-icwa %>%
          White = NA)
 
 icwa_fc<-icwa %>% 
+  mutate(AIAN = fc_aian,
+         White = fc_non) %>% 
   left_join(xwalk) %>% 
   ungroup() %>% 
   mutate(AIAN = fc_aian,

### 2020-12-16 MS_descriptives.R
cleanup

@@ -1,419 +1,0 @@
-rm(list=ls()); gc()
-library(tidyverse)
-library(lubridate)
-library(gridExtra)
-library(RColorBrewer)
-theme_set(theme_bw())
-source("lifetable.r")
-pop<-read_fwf("./data/us.1990_2018.singleages.adjusted.txt",
-              fwf_widths(c(4, 2, 2, 3, 2, 1, 1, 1, 2, 8),
-                         c("year", "state", "st_fips",
-                           "cnty_fips", "reg", "race",
-                           "hisp", "sex", "age", "pop")),
-              col_types = "iccciiiiii")
-
-pop<-pop %>% 
-  filter(age<=18, race == 3 | (race==1 & hisp==0), year >=2003) %>% 
-  mutate(pop = as.integer(pop),
-         age = as.integer(age),
-         race_ethn = 
-           case_when(
-             race==1 & hisp ==0 ~ "White",
-             race==3 ~ "AI/AN"))%>%
-  group_by(year, age, race_ethn) %>% 
-  summarise(pop = sum(pop))
-
-format_data<-function(dat){
-  dat<-dat %>% 
-    filter(year>=2003, year<2017)%>% 
-    filter(age<=18) %>% 
-    filter(.imp!=0) %>% 
-    filter(race_ethn == "AI/AN" | race_ethn == "White") %>% 
-    select(-pop) %>% 
-    complete(.imp, age, race_ethn, year, fill = list(var = 0)) %>% 
-    left_join(pop)
-}
-
-ncands_inv<-read_csv("./data/ncands_first_inv.csv") %>% 
-  format_data()
-ncands_sub<-read_csv("./data/ncands_first_sub.csv")%>% 
-  format_data()
-
-
-afcars_fc<-read_csv("./data/afcars_first_fc.csv")%>% 
-  format_data()
-afcars_tpr<-read_csv("./data/afcars_tpr.csv")%>% 
-  format_data()
-afcars_non_icwa<-read_csv("./data/afcars_non_icwa.csv")%>% 
-  format_data()
-afcars_fc_inv<-read_csv("./data/afcars_fc_post_inv.csv")%>% 
-  format_data()
-afcars_fc_sub<-read_csv("./data/afcars_fc_post_sub.csv")%>% 
-  format_data()
-
-### read in new first investigation csv
-### run lifetable by year
-
-investigation_tables<-list()
[…]

### 2020-12-18 cfp_final_report.Rmd
update through lifetime risks

@@
 ### load packages and suppress warnings
-knitr::opts_chunk$set(echo = F, message = F, warning = F)
 library(tidyverse, warn.conflicts = FALSE)
-options(dplyr.summarise.inform = FALSE,
-        knitr.kable.NA = " ")
-
 library(bookdown)
@@
 
+# options
+knitr::opts_chunk$set(echo = F, message = F, warning = F)
+options(dplyr.summarise.inform = FALSE,
+        knitr.kable.NA = " ")
+options(scipen=999)
 ```
@@
 Since the early 2010s, all 50 states have submitted these report-level data annually to the US Children's Bureau.
+
 Here, we use the NCANDS data to measure two events for AIAN (defined as American Indian or Alaska Native alone or in combination with any other racial or ethnic identification) children and white children (defined as non-hispanic white alone).
 Using within-state unique child identifiers, we document the first time a child was the subject of a screened-in CPS report, and the first time a child was the subject of a confirmed or substantiated CPS report over a five-year period (2014 - 2018).
+Note that child identifiers for Georgia, Pennsylvania, and Rhode Island were not reliable over time for children with unsubstantiated screened-in cases.
+These states are exluded from analysis of risk of investigation, but included in all other analyses.
 
@@
 
-To obtain valid estimates of the AIAN child population by age for US states between 2014 and 2018, we rely on two data sources.
+To obtain valid estimates of the AIAN child population by age for US states between 2014 and 2018, we rely on the US Census Population and Housing Units Estimate Program (CITE STATE BY 5 RACE ALONE OR COMB FILE).
 Our definition of American Indian and Alaska Native includes any individual who identifies as American Indian or Alaska Native alone or in combination with any other group.
 The US Census Bureau's Population and Housing Unit Estimates program uses baseline data from the 2010 Census along with various sources of demographic data to estimate age and race specific population totals for all counties in the US annually.
-It
+At the state-level, this program provide single-age estimates of the population that identifies as American Indian / Alaska Native alone or in combination with any other group(s).
+These estimates are imputed by the US Census based on 2010 US Census baseline estimates and various demographic indicators for each locality to project population change.
 
-SEER POP DATA
+Unfortunately, we are not able to leverage Census population data directly to make comparisons of the incidence rate of family separation over time expressed as rates per capita.
+Changes in the enumeration of the AIAN population between 1976 and 2019 make such comparisons impractical.
+For example, the US Census enumerated approximately 1.4 million American Indians and Alaska Natives in the 1980 decennial count, a 70 percent increase over the 1970 decennial census.
+Such dramatic shifts are largely explained by the Census changing its methodology for determining race/ethnicity of a subject.
+Until 1970, the Census worker reported a subject's race/ethnicity based on worker observations.
+In 1980, the Census changed to self-identification of race/ethnicity (Lujan: American Indians and Alaska Natives Count).
+In 1990, the Census counted about 2 million American Indians and Alaska Native, an approximately 40 percent increase from the prior Census.
+By 2000, the Census reported 4.2 million American Indians and Alaska Natives living in the United States, more than double the 2000 population count.
+Such changes are not driven by demography alone (fertlity, mortality, migration), but instead are an artifact of shifting methods of enumerating the racial and ethnic identities of US residents by the Census Bureau.
+Shifting Census methods, enabling individuals to identify their own group membership, and to identify membership in multiple groups have dramatically changed official estimates of the size of the AIAN population.
+This makes direct comparisons of historical AIAN population counts to contemporary AIAN population counts problematic.
 
-ALONE OR IN COMBINATION PROBLEMS WITH BRIDGED ESTIMATES. CITE DEMOGRAPHY ON SHIFTS IN BASE POPULATION.
-
-PEP POP DATA
+For contemporary risk estimates, we rely on age-specific estimates of the population who identify as AIAN alone or in combination with any other group between 2014 and 2018 by state computed by the Census Population and Housing Units Estimation Program.
+We do not attempt here to make historical comparisons of incident rates of events between the contemporary period and the pre-ICWA period.
 
@@
 
-0. age distribution from SEER onto pep
-1. point-in-time comps
+## Historical comparisons
 
-AAIA compiled point-in-time estimates of the numbers of AIAN children 21 years of age and under in adoption or foster care in each state for which they were able to collect complete data through surveys of state agencies and private providers.
-We compare these numbers to counts of children in foster care at year-end and to total numbers of children in state-sponsored adoptions.
-Annual foster care counts are straightforward to compute with AFCARS, by restricting caseload counts to
+We directly compare counts of children experiencing family separation through adoption and foster care in the states surveyed by AAIA in 1976 to counts of children in foster care or adoption in 2019.
+Because historical and contemporary AIAN population estimates are incompatible, we do not provide estimates as per capita rates.
+AAIA reports point-in-time counts of children in foster care or adoption by state.
+For comparability, we compute point-in-time contemporary foster care caseloads by counting the number of children in foster care at the end of the reporting year in 2019, and adoption caseloads by using annual counts of new adoptions from 2010-2019 and estimating the number of children concurrently in adoptive homes under the age of 21 (for comparability with AAIA measures, which used 21 as the age threshold for inclusion).
+We compare counts of AIAN and non-AIAN children in each system for states included in the AIAN data at these two time points, then compare aggregate totals of children in some form of state-involved family separation across these two time periods.
 
-EXPLAIN VARIABLE YEAR OF DATA COLLECTION FOR AAIA
+## Age-specific and lifetime risks
 
-2. period lifetables (age-spec / lifetime)
-3. conditional probs
-4. imputation
+For each child welfare system outcome examined here (investigations, substantiations, foster care entries, terminations of parental rights), we use 2014 - 2018 NCANDS and AFCARS data to estimate period life tables for each state for AIAN children and white children separately.
+Period life tables share some of the analytic benefits of cohort life tables.
+In a cohort life table, analysts track the incidence of an outcome across a group over the life course, adjusting for mortality and prior-event incidence.
+When a cohort life table is impractical because of data limitations or shifts in risk over time, a period life table allows for a simulation of risk over the life course by assuming that age-specific risks remain static at the levels observed during the focal period.
 
-# Findings:Change since ICWA
+In this study, we use 5 years of child welfare and population data to estimate the risk of a first child welfare event for each group and age.
+We then apply these age-specific risk estimates to a hypothetical cohort of 100,000 infants for each group and state, counting the number of events that would have occurred in this cohort at 2014-2018 levels of age-specific risk and age-specific mortality (CITE THE DEMOGRAPHY BOOK).
+The key assumption of this method is that event and mortality risk levels remain relatively stable over the period.
+Caution is due in using these data to extrapolate to future periods, as both event risk and mortality risk may shift over time.
+
+## Case processing and conditional probabilities
+
+Period life tables provide us with rigorous estimates of the marginal probability that a child would experience a particular event in a focal population at a fixed level of risk.
+In this study, we are interested in identifying the institutional sites at which inequalities emerge or are exacerbated.
+To do so, we compute the conditional probability of successive steps in child welfare case processing -- the probability of substantiation after investigation; the probability of foster care after investigation; the probability of foster care after substantiation; and the probability of parental rights termination after foster care.
+
+To do so, we first estimates period life tables for the joint event of each combination of outcomes.
+For example, we link the NCANDS and AFCARS foster care data using unique child identifiers to estimate the joint probability that a child experience both a CPS investigation *and* a later foster care entry.
+Once we estimate this join probability, we can compute the conditional probability as follows:
+
+\[\textrm{Pr}(a|b) = \frac{\textrm{Pr}(a,b)}{\textrm{Pr}(b)}\]
+
+These conditional probabilities provide information on the likelihood that a child will move up the case processing ladder from one state to the next.
+By comparing the conditional probabilities of case processing outcomes separately for AIAN and white children across states, we can identify the states of child protection case processing during which inequalities emerge for AIAN children relative to white children.
+
+It is important to note that while state-level analyses provide far more insight into local processes than do national analyses, working at the state-level may obscure important differences in system practices and outcomes occurring at county or local levels.
+This is particularly relevant for states with county-administered child protection systems.
+Because of data-masking procedures in the NCANDS/AFCARS, analyses at the county-level are not consistently possible for non-urban locales.
+
+## Missing data and imputation
+
+A small proportion of cases in the NCANDS and AFCARS are missing data on focal variables including child race/ethnicity and caretaker race/ethnicity.
+To address this missing data and ensure that our results are not negatively biased through listwise deletion, we compute multiple imputations of the NCANDS and AFCARS at the report or child level using pseudo Bayesian methods as developed by Van Buuren (https://www.jstatsoft.org/article/view/v045i03).
+These models include child-level predictors such as age and sex in addition to including state-level population composition predictors to adjust for hetereogeneity in the racial composition of populations across states.
+All results reported in the figures below display uncertainty intervals that reflect the maximum and minimum estimates of each statistic, reflecting uncertainty driven by missing data.
+In most cases, this uncertainty is relatively small.
+
+# Findings
+
+## Change Since ICWA
 
@@
 
-`r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,1)`
-
 ```{r icwa-adopt-counts, fig.cap = "Children 21 and under in state-involved adoptions on reporting date by state and period. Data from AAIA surveys and AFCARS"}
@@
 
-\newpage
-
 ```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}
@@
 ```
-
-\newpage
 
@@
 Foster care caseloads grew in many states for both non-AIAN and AIAN children.
-Of the 19 states included here, 12 saw higher growth in AIAN foster care caseloads than they saw in non-AIAN fooster care caseloads.
+Of the 19 states included here, 12 saw higher growth in AIAN foster care caseloads than they saw in non-AIAN foster care caseloads.
 Adoption caseloads also grew faster for AIAN children in these states than they did for non-AIAN children.
@@
 
-\newpage
+## Contemporary risks of child welfare system contact
 
-# Findings:Contemporary risks of child welfare system contact
+We now explore in detail the contours of the contemporary exposure of AIAN children and families to various forms of child welfare system contact.
+We first present age-specific risks of successive child welfare events.
+Next, we provide cumulative (life time) risks of child welfare event incidence.
+Finally, we estimate and visualize conditional probabilities of experiencing successive child welfare events, illustrating differences in case processing for Native and non-Native children across states.
+All analyses below rely on pooled state-level 2014 - 2018 data on child welfare system events from the NCANDS child file and AFCARS foster care file, along with pooled state-level data on child populations from the Census Population and Housing units Estimation Program.
 
-INTRO WITH MOTIVATION - INCREASING SERIOUSNESS OF CONTACT, POSSIBILITY OF ESTIMATING POINTS OF INEQUALITY ACCELERATION
+All estimates should be interpreted as risks at 2014 - 2018 levels. Caution should be exercised in using these estimates to project risk to future periods.
 
-## Age-specific risks
+### Age-specific risks
 
-Figure \@ref(nat-marginal) shows the marginal probabilities of experiencing a series of child welfare system events for both Native and non-Native children.
+Figure \@ref(fig:nat-marginal) shows the marginal probabilities of experiencing a series of child welfare system events for both Native and non-Native children.
 This plot shows national age-specific event probabilities at 2014-2018 risk levels for AIAN and non-AIAN children for experiencing 1) a screened-in investigation, 2) a confirmed maltreatment case, 3) a foster care placement, and 4) termination of parental rights through the child welfare system.
-DISCUSSION OF THIS.
 
@@
 
+Across all outcomes, and for both Native and white children, infants are at highest risk of child welfare system contact.
+About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Investigation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of Native infants experience a CPS investigation nationally, and about `r plot_dat %>% filter(age==0, race_ethn=="White", var=="Investigation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of white infants experience a CPS investigation.
+About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Substantiation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of AIAN children experience an agency confirmed maltreatment case before their first birthday, compared with about `r plot_dat %>% filter(age==0, race_ethn=="White", var=="Substantiation") %>% select(q) %>% mutate(q = round(q*100, 1))`.
+About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Foster Care") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of AIAN children are removed from their families and placed into foster care before their first birthday, and about `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Termination") %>% select(q) %>% mutate(q = round(q*100, 1))`.
+Risk of a first event declines steadily with age for both groups.
+
+Nationally at 2014 - 2018 risk levels, white children are at higher risk of investigation at all ages, risks of substantiation are similar for white and Native children, risks of foster care are higher for native children, and risks for termination are slightly higher for native children.
+
 Of course, there is substantial variation in risk of child welfare system contact across US states.
-EXPLAIN WHY THIS IS.
+States have tremendous flexibility in designing and implementing both child welfare surveillance systems and foster care systems.
+The character of government-to-government relationships between tribes and states varies tremendously across places.
+Federal law also differentially affects tribal governance and the relationships between states, counties and tribes across places (CITES).
 Figure\@ref(fig:nat-inv-age) shows the age-specific risk of experiencing a first screened-in investigation by US state for both AIAN and white children at 2014 - 2018 levels of risk.
-DESCRIBE FINDINGS
+
+```{r}
+### for state infant comparisons
+temp<-investigation_tables %>% 
+         group_by(state, race_ethn, age) %>% 
+         summarise(qmax = max(q), qmin=min(q), q = mean(q)) %>% 
+  filter(age==0) %>% 
+  arrange(desc(q)) %>% 
+  ungroup()
+```
+
+Infants are at highest risk of investigation in all states, and the shape of the age-risk curve is similar to the national pattern shown in Figure \@ref(fig:nat-marginal).
+Risk tends to be markedly higher for infants, and decline slowly with age.
+However, there is tremendous variation in levels of risk across states, and clear regional patterning in inequalities in the exposure of infants to CPS investigation.
+AIAN infant risks of investigation are highest in Alaska.
+In 2014 - 2018, about `r temp %>% filter(state=="AK", race_ethn=="AIAN") %>% select(q) %>% mutate(q = round(q * 100, 1))` percent of AIAN infants had a screened-in CPS report.
+More than 10 percent of AIAN infants experienced an investigation in 3 other states: Minnesota, Oklahoma, and Montana.
+White children in Southern states like Mississippi, West Virginia, Florida, and Kentucky are typically at  higher levels of risk of investigation at all ages that are AIAN children, whose risk levels are typically quite low.
+In The Great Plains and Upper Midwest states, AIAN children typically have much higher investigation risk than do white children.
+Inequalities are pronounced in Wisconsin, Minnesota, North and South Dakota, Nebraska, and Oklahoma.
+Other than Alaska, states in the West typically have similar investigation rates for AIAN and white children.
 
@@
 
-There are also dramatic differences in risk of experiencing foster care across states.
-Figure \@ref(fig:nat-fc-age) shows the distribution of age-specific risks of first foster care entries for Native and white children by US state at 2014 - 2018 levels of risk.
-DESCRIBE FINDINGS
+```{r}
+### for state infant comparisons
+temp<-fc_tables %>% 
+         group_by(state, race_ethn, age) %>% 
+         summarise(qmax = max(q), qmin=min(q), q = mean(q)) %>% 
+  filter(age==0, race_ethn=="AIAN") %>% 
+  arrange(desc(q)) %>% 
+  ungroup() %>% 
+  mutate(q = round(q* 100,1))
+```
+
+In Minnesota, at 2014 - 2018 levels of risk, 9.7 percent of AIAN children could be expected to be removed from their families and placed into foster care before their first birthday.
+This is the highest level of risk in the country for infants of any racial or ethnic group (CITE YI EDWARDS WILDEMAN WORKING PAPER).
+AIAN infants are also at exceptionally high risk of foster care placement in Montana, Alaska, South Dakota, North Dakota, Oklahoma, Washington, Oregon, Nebraska, and Wisconsin.
+At 2014 - 2018 risk levels, more than 3 percent of all AIAN infants would be placed into foster care.
+White children face a 3 percent risk of foster care before their first birthday in only one state: West Virginia.
+Unlike with investigation, where white children in some Southern states experienced levels of extreme risk similar to risk levels experienced by AIAN children in Plains and Midwestern states, Native risk of foster care in high-risk states is categorically different from white risk in any state.
+
+Nationally, white infants experienced about a 1.1 percent risk of foster care entry before their first birthday.
+Teenagers are typically at relatively low-risk of entering foster care for the first time, but in Minnesota, AIAN 16 year olds face a 1.4 percent risk of a first foster care entry, higher than the national risk faced by white infants.
+Risk for 16 year old AIAN children is also higher than white infant risk in North Dakota.
 
@@
 
-THEN HIGH RISK STATES FOR EACH OUTCOME FOR AGESPEC
+In states where Native children are at high risk of investigation, they are typically also at higher risk of substantiation.
+Figure \@ref(fig:sub-high) shows the age-specific risk of a first substantiated maltreatment investigation for the states with the six highest levels of risk for AIAN infants: Alaska, Iowa, Massachusetts, Minnesota, Montana, and Oklahoma.
+In Alaska, Minnesota, and Oklahoma, about 6 percent of AIAN infants experience an investigation at 2014 - 2018 risk levels.
+About 5 percent of AIAN children experience a substantiated case before their first birthday in Massachusetts, Iowa, and Montana.
+AIAN children are at consistently higher risk of substantiation over the life course in each of these states except Massachusetts, where white children's risk is similar to AIAN children's risk after infancy.
 
-ADD PLOTS HERE
+```{r sub-high, fig.cap = "Age-specific probability of agency-confirmed maltreatment case in high-risk states, 2014 - 2018 risk levels"}
+sub_high <- subst_tables %>% 
+  filter(age==0, .imp==1) %>% 
+  filter(race_ethn=="AIAN") %>% 
+  arrange(desc(q)) %>% 
+  ungroup() %>% 
+  slice(1:6)
 
-## Lifetime event risks and inequalities in lifetime risk
+ggplot(subst_tables %>% 
+         filter(state %in% sub_high$state) %>% 
+         group_by(state, race_ethn, age) %>% 
+         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
+       aes(x = age, y = q, ymin = qmin, ymax = qmax,
+           color = race_ethn, fill = race_ethn)) + 
+  geom_ribbon(alpha = 0.5, color = NA) + 
+  geom_line() + 
+  labs(color = "", fill = "") + 
+  facet_wrap(~state)+
+  labs(x = "Age", y = "Proportion of children") + 
+  # theme(strip.text.x = element_text(size = 7),
+  #       axis.text.x = element_text(size = 6),
+  #       axis.text.y = element_text(size = 6)) + 
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+```
 
-Figure \@ref(fig:cumul-map) displays the AIAN childrens' risk of ever experiencing each included child welfare system event by their age 18 at 2014 - 2018 levels of risk. DESCRIBE FINDINGS HERE
+AIAN children are at higher risk nationally of experiencing the full termination of parental rights while in foster care than are children from any other racial or ethnic group (CITE WILDEMAN EDWARDS WAKEFIELD).
+Like other outcomes, risk is at its peak for infants.
+Figure \@ref(fig:tpr-high) shows the age-specific probability of termination for Native and white children in the six states with the highest risk of termination for native infants.
+Risk for infants is highest in Minnesota.
+About 1.8 percent of AIAN children would experience full termination of parental rights before their first birthday at 2014 - 2018 levels of risk.
+Risk for infants is also at or above one percent in Alaska, Iowa, Montana, North Dakota, and Oklahoma.
+
+```{r tpr-high, fig.cap = "Age-specific probability of full termination of parental rights, 2014 - 2018 risk levels"}
+tpr_high <- tpr_tables %>% 
+  filter(age==0, .imp==1) %>% 
+  filter(race_ethn=="AIAN") %>% 
+  arrange(desc(q)) %>% 
+  ungroup() %>% 
+  slice(1:6)
+
+ggplot(tpr_tables %>% 
+         filter(state %in% tpr_high$state) %>% 
+         group_by(state, race_ethn, age) %>% 
+         summarise(qmax = max(q), qmin=min(q), q = mean(q)),
+       aes(x = age, y = q, ymin = qmin, ymax = qmax,
+           color = race_ethn, fill = race_ethn)) + 
+  geom_ribbon(alpha = 0.5, color = NA) + 
+  geom_line() + 
+  labs(color = "", fill = "") + 
+  facet_wrap(~state)+
+  labs(x = "Age", y = "Proportion of children") + 
+  # theme(strip.text.x = element_text(size = 7),
+  #       axis.text.x = element_text(size = 6),
+  #       axis.text.y = element_text(size = 6)) + 
+  scale_fill_brewer(palette = "Dark2")+
+  scale_color_brewer(palette = "Dark2")
+```
+
+While there is a very clear relationship between the risk of experiencing child welfare system events and age -- infants are at much greater risk than other children, then risk tpyically declines slowly with age -- there is tremendous difference in the frequency of these events across places, and clear regional patterns in differences in risk between Native and white children.
+
+### Lifetime event risks and inequalities in lifetime risk
+
+Next, we provide a summary of lifetime incidence of each child welfare event for Native and white children across US states.
+These cumulative risks simulate the total exposure of the population of children to a first event by applying age-specific risks and mortality risks (computed from the observed data) to a hypothetical cohort of 100,000 children in each state.
+They can be interpreted as 'lifetime' risks of child welfare system event incidence for each group.
+
+Figure \@ref(fig:cumul-map) displays AIAN risk of ever experiencing each child welfare system event by age 18 at 2014 - 2018 levels of risk. **START HERE**
 
@@
 
-\newpage
-
-# Findings: conditional probability of subsequent child welfare events
+### Conditional probability of subsequent child welfare events
 
@@
 
-\newpage
-
 # Discussion

### 2020-12-21 cfp_final_report.Rmd
update with conditional text

@@
 These cumulative risks simulate the total exposure of the population of children to a first event by applying age-specific risks and mortality risks (computed from the observed data) to a hypothetical cohort of 100,000 children in each state.
-They can be interpreted as 'lifetime' risks of child welfare system event incidence for each group.
-
-Figure \@ref(fig:cumul-map) displays AIAN risk of ever experiencing each child welfare system event by age 18 at 2014 - 2018 levels of risk. **START HERE**
+They can be interpreted as lifetime risks of child welfare system event incidence for each group.
+
+Figure \@ref(fig:cumul-map) displays AIAN risk of ever experiencing each child welfare system event by age 18 at 2014 - 2018 levels of risk.
+Note the variable scale for each outcome, and note that Georgia, Pennsylvania, and Rhode Island lack reliable identifiers with unsubstantiated screened-in investigations for the period.
+More than half of all American Indian and Alaska Native children living in Alaska, Montana, Oklahoma, Nebraska, Montana, and North Dakota would experience a CPS investigation by age 18 at 2014 - 2018 levels of risk.
+Risk is highest in Alaska, where nearly 80 percent of AIAN children receive an investigation during their childhood.
+There are 7 states where white children have risks at or above 50 percent; West Virginia, Arkansas, Indiana, Kentucky, Florida, Tennessee, and Michigan.
+
+More than one in five Native children can expect to have a substantiated or confirmed CPS case in Alaska, Minnesota, Oklahoma, North Dakota, and Montana at 2014 - 2018 levels of risk.
+AIAN children in Alaska are at greatest risk, about 3 in 10 children have a confirmed case by age 18.
+There are only two states where white children face a one in five risk of a substantiated case: Kentucky and West Virginia.
+
+We estimate that about 1 in 3 Native children in Minnesota will ever enter foster care at levels of risk similar to those observed between 2014 - 2018.
+In Montana, this risk is about 1 in 4.
+In North Dakota, Alaska, and South Dakota, the risk is about 1 in 5.
+There is nowhere in the country where white children face levels of risk as high as 1 in 5.
+AIAN lifetime risk of foster care is also high (greater than 1 in 10) in Nebraska, Wisconsin, Iowa, Oklahoma, Oregon, and Washington.
+
+Termination of parental rights through the foster care system occurs for about 4 percent of all AIAN children in Minnesota.
+About 3 percent of AIAN children in Oklahoma, Montana, North Dakota, Iowa, and Alaska can expect to experience the full termination of parental rights by age 18 at 2014 - 2018 levels of risk.
+White children in West Virginia and Arizona also face risks of termination of parental rights of around 3 percent.
+
+Child welfare system lifetime event risks are highest for AIAN children in the same Great Plains, Northwestern, and Midwestern states that showed high levels of event risk for infants.
+In many states, more than half of all AIAN children ever have contact with the child welfare system, and foster care is disturbingly common for Native children in Minnesota, Montana, North Dakota, Alaska, and South Dakota, where more than one in five Native children can expect to ever enter foster care at current levels of risk.
 
@@
 
-Figure \@ref(fig:ineq-map) shows the ratio of AIAN lifetime event risk to white lifetime event risk for each state. DISCUSS FINDINGS HERE.
+Figure \@ref(fig:ineq-map) shows the ratio of AIAN lifetime event risk to white lifetime event risk for each state.
+These rate ratios provide a measure of the magnitude of inequality in AIAN system contact compared to white system contact.
+The same states that have high levels of event risk for AIAN children also have high levels of inequality in the rate at which AIAN children experience system contact compared to white children.
+Inequalities in investigations are less pronounced than inequalities in other outcomes.
+South Dakota, Minnesota, and Alaska exhibit the highest levels of inequality in CPS investigations.
+AIAN children are about 3 times more likely than white children to be investigated in these states.
+In these same states (South Dakota, Minnesota, and Alaska), AIAN children are at least 4 times more likely than white children to ever experience an agency-confirmed maltreatment case.
+
+Foster care inequalities are particularly pronounced.
+In Minnesota, AIAN children are about 8 times more likely than white children to ever enter foster care.
+In South Dakota, AIAN children are about 7 times more likely than white children to ever enter foster care.
+Native children in North Dakota, Wisconsin, and Alaska are about 4 times more likely than white children to ever enter foster care.
+Native children are more likely than white children to enter foster care in 20 states.
+Inequalities in termination of parental rights have a similar geography.
+In South Dakota, North Dakota, and Minnesota, AIAN children are at least 5 times more likely than white children to experience full parental rights termination through the child welfare system.
+In Massachusetts, Montana, Alaska, Iowa, Wisconsin, Nebraska, and Washington, AIAN children are about twice as likely to experience termination of parental rights as are their white peers.
 
@@
 
-SUMMARY GRAF ON MARGINAL PROBABILITIES (AGE-SPEC AND CUMULATIVE)
-
-### Conditional probability of subsequent child welfare events
-
-Table \@ref(tab:mn-demo) show some cool things. EXPLAIN WHAT THESE MEASURE AND HOW THEY ARE COMPUTING
+## Case processing and conditional risks
+
+The marginal probabilities of event incidence discussed above provide compelling pictures of the magnitude of the exposure of AIAN children to the child welfare system across the U.S.
+in the post-ICWA period.
+However, they do not answer core questions about the sources of such routine contact and disruptive intervention.
+Below, we explore how AIAN children are processed through state child welfare systems.
+We identify the risk of experiencing a child welfare system event, conditional on experiencing a prior case-processing event.
+These conditional probabilities allow us to examine how inequalities emerge or are exacerbated as AIAN children are processed through child welfare systems.
+
+Table \@ref(tab:mn-demo) demonstrates the basic logic of this approach.
+We display the average numbers of AIAN children subject to either an investigation or a substantiated investigation.
+The crude event rate is simply the number of children experiencing each event divided by the size of the population at-risk of experiencing the event.
+To compute the rate at which children transition from one case processing event to the next stage of case processing, we first compute the joint probability of experiencing both events, then divide by the marginal probability of experiencing the first event.
+In this case, because all children with confirmed maltreatment cases also must have had an investigation, we treat the confirmed maltreatment rate as the joint probability $\textrm{Pr(investigation, substantiation)} = \textrm{Pr(substantiation)}$ and simply divide this quantity by the probability of investigation.
+We learn that about 41 percent of all AIAN children infants in Minnesota with a screened-in case can expect to be substantiated at 2014 - 2018 rates.
+
+Because we cannot assume that all children entering foster care experienced either a) an investigation or b) a substantiation, we match unique identifiers across AFCARS and NCANDS for first events of each type, then separately estimate life tables for the joint probability of the focal event.
+
+To identify sites of institutional decision making where inequalities in AIAN child welfare system case processing emerge or are accelerated, we compute conditional event probabilities for 1) case substantiation, conditional on investigation; 2) foster care placement, conditional on investigation; 3) foster care placement, conditional on substantiation; and 4) termination of parental rights after foster care placement for both AIAN and white children.
+Figure \@ref(fig:nat-cond) displays the national risk levels for each event.
 
@@
 
-To identify sites of institutional decision making where inequalities in AIAN child welfare system case processing emerge or are accelerated, we compute conditional event probabilities for 1) case substantiation, conditional on investigation; 2) foster care placement, conditional on investigation; 3) foster care placement, conditional on substantiation; and 4) termination of parental rights after foster care placement for both AIAN and white children.
-Figure \@ref(fig:nat-cond) displays the national risk levels for each event.
-
-DISCUSS FINDINGS
+Nationally, at each stage of case processing other than the termination of parental rights, AIAN children are more likely than their white peers to have their cases move up to the next, more serious stage of case processing.
+About 40 percent of AIAN infants who are the subject of an investigation experience a substantiated case, compared to about 33 percent of white infants.
+About 28 percent of AIAN infants who experience a screened-in investigation enter foster care, compared to about 15 percent of white infants.
+About 62 percent of AIAN children who experience an agency-confirmed maltreatment case enter foster care, compared to about 41 percent of white children.
+About 27 percent of AIAN infants who enter foster care experience full termination of parental rights, compared to about 31 percent of white children.
+For each outcome, these gaps are relatively stable as children age.
 
@@
 
-STATE LEVEL FINDINGS
-
-```{r sub-inv}
+We display the distribution of age-specific probabilities of substantiation, conditional on investigation across US states in Figure \@ref(fig:sub-inv).
+Note that states with relatively small AIAN populations, such as Vermont and Delaware, have relatively unstable estimate for these conditional probabilities, as small changes in event counts can drive large fluctuation in relative rate estimates when population denominators are relatively small.
+The probability of substantiation, conditional on investigation is highest for infants.
+White children in Louisiana experience the highest conditional probability of substantiation, at about 65 percent.
+Similar levels of risk occur for AIAN infants in Idaho, Utah, Massachusetts, and Iowa.
+
+```{r sub-inv, fig.cap = "Probability of substantiation, conditional on investigation, 2014 - 2018 levels of risk"}
 ncands_sub_inv_tab<-subst_tables %>%
@@
   scale_fill_brewer(palette = "Dark2")+
-  scale_color_brewer(palette = "Dark2")
-```
-
-```{r fc-inv}
+  scale_color_brewer(palette = "Dark2")+ 
+  theme(axis.text.y = element_text(size = 5),
+        axis.text.x = element_text(size = 5))
+```
+
+Figure \@ref(fig:fc-inv) shows the age specific probability of entering foster care, conditional on experiencing a CPS investigation.
+American Indian and Alaska Native infants in California, Minnesota, Hawaii, Idaho, Oregon, and Montana all have a greater than 40 percent chance of entering foster care if they experience a child welfare investigation, at 2014 - 2018 risk levels.
+
+Figure \@ref(fig:fc-sub) shows the age specific probability of entering foster care, conditional on experiencing an agency confirmed maltreatment investigation.
+In Oregon, California, Missouri, Montana, New Hampshire, and Washington, more than 80 percent of all AIAN infants with substantiated cases will be placed into foster care at 2014 - 2018 levels of risk.
+White infants do not experience a conditional risk of foster care this high in any state.
+
+```{r fc-inv, fig.cap = "Probability of foster care, conditional on investigation, 2014 - 2018 levels of risk"}
 fc_cond_inv_tabs<-fc_post_inv_tables %>%
@@
               select(.imp, age, race_ethn, state, inv)) %>% 
-  mutate(fc_cond_inv = fc_inv / inv)
+  mutate(fc_cond_inv = fc_inv / inv) %>% 
+  filter(!is.infinite(fc_cond_inv))
 
@@
   scale_fill_brewer(palette = "Dark2")+
-  scale_color_brewer(palette = "Dark2")
-```
-
-```{r fc-sub}
-
+  scale_color_brewer(palette = "Dark2")+ 
+  theme(axis.text.y = element_text(size = 5),
+        axis.text.x = element_text(size = 5))
+```
+
+```{r fc-sub, fig.cap = "Probability of foster care, conditional on substantiation, 2014 - 2018 levels of risk"}
 fc_cond_sub_tabs<-fc_sub_tables %>%
@@
               select(.imp, age, race_ethn, state, sub)) %>% 
-  mutate(fc_cond_sub = fc_sub / sub)
+  mutate(fc_cond_sub = fc_sub / sub)%>% 
+  filter(!is.infinite(fc_cond_sub)) %>% 
+  filter(fc_cond_sub<1)
 
@@
   scale_fill_brewer(palette = "Dark2")+
-  scale_color_brewer(palette = "Dark2")
-```
-
-```{r non-icwa}
+  scale_color_brewer(palette = "Dark2")+ 
+  theme(axis.text.y = element_text(size = 5),
+        axis.text.x = element_text(size = 5)) + 
+  coord_cartesian(ylim = c(0,1))
+```
+
+```{r non-icwa, fig.cap = "Probability AIAN children in foster care ever placed in non-kin, non-native foster setting, 2014 - 2018 levels of risk"}
 non_icwa_cond_fc<-non_icwa_tables %>%
@@
   facet_geo(~state) + 
-  labs(x = "Age", y = "Proportion") 
+  labs(x = "Age", y = "Proportion") + 
+  theme(axis.text.y = element_text(size = 5),
+        axis.text.x = element_text(size = 5))
 ```

### 2020-12-21 cfp_final_report.Rmd
draft push

@@
 aim of report is to 1) evaluate magnitude of change since ICWA, 2) ID geographic variation in inequality, 3) ID sites and places where inequality is accelerated
-
-# lit review
-
-1. historical importance of family separation
-2. ICWA and policy
-3. prior empirics on AIAN inequality
 
@@
 American Indian and Alaska Native infants in California, Minnesota, Hawaii, Idaho, Oregon, and Montana all have a greater than 40 percent chance of entering foster care if they experience a child welfare investigation, at 2014 - 2018 risk levels.
-
 Figure \@ref(fig:fc-sub) shows the age specific probability of entering foster care, conditional on experiencing an agency confirmed maltreatment investigation.
 In Oregon, California, Missouri, Montana, New Hampshire, and Washington, more than 80 percent of all AIAN infants with substantiated cases will be placed into foster care at 2014 - 2018 levels of risk.
-White infants do not experience a conditional risk of foster care this high in any state.
+
+When interpreting Figures \@ref(fig:sub-inv) \@ref(fig:fc-inv) \@ref(fig:fc-sub) jointly, along with Figure \@ref(fig:nat-inv-age), a clear pattern emerges.
+Much of the inequality experienced by AIAN children emerges once a child welfare system begins processing a case.
+With certain exceptions (Alaska, in particular), inequalities typically emerge post-investigation.
+Once AIAN children come into contact with child welfare systems, they are often at much higher risk of substantiation that their peers, and much more likely to be placed into foster care than their peers in both substantiated and unsubstantiated cases.
+The unequal treatment of AIAN children in the child welfare system is dramatically accelerated at the at the point of the decision to place a child into foster care.
 
@@
 
+Once a decision has been made to place a Native child into foster care, the Indian Child Welfare Act specifies a preference for placements with kin, with members of the child's tribe, or with another native family.
+Non-native and non-kin foster families are intended to be a last resort, when preferred placement settings are unavailable.
+Figure \@ref(fig:non-icwa) shows the proportion of AIAN children in foster care who were ever placed into a non-kin non-native foster setting during their time in foster care.
+Infants have the highest likelihood of being placed with kin or with a native family.
+However, if 41 US States, we estimate that more than 50 percent of AIAN infants are placed in a non-native non-kin foster setting.
+Maryland, New Hampshire, Connecticut, Pennsylvania, North Carolina, and Florida have the highest rate of ICWA preferred placements for infants, ranging between 54 and 61 percent of infants placed with either kin or an AIAN foster family.
+In Georgia, South Carolina, and Kentucky, more than 80 percent of AIAN infants were placed with a non-kin non-Native foster caretaker.
+States with the greatest frequency of AIAN foster care also routinely place AIAN children with non-kin non-AIAN caretakers.
+In South Dakota, 70 percent of infants are placed in non-kin non-Native settings, In Washington and North Dakota, 65 percent of AIAN infants are placed in non-kin non-AIAN settings.
+In Oklahoma, Minnesota, Wisconsin, and Montana, more than half of all AIAN infants placed in foster care are placed with non-kin non-AIAN caretakers.
+The likelihood that an AIAN child weill be placed with a family member or other Native foster caretaker typically decreases with age.
+There is generally a very small chance that an AIAN teenager placed into foster care will be place with kin or with an AIAN foster family.
+
 ```{r non-icwa, fig.cap = "Probability AIAN children in foster care ever placed in non-kin, non-native foster setting, 2014 - 2018 levels of risk"}

### 2020-12-22 cfp_final_report.Rmd
add discussion

@@
 We do not attempt here to make historical comparisons of incident rates of events between the contemporary period and the pre-ICWA period.
-
-5. Population data
 
@@
 In Oklahoma, Minnesota, Wisconsin, and Montana, more than half of all AIAN infants placed in foster care are placed with non-kin non-AIAN caretakers.
-The likelihood that an AIAN child weill be placed with a family member or other Native foster caretaker typically decreases with age.
+The likelihood that an AIAN child will be placed with a family member or other Native foster caretaker typically decreases with age.
 There is generally a very small chance that an AIAN teenager placed into foster care will be place with kin or with an AIAN foster family.
@@
 
+This report has provided two sets of empirical findings.
+
+First, we examine data on the breadth of the separation of AIAN children prior to the passage of the Indian Child Welfare Act, and compare these data to contemporary AIAN family separation through the child welfare system.
+We find that, while the overall scale of AIAN family separation has decreased since ICWA due to the closing of BIA residential boarding schools, the scale of AIAN family separation through fostering and adoption has increased in the states for which we have comparable data.
+In some states, like Oklahoma, Washington, California, and Alaska, the frequency of AIAN foster care and adoption has increased dramatically since the mid-1970s.
+
+These results should be interpreted with several limitations in mind, however.
+Because of changes in the ways that the US Census tabulates AIAN populations, and because of cultural shifts in how AIAN people identify themselves when asked questions about racial identity on surveys like the Census (CITES), estimates of the size of the AIAN child population are not comparable between 1976 and 2014 - 2018.
+As such, we cannot directly compare population adjusted incidence rates, only raw counts.
+Nonetheless, the magnitude of growth in the frequency of AIAN foster care and adoption in many states has outpaced the kinds of growth we would expect from demographic processes alone, and suggests that the scope of AIAN fostering and adoption has indeed grown over this period.
+Future research should explore methods for normalizing population counts across these two periods.
+
+We should also consider AAIA's counts of Native children in foster care and adoption in the pre-ICWA period to be conservative.
+AAIA relied on agency surveys to count AIAN children that had been separated from their families, and limited survey non-response negatively biases these counts.
+Additionally, many state and local agencies did not collect data on a child's racial or ethnic identity during this period.
+In some states, AAIA imputed the total number of AIAN children in foster care and adoption by applying the state or local population composition to the total count of children in foster care or adoption.
+Assuming any disparity in placement (a reasonable assumption), these imputations would be negatively biased, as AAIA suggests in their report.
+As such, the pre-ICWA placement numbers are likely conservative estimates of the scale of AIAN family separation in 1976.
+Because there is likely no such bias in contemporary numbers, estimates of proportional change (as displayed in Figure \@ref(fig:icwa-pct)) may overstate the magnitude of change since ICWA.
+
+Second, we examine data on the contemporary contact of AIAN children with the child welfare system at 5 distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights.
+We find that AIAN children are more likely than white children to be placed into foster care, even when rates of investigation and substantiation of cases are similar to those for white children.
+We find that inequalities in case outcomes vary dramatically across regions of the country.
+Inequalities for AIAN children are typically low in the South and Northeast, and typically very high in the Midwest, Great Plains, Mountain West, and Northwest.
+While some states, such as Alaska, Minnesota, and North and South Dakota exhibit high inequalities in risks of investigation for AIAN children, we find that inequalities in family separation are typically a function of later case processing decisions.
+In many places, decisions made once a child is in the system, are critical in exacerbating the unequal exposure of AIAN children to family separation.
+
+Efforts to reduce AIAN family separation should be mindful of where inequalities emerge for Native children and families.
+In some places, the front door of reporting and investigations accounts for a great deal of inequality.
+In others, AIAN children are treated categorically differently than white children when it comes time to make a decision about removal into foster care after an investigation or substantiation.
+Our data also show that all states are not adequately addressing ICWA placement preferences.
+In nearly all states, AIAN children can expect to be placed in a non-kin non-Native setting at some point during their time in foster care.
+
 # Conclusion

### 2020-12-31 cfp_final_report.Rmd
draft complete

@@
 author: "Frank Edwards and Theresa Rocha Beardall"
-date: "12/2020"
+date: "December, 2020"
+bibliography: "casey_rpt.bib"
+link-citations: true
+csl: "vancouver-brackets-no-et-al.csl"
 abstract: "American Indian and Alaska Native children are separated from their families by state child welfare agencies at exceptionally high rates.
-This study connects contemporary trends in Native family separation to histories of Indian child removal, and locates places and institutional sites where inequality emerges."
+This study connects contemporary trends in Native family separation to histories of Indian child removal, and provides insights into the geographies and institutional sites where inequalities emerge.
+We find that the total number of AIAN children in foster care or adoptive homes in states with large Native populations has increased since the passage of the Indian Child Welfare Act (1978).
+We find that risks of child welfare system contact are highest for AIAN infants, and that risk is highly variable across states.
+Our estimates show that in high risk states at 2014 - 2018 levels of risk, more than half of AIAN children will ever be investigated by a child welfare agency, more than one in five will experience a substantiated maltreatment case, and more than one in five will ever enter foster care.
+We further find that child welfare agency case processing exacerbates inequality.
+AIAN children are more likely than white children to enter foster care, conditional on experiencing a substantiated maltreatment case, than are white children.
+These exceptionally high levels of risk indicate that the crisis of Indian family separation is ongoing.
+For AIAN children in states like Minnesota, Alaska, South Dakota, North Dakota, and Oklahoma, contact with the child welfare system is a routine part of growing up."
 output: bookdown::pdf_document2
@@
 
-clear statement on ongoing crisis of native family separation.
-aim of report is to 1) evaluate magnitude of change since ICWA, 2) ID geographic variation in inequality, 3) ID sites and places where inequality is accelerated
+Despite growing attention to threats against the health, safety, and wellbeing of tribal communities in recent years, the child welfare system remains a confounding and persistently complex source of harm in the lives of Native families throughout the U.S.
+These harms stem from institutional acts of violence waged against Native Peoples by the U.S.
+federal government.
+This began with early Americans who were eager to physically remove Native Peoples from their homelands in order to establish commerce, statehood, and homes for non-Native families.
+Often, the intention of such violence was to break up Native families and assimilate them into the nation’s population.
+Boarding schools and adoption programs carried the U.S.
+mission of assimilating Native children well into the 20th century.
+
+Family separation continues to jeopardize and threaten the autonomy and wellbeing of Native families, and as a result, the very existence of tribal communities.
+These threats remain despite the passage of the landmark Indian Child Welfare Act (ICWA) of 1978, a law that was implemented to address and end generations of abuse, mistreatment, and forcible removal of Native children from their homes.
+
+In the following study, we describe the scale of the ongoing crisis of family separation by 1) evaluating the magnitude of change of family separation since ICWA, 2) identifying geographic variation in child welfare inequality, and 3) identifying the stages in case processing where inequality is accelerated.
+This study provides new empirical evidence about where and how Native families are being pushed into the child welfare system, and situates these findings in the long historical legacy of Native family separation.
 
@@
 This study relies on a number of data sources.
-First, we compile and digitize statistics collected by the Association for American Indian Affairs during the late 1960s and early to mid 1970s in an effort to document the breadth and depth of the Indian family separation crisis.
-To capture the contemporary exposure of AIAN children to the child welfare system, we rely on federal data systems collected by the US Children's Bureau.
-We use the 2014 - 2018 National Child Abuse and Neglect Data System (NCANDS) child files (CITE), the 2014 - 2018 Adoption and Foster Care Analysis and Reporting System (AFCARS) foster care files (CITE) and the 2010 - 2018 AFCARS adoption files (CITE).
-We use two data sources for contemporary population estimates: the US Census Bureau's Population Estimates Program (PEP) (CITE), and the National Institute of Healths' SEER population estimates (CITE).
+First, we compile and digitize statistics collected by the Association for American Indian Affairs during the late 1960s and early to mid 1970s in an effort to document the breadth and depth of the Indian family separation crisis @ninety-fifth_congress_indian_1977.
+To capture the contemporary exposure of AIAN children to the child welfare system, we rely on federal data systems collected by the U.S.
+Children's Bureau.
+We use the 2014 - 2018 National Child Abuse and Neglect Data System (NCANDS) child files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_national_2018, the 2014 - 2018 Adoption and Foster Care Analysis and Reporting System (AFCARS) foster care files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_adoption_2018 and the 2010 - 2018 AFCARS adoption files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_adoption_2018-1.
+We rely on the U.S.
+Census Bureau's Population Estimates Program (PEP) for contemporary population estimates @us_census_bureau_age_2020.
 
@@
 When this was the case, AAIA imputed foster care caseload totals for Native and non-Native children by multiplying the total volume of children in foster care by the proportion of the population that was Native and non-Native respectively.
-This approach (as AAIA) acknowledges, likely produced conservative estimates of AIAN foster care caseloads in these states, as Native children were more likely than non-Native children to enter foster care in these states during this period.
-
-These landmark data provided a first close-to-comprehensive national picture of the scale of Indian family separation by combining federal, state, and local data sources.
-These data provided much of the quantitative empirical foundation for the landmark reforms implemented in the Indian Child Welfare Act.
-While AAIA's imputation methods and geographic scope do not match contemporary data collection procedures, these data provide the most comprehensive historical data available to examine the scope of Indian family separation in the period immediately prior to the passage of the Indian Child Welfare Act.
+This approach (as AAIA acknowledges), likely produced conservative estimates of AIAN foster care caseloads in these states, as Native children were more likely than non-Native children to enter foster care in these states during this period.
+
+These landmark data provided a first close-to-comprehensive national picture of the scale of Indian family separation by combining federal, state, and local data sources, and data provided much of the quantitative empirical foundation for the landmark reforms implemented in the Indian Child Welfare Act.
+We use them here to measure change in the exposure of AIAN families to foster care and adoption since the passage of ICWA.
 
@@
 The National Child Abuse and Neglect Data System is a voluntary reporting system that documents all cases of alleged child maltreatment reported to a state or local child welfare hotline that are screened-in and receive an agency response (typically an investigation from a caseworker).
-Since the early 2010s, all 50 states have submitted these report-level data annually to the US Children's Bureau.
-
-Here, we use the NCANDS data to measure two events for AIAN (defined as American Indian or Alaska Native alone or in combination with any other racial or ethnic identification) children and white children (defined as non-hispanic white alone).
+Since the early 2010s, all 50 states have submitted these report-level data annually to the U.S.
+Children's Bureau.
+
+Here, we use the NCANDS data to measure two events for AIAN children (defined as American Indian or Alaska Native alone or in combination with any other racial or ethnic identification) and white children (defined as non-Hispanic white alone).
 Using within-state unique child identifiers, we document the first time a child was the subject of a screened-in CPS report, and the first time a child was the subject of a confirmed or substantiated CPS report over a five-year period (2014 - 2018).
 Note that child identifiers for Georgia, Pennsylvania, and Rhode Island were not reliable over time for children with unsubstantiated screened-in cases.
-These states are exluded from analysis of risk of investigation, but included in all other analyses.
+These states are excluded from analysis of risk of investigation, but included in all other analyses.
 
@@
 
-The AFCARS adoption file provides detailed information on all children newly adopted with state or local child welfare agency involvement in the US each year.
+The AFCARS adoption file provides detailed information on all children newly adopted with state or local child welfare agency involvement in the U.S.
+each year.
 To provide point-in-time estimates of total numbers of children in adoptive homes, we aggregate data from 2010 - 2019, then evaluate the number of both Native and non-Native children in adoptive homes who would have been 21 years old or younger in 2019.
@@
 
-To obtain valid estimates of the AIAN child population by age for US states between 2014 and 2018, we rely on the US Census Population and Housing Units Estimate Program (CITE STATE BY 5 RACE ALONE OR COMB FILE).
-Our definition of American Indian and Alaska Native includes any individual who identifies as American Indian or Alaska Native alone or in combination with any other group.
-The US Census Bureau's Population and Housing Unit Estimates program uses baseline data from the 2010 Census along with various sources of demographic data to estimate age and race specific population totals for all counties in the US annually.
-At the state-level, this program provide single-age estimates of the population that identifies as American Indian / Alaska Native alone or in combination with any other group(s).
-These estimates are imputed by the US Census based on 2010 US Census baseline estimates and various demographic indicators for each locality to project population change.
-
-Unfortunately, we are not able to leverage Census population data directly to make comparisons of the incidence rate of family separation over time expressed as rates per capita.
+To obtain valid estimates of the AIAN child population by age for U.S.
+states between 2014 and 2018, we rely on data from the U.S.
+Census Population Estimates Program.
+Our definition of American Indian and Alaska Native includes any individual who identifies as American Indian or Alaska Native alone or in combination with any other racial or ethnic identity.
+The U.S.
+Census Bureau's Population Estimates program uses baseline data from the 2010 Census along with various sources of demographic data to estimate age and race specific population totals for all counties in the U.S.
+annually.
+At the state-level, this program provides single-age estimates of the population that identifies as American Indian / Alaska Native alone or in combination with any other group(s).
+
+We are not able to leverage Census population data directly to make comparisons of the incidence rate of family separation between the 1970s and 2010s.
 Changes in the enumeration of the AIAN population between 1976 and 2019 make such comparisons impractical.
-For example, the US Census enumerated approximately 1.4 million American Indians and Alaska Natives in the 1980 decennial count, a 70 percent increase over the 1970 decennial census.
-Such dramatic shifts are largely explained by the Census changing its methodology for determining race/ethnicity of a subject.
+For example, the U.S.
+Census counted approximately 1.4 million American Indians and Alaska Natives in the 1980 decennial count, a 70 percent increase over the 1970 decennial census.
+Such dramatic shifts are largely explained by the Census changing its methodology for determining race/ethnicity of a respondent.
 Until 1970, the Census worker reported a subject's race/ethnicity based on worker observations.
-In 1980, the Census changed to self-identification of race/ethnicity (Lujan: American Indians and Alaska Natives Count).
-In 1990, the Census counted about 2 million American Indians and Alaska Native, an approximately 40 percent increase from the prior Census.
+In 1980, the Census changed to self-identification of race/ethnicity @lujan_american_2014.
+In 1990, the Census counted about 2 million American Indians and Alaska Natives, an approximately 40 percent increase from the prior Census.
 By 2000, the Census reported 4.2 million American Indians and Alaska Natives living in the United States, more than double the 2000 population count.
-Such changes are not driven by demography alone (fertlity, mortality, migration), but instead are an artifact of shifting methods of enumerating the racial and ethnic identities of US residents by the Census Bureau.
-Shifting Census methods, enabling individuals to identify their own group membership, and to identify membership in multiple groups have dramatically changed official estimates of the size of the AIAN population.
+Such changes are not driven by population processes alone, but instead are an artifact of shifting methods of enumerating the racial and ethnic identities of U.S.
+residents by the Census Bureau.
 This makes direct comparisons of historical AIAN population counts to contemporary AIAN population counts problematic.
-
-For contemporary risk estimates, we rely on age-specific estimates of the population who identify as AIAN alone or in combination with any other group between 2014 and 2018 by state computed by the Census Population and Housing Units Estimation Program.
-We do not attempt here to make historical comparisons of incident rates of events between the contemporary period and the pre-ICWA period.
 
@@
 For comparability, we compute point-in-time contemporary foster care caseloads by counting the number of children in foster care at the end of the reporting year in 2019, and adoption caseloads by using annual counts of new adoptions from 2010-2019 and estimating the number of children concurrently in adoptive homes under the age of 21 (for comparability with AAIA measures, which used 21 as the age threshold for inclusion).
-We compare counts of AIAN and non-AIAN children in each system for states included in the AIAN data at these two time points, then compare aggregate totals of children in some form of state-involved family separation across these two time periods.
+We compare counts of AIAN and non-AIAN children in each system for states included in the AAIA data at these two time points, then compare aggregate totals of children in some form of state-involved family separation (inclusive of Bureau of Indian Affairs boarding schools) across these two time periods.
 
@@
 
-For each child welfare system outcome examined here (investigations, substantiations, foster care entries, terminations of parental rights), we use 2014 - 2018 NCANDS and AFCARS data to estimate period life tables for each state for AIAN children and white children separately.
+For each child welfare system outcome examined here (investigations, substantiations, foster care entries, terminations of parental rights), we use 2014-2018 NCANDS and AFCARS data to estimate period life tables for each state for AIAN children and white children separately.
 Period life tables share some of the analytic benefits of cohort life tables.
-In a cohort life table, analysts track the incidence of an outcome across a group over the life course, adjusting for mortality and prior-event incidence.
+In a cohort life table, analysts track the incidence of an outcome across a group over the life course, adjusting for mortality and prior incidence of the focal event.
 When a cohort life table is impractical because of data limitations or shifts in risk over time, a period life table allows for a simulation of risk over the life course by assuming that age-specific risks remain static at the levels observed during the focal period.
@@
 In this study, we use 5 years of child welfare and population data to estimate the risk of a first child welfare event for each group and age.
-We then apply these age-specific risk estimates to a hypothetical cohort of 100,000 infants for each group and state, counting the number of events that would have occurred in this cohort at 2014-2018 levels of age-specific risk and age-specific mortality (CITE THE DEMOGRAPHY BOOK).
+We then apply these age-specific risk estimates to a hypothetical cohort of 100,000 infants for each group and state, counting the number of events that would have occurred in this cohort at 2014-2018 levels of age-specific risk and age-specific mortality @preston_demography_2001.
 The key assumption of this method is that event and mortality risk levels remain relatively stable over the period.
-Caution is due in using these data to extrapolate to future periods, as both event risk and mortality risk may shift over time.
+Caution is due in using these data to extrapolate to future event risk levels, as both event risk and mortality risk may shift over time.
 
@@
 
-Period life tables provide us with rigorous estimates of the marginal probability that a child would experience a particular event in a focal population at a fixed level of risk.
+Period life tables provide us with rigorous estimates of the marginal probability that a child would experience a particular event in a focal population at fixed levels of risk.
 In this study, we are interested in identifying the institutional sites at which inequalities emerge or are exacerbated.
@@
 
-To do so, we first estimates period life tables for the joint event of each combination of outcomes.
-For example, we link the NCANDS and AFCARS foster care data using unique child identifiers to estimate the joint probability that a child experience both a CPS investigation *and* a later foster care entry.
-Once we estimate this join probability, we can compute the conditional probability as follows:
+To do so, we first estimate period life tables for the joint event of each combination of outcomes.
+For example, we link the NCANDS and AFCARS foster care data using unique child identifiers to estimate the joint probability that a child experiences both a CPS investigation *and* a later foster care entry.
+Once we estimate this joint probability, we can compute the conditional probability as follows:
 
@@
 These conditional probabilities provide information on the likelihood that a child will move up the case processing ladder from one state to the next.
-By comparing the conditional probabilities of case processing outcomes separately for AIAN and white children across states, we can identify the states of child protection case processing during which inequalities emerge for AIAN children relative to white children.
-
-It is important to note that while state-level analyses provide far more insight into local processes than do national analyses, working at the state-level may obscure important differences in system practices and outcomes occurring at county or local levels.
-This is particularly relevant for states with county-administered child protection systems.
-Because of data-masking procedures in the NCANDS/AFCARS, analyses at the county-level are not consistently possible for non-urban locales.
+By comparing the conditional probabilities of case processing outcomes separately for AIAN and white children across states, we can identify the sites of child protection case processing during which inequalities emerge for AIAN children relative to white children.
+
+It is important to note that while state-level analyses provide far more insight into local processes than do national analyses, working at the state-level may obscure important differences in system practices and outcomes occurring at county, local, or tribal levels.
 
@@
 A small proportion of cases in the NCANDS and AFCARS are missing data on focal variables including child race/ethnicity and caretaker race/ethnicity.
-To address this missing data and ensure that our results are not negatively biased through listwise deletion, we compute multiple imputations of the NCANDS and AFCARS at the report or child level using pseudo Bayesian methods as developed by Van Buuren (https://www.jstatsoft.org/article/view/v045i03).
-These models include child-level predictors such as age and sex in addition to including state-level population composition predictors to adjust for hetereogeneity in the racial composition of populations across states.
+We assume these data are missing at random and compute multiple imputations of the NCANDS and AFCARS at the report or child level using pseudo Bayesian methods developed by Van Buuren @buuren_mice_2011.
+These models include child-level predictors such as age and sex in addition to including state-level population composition predictors to adjust for heterogeneity in the racial composition of populations across states.
 All results reported in the figures below display uncertainty intervals that reflect the maximum and minimum estimates of each statistic, reflecting uncertainty driven by missing data.
@@
 Figure \@ref(fig:icwa-fc-counts) shows the total numbers of children in foster care as reported by AAIA in the mid-1970s and in AFCARS in 2019.
-There were `r icwa_nat_na$fc[1]` Native children in the 19 states for which AAIA collected data in the mid-1970s.
-In 2019, there were `r icwa_nat_na$fc[3]` Native children in foster care, an increase of `r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,1)` AIAN children in foster care over this period.
-For non-Native children in these 19 states, caseloads increased by `r round(((icwa_nat_na$fc[4] / icwa_nat_na$fc[2]) - 1) * 100,1)` percent over the same period.
+There were `r icwa_nat_na$fc[1]` Native children in foster care in the 19 states for which AAIA collected data in the mid-1970s.
+In 2019, there were `r icwa_nat_na$fc[3]` Native children in foster care in these states, an increase of `r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,0)` percent.
+For non-Native children in these 19 states, caseloads increased by `r round(((icwa_nat_na$fc[4] / icwa_nat_na$fc[2]) - 1) * 100,0)` percent over the same period.
 
@@
 Note that 6 of the 19 included states lack complete data on AIAN children in adoption in the AAIA data (Idaho, Maine, Nevada, New Mexico, New York, and Wyoming).
-For these 13 states with complete adoption data, AAIA estimated that there were `r icwa_nat$adopted[1]` Native children in state-involved adoptions in the mid-1970s.
-In these same 13 states, there were `r icwa_nat$adopted[3]` AIAN children in adoptions in 2019, an increase of `r round(((icwa_nat_na$adopted[3] / icwa_nat_na$adopted[1]) - 1) * 100,1)` percent.
-For non-native children in these states, there was a `r round(abs(((icwa_nat_na$adopted[2] / icwa_nat_na$adopted[4]) - 1) * 100),1)` percent reduction in the numbers of adopted children over the same period.
+For the 13 states with complete adoption data, AAIA estimated that there were `r icwa_nat$adopted[1]` Native children in state-involved adoptions in the mid-1970s.
+In these same 13 states, there were `r icwa_nat$adopted[3]` AIAN children in adoptions in 2019, an increase of `r round(((icwa_nat_na$adopted[3] / icwa_nat_na$adopted[1]) - 1) * 100,0)` percent.
+For non-native children in these states, there was a `r round(abs(((icwa_nat_na$adopted[2] / icwa_nat_na$adopted[4]) - 1) * 100),0)` percent reduction in the numbers of adopted children over the same period.
 
@@
 
-Figure \@ref(fig:icwa-pct) shows the relative change in adoptive and foster care caseloads between the mid-1970s and 2019 for these large Native population states.
-Of the 19 states with complete data, 14 saw growth in the numbers of Native children in foster care since the passage of ICWA.
+Figure \@ref(fig:icwa-pct) shows the relative change in adoption and foster care caseloads between the mid-1970s and 2019 for these states.
+Of the 19 states, 14 saw growth in the numbers of Native children in foster care since the passage of ICWA.
 Of the 13 states with complete adoption data, 8 had more Native children in foster care in 2019 than in the mid-1970s prior to the passage of ICWA.
@@
 Oklahoma saw the largest absolute growth in Native foster care caseloads.
-In the mid-1970s, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 1976) %>% select(fc) ` Native chlidren in foster care in Oklahoma.
-In 2019, here were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 2019) %>% select(fc) ` Native children in foster care in Oklahoma, a growth of nearly `r round(icwa_delta_pct %>% filter(state=="OK", name == "Foster care", race_ethn == "AIAN") %>% select(value), 0)` percent.
+In the mid-1970s, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 1976) %>% select(fc) ` Native children in foster care in Oklahoma.
+In 2019, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 2019) %>% select(fc) ` Native children in foster care in Oklahoma, a growth of nearly `r round(icwa_delta_pct %>% filter(state=="OK", name == "Foster care", race_ethn == "AIAN") %>% select(value), 0)` percent.
 
@@
 There were more AIAN children in boarding schools in 1973 than were in foster care in 2019.
-While the scale of separation of Native children from their families through foster care and adoption has increased over time in these states, the absolute scale of the separation of Native children from their families has likely declined nationally with the end of the boarding school era.
+While the scale of separation of Native children from their families through foster care and adoption has increased over time in these states, the absolute scale of the separation of Native children from their families has declined nationally with the end of the boarding school era.
 
@@
 
-All estimates should be interpreted as risks at 2014 - 2018 levels. Caution should be exercised in using these estimates to project risk to future periods.
-
 ### Age-specific risks
@@
 Figure \@ref(fig:nat-marginal) shows the marginal probabilities of experiencing a series of child welfare system events for both Native and non-Native children.
-This plot shows national age-specific event probabilities at 2014-2018 risk levels for AIAN and non-AIAN children for experiencing 1) a screened-in investigation, 2) a confirmed maltreatment case, 3) a foster care placement, and 4) termination of parental rights through the child welfare system.
-
-```{r nat-marginal, fig.cap = "Marginal probability of child welfare event incidence for AIAN children by age 18, 2014 - 2018 risk levels, US totals"}
+This plot shows national age-specific event probabilities at 2014-2018 risk levels for AIAN and non-AIAN children for 1) a screened-in investigation, 2) a confirmed maltreatment case, 3) a foster care placement, and 4) termination of parental rights through the child welfare system.
+
+```{r nat-marginal, fig.cap = "Age-specific risks of child welfare event incidence for AIAN children by age 18, 2014 - 2018 risk levels, U.S. totals"}
 
@@
 About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Investigation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of Native infants experience a CPS investigation nationally, and about `r plot_dat %>% filter(age==0, race_ethn=="White", var=="Investigation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of white infants experience a CPS investigation.
-About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Substantiation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of AIAN children experience an agency confirmed maltreatment case before their first birthday, compared with about `r plot_dat %>% filter(age==0, race_ethn=="White", var=="Substantiation") %>% select(q) %>% mutate(q = round(q*100, 1))`.
-About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Foster Care") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of AIAN children are removed from their families and placed into foster care before their first birthday, and about `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Termination") %>% select(q) %>% mutate(q = round(q*100, 1))`.
+About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Substantiation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of AIAN children experience an agency confirmed maltreatment case before their first birthday, compared with about `r plot_dat %>% filter(age==0, race_ethn=="White", var=="Substantiation") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of white children.
+About `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Foster Care") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of AIAN children are removed from their families and placed into foster care before their first birthday, compared to about `r plot_dat %>% filter(age==0, race_ethn=="AIAN", var=="Termination") %>% select(q) %>% mutate(q = round(q*100, 1))` percent of white children.
 Risk of a first event declines steadily with age for both groups.
 
-Nationally at 2014 - 2018 risk levels, white children are at higher risk of investigation at all ages, risks of substantiation are similar for white and Native children, risks of foster care are higher for native children, and risks for termination are slightly higher for native children.
-
-Of course, there is substantial variation in risk of child welfare system contact across US states.
+Nationally at 2014 - 2018 risk levels, white children are at higher risk of investigation at all ages, risks of substantiation are similar for white and Native children, risks of foster care are higher for Native children, and risks for termination are slightly higher for Native children.
+
+Of course, there is substantial variation in risk of child welfare system contact across U.S.
+states.
 States have tremendous flexibility in designing and implementing both child welfare surveillance systems and foster care systems.
 The character of government-to-government relationships between tribes and states varies tremendously across places.
-Federal law also differentially affects tribal governance and the relationships between states, counties and tribes across places (CITES).
-Figure\@ref(fig:nat-inv-age) shows the age-specific risk of experiencing a first screened-in investigation by US state for both AIAN and white children at 2014 - 2018 levels of risk.
+Federal law also differentially affects tribal governance and the relationships between states, counties and tribes across places @pevar_rights_2012.
+Figure \@ref(fig:nat-inv-age) shows the age-specific risk of experiencing a first screened-in investigation by U.S.
+state for both AIAN and white children at 2014 - 2018 levels of risk.
 
@@
 Infants are at highest risk of investigation in all states, and the shape of the age-risk curve is similar to the national pattern shown in Figure \@ref(fig:nat-marginal).
-Risk tends to be markedly higher for infants, and decline slowly with age.
+Risk tends to be markedly higher for infants and declines slowly with age.
 However, there is tremendous variation in levels of risk across states, and clear regional patterning in inequalities in the exposure of infants to CPS investigation.
@@
 In 2014 - 2018, about `r temp %>% filter(state=="AK", race_ethn=="AIAN") %>% select(q) %>% mutate(q = round(q * 100, 1))` percent of AIAN infants had a screened-in CPS report.
-More than 10 percent of AIAN infants experienced an investigation in 3 other states: Minnesota, Oklahoma, and Montana.
-White children in Southern states like Mississippi, West Virginia, Florida, and Kentucky are typically at  higher levels of risk of investigation at all ages that are AIAN children, whose risk levels are typically quite low.
+Risk levels were at 10 percent or higher in 3 other states: Minnesota, Oklahoma, and Montana.
+White children in Southern states like Mississippi, West Virginia, Florida, and Kentucky are typically at  higher levels of risk of investigation at all ages that are AIAN children, whose risk levels are typically low.
 In The Great Plains and Upper Midwest states, AIAN children typically have much higher investigation risk than do white children.
@@
   theme_minimal() + 
-  labs(x = "Age", y = "Proportion of children") + 
+  labs(x = "Age", y = "Probability") + 
   theme(strip.text.x = element_text(size = 7),
@@
 
-In Minnesota, at 2014 - 2018 levels of risk, 9.7 percent of AIAN children could be expected to be removed from their families and placed into foster care before their first birthday.
-This is the highest level of risk in the country for infants of any racial or ethnic group (CITE YI EDWARDS WILDEMAN WORKING PAPER).
+Figure \@ref(fig:nat-fc-age) shows the age-specific risk of entering foster care for the first time for all U.S.
+States at 2014 - 2018 levels of risk.
+In Minnesota, 9.7 percent of AIAN children could be expected to be removed from their families and placed into foster care before their first birthday.
+This is the highest level of risk in the country for infants of any racial or ethnic group.
 AIAN infants are also at exceptionally high risk of foster care placement in Montana, Alaska, South Dakota, North Dakota, Oklahoma, Washington, Oregon, Nebraska, and Wisconsin.
-At 2014 - 2018 risk levels, more than 3 percent of all AIAN infants would be placed into foster care.
+At 2014 - 2018 risk levels, more than 3 percent of all AIAN infants would be placed into foster care in these states.
 White children face a 3 percent risk of foster care before their first birthday in only one state: West Virginia.
@@
 
-```{r nat-fc-age, fig.cap = "Age specific probability of entering foster care, 2014 - 2018 risk levels"}
+```{r nat-fc-age, fig.cap = "Age specific risk of entering foster care, 2014 - 2018 risk levels"}
 ggplot(fc_tables %>% 
@@
 
-```{r sub-high, fig.cap = "Age-specific probability of agency-confirmed maltreatment case in high-risk states, 2014 - 2018 risk levels"}
+```{r sub-high, fig.cap = "Age-specific risk of agency-confirmed maltreatment case in high-risk states, 2014 - 2018 risk levels"}
 sub_high <- subst_tables %>% 
@@
   facet_wrap(~state)+
-  labs(x = "Age", y = "Proportion of children") + 
+  labs(x = "Age", y = "Probability") + 
   # theme(strip.text.x = element_text(size = 7),
@@
 
-AIAN children are at higher risk nationally of experiencing the full termination of parental rights while in foster care than are children from any other racial or ethnic group (CITE WILDEMAN EDWARDS WAKEFIELD).
+AIAN children are at higher risk nationally of experiencing the full termination of parental rights while in foster care than are children from any other racial or ethnic group @wildeman_cumulative_2020.
 Like other outcomes, risk is at its peak for infants.
@@
 
-```{r tpr-high, fig.cap = "Age-specific probability of full termination of parental rights, 2014 - 2018 risk levels"}
+```{r tpr-high, fig.cap = "Age-specific risk of full termination of parental rights, 2014 - 2018 risk levels"}
 tpr_high <- tpr_tables %>% 
@@
 
-While there is a very clear relationship between the risk of experiencing child welfare system events and age -- infants are at much greater risk than other children, then risk tpyically declines slowly with age -- there is tremendous difference in the frequency of these events across places, and clear regional patterns in differences in risk between Native and white children.
-
 ### Lifetime event risks and inequalities in lifetime risk
 
-Next, we provide a summary of lifetime incidence of each child welfare event for Native and white children across US states.
+Next, we provide a summary of lifetime incidence of each child welfare event for Native and white children across U.S.
+states.
 These cumulative risks simulate the total exposure of the population of children to a first event by applying age-specific risks and mortality risks (computed from the observed data) to a hypothetical cohort of 100,000 children in each state.
-They can be interpreted as lifetime risks of child welfare system event incidence for each group.
 
@@
 
-We estimate that about 1 in 3 Native children in Minnesota will ever enter foster care at levels of risk similar to those observed between 2014 - 2018.
+We estimate that about 1 in 3 Native children in Minnesota will enter foster care at levels of risk similar to those observed between 2014 - 2018.
 In Montana, this risk is about 1 in 4.
@@
 
-```{r nat-cond, fig.cap = "Risk of child welfare system event, conditional on prior system event at 2014 - 2018 levels of risk"}
+```{r nat-cond, fig.cap = "Age-specific risk of child welfare system event, conditional on prior system event at 2014 - 2018 levels of risk"}
 
@@
 
-We display the distribution of age-specific probabilities of substantiation, conditional on investigation across US states in Figure \@ref(fig:sub-inv).
-Note that states with relatively small AIAN populations, such as Vermont and Delaware, have relatively unstable estimate for these conditional probabilities, as small changes in event counts can drive large fluctuation in relative rate estimates when population denominators are relatively small.
+We display the distribution of age-specific probabilities of substantiation, conditional on investigation across U.S.
+states in Figure \@ref(fig:sub-inv).
+Note that states with relatively small AIAN populations, such as Vermont and Delaware, have relatively unstable estimates for these conditional probabilities, as small changes in event counts can drive large fluctuations in relative rate estimates when population denominators are relatively small.
 The probability of substantiation, conditional on investigation is highest for infants.
@@
 
-When interpreting Figures \@ref(fig:sub-inv) \@ref(fig:fc-inv) \@ref(fig:fc-sub) jointly, along with Figure \@ref(fig:nat-inv-age), a clear pattern emerges.
+When interpreting Figures \@ref(fig:sub-inv), \@ref(fig:fc-inv), and \@ref(fig:fc-sub) jointly, along with Figure \@ref(fig:nat-inv-age), a clear pattern emerges.
 Much of the inequality experienced by AIAN children emerges once a child welfare system begins processing a case.
-With certain exceptions (Alaska, in particular), inequalities typically emerge post-investigation.
-Once AIAN children come into contact with child welfare systems, they are often at much higher risk of substantiation that their peers, and much more likely to be placed into foster care than their peers in both substantiated and unsubstantiated cases.
-The unequal treatment of AIAN children in the child welfare system is dramatically accelerated at the at the point of the decision to place a child into foster care.
+With certain exceptions (Alaska, in particular), inequalities emerge post-investigation.
+In most states, it is not surveillance and reporting alone driving unequal outcomes.
+Once AIAN children come into contact with child welfare systems, they are often at much higher risk of substantiation than their peers, and much more likely to be placed into foster care.
+The unequal treatment of AIAN children in the child welfare system is dramatically accelerated during the decision to place a child into foster care.
 
@@
 Infants have the highest likelihood of being placed with kin or with a native family.
-However, if 41 US States, we estimate that more than 50 percent of AIAN infants are placed in a non-native non-kin foster setting.
+However, in 41 U.S.
+States, we estimate that more than 50 percent of AIAN infants are placed in a non-native non-kin foster setting.
 Maryland, New Hampshire, Connecticut, Pennsylvania, North Carolina, and Florida have the highest rate of ICWA preferred placements for infants, ranging between 54 and 61 percent of infants placed with either kin or an AIAN foster family.
@@
 The likelihood that an AIAN child will be placed with a family member or other Native foster caretaker typically decreases with age.
-There is generally a very small chance that an AIAN teenager placed into foster care will be place with kin or with an AIAN foster family.
+There is generally a very small chance that an AIAN teenager placed into foster care will be placed with kin or with an AIAN foster family.
 
@@
 
-This report has provided two sets of empirical findings.
+This report provides two sets of empirical findings.
 
 First, we examine data on the breadth of the separation of AIAN children prior to the passage of the Indian Child Welfare Act, and compare these data to contemporary AIAN family separation through the child welfare system.
-We find that, while the overall scale of AIAN family separation has decreased since ICWA due to the closing of BIA residential boarding schools, the scale of AIAN family separation through fostering and adoption has increased in the states for which we have comparable data.
+We find that, while the overall scale of AIAN family separation has decreased since ICWA due to the closing of BIA boarding schools, the scale of AIAN family separation through fostering and adoption has increased in the states for which we have comparable data.
 In some states, like Oklahoma, Washington, California, and Alaska, the frequency of AIAN foster care and adoption has increased dramatically since the mid-1970s.
 
-These results should be interpreted with several limitations in mind, however.
-Because of changes in the ways that the US Census tabulates AIAN populations, and because of cultural shifts in how AIAN people identify themselves when asked questions about racial identity on surveys like the Census (CITES), estimates of the size of the AIAN child population are not comparable between 1976 and 2014 - 2018.
+However, these results should be interpreted with several limitations in mind.
+Because of changes in the ways that the U.S.
+Census tabulates AIAN populations, and because of cultural shifts in how AIAN people identify themselves when asked questions about racial identity on surveys like the Census @lujan_american_2014, estimates of the size of the AIAN child population are not comparable between 1976 and 2014 - 2018.
 As such, we cannot directly compare population adjusted incidence rates, only raw counts.
@@
 
-Second, we examine data on the contemporary contact of AIAN children with the child welfare system at 5 distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights.
+Second, we examine data on the contemporary contact of AIAN children with the child welfare system at five distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights.
 We find that AIAN children are more likely than white children to be placed into foster care, even when rates of investigation and substantiation of cases are similar to those for white children.
@@
 In nearly all states, AIAN children can expect to be placed in a non-kin non-Native setting at some point during their time in foster care.
+This provides further evidence that full compliance with ICWA's placement preferences is yet to be achieved.
 
 # Conclusion
+
+This study provides new evidence on the enduring and punitive impact of the child welfare system on Native families.
+We detail the institutional pathways of Native families into the child welfare system, evaluate the magnitude of change of family separation since the passage of the ICWA, and identify geographic variation in child welfare inequality.
+These findings suggest that the troubling history of Native family separation has not yet ended.
+
+# Acknolwedgments {-}
+
+This project received financial support from Casey Family Programs.
+The authors thank Wade Carlson, Brian Clapier, Steve Ellis, Anita Fineday, Casey Groat, Bert Hirsch, and Christopher Wildeman for advice and feedback.
+The contents of this report are solely the responsibility of the authors and do not necessarily represent the views of Casey Family Programs.
+
+# References

### 2020-12-31 visuals_cjlr.r
update cjlr images with PEP data change

@@ -23,7 +23,9 @@ theme_set(theme_bw())
 # 
 # save.image("./cjlr.RData")
 
-load("./cjlr.RData")
+### use data from cfp final report, update cjlr.RData later
+load("./cfp_final_report.RData")
+
 
 ####################################################################################
 ############## ICWA figures
@@ -311,7 +313,7 @@ ggplot(plot_dat,
 map_dat<-investigation_tables_c %>% 
   filter(race_ethn=="AIAN") %>% 
   group_by(state) %>% 
-  summarize(c = mean(c_c)) 
+  summarize(c = mean(c)) 
 
 ### xwalk state maps onto abbreviations
 us_map<-us_map()
@@ -339,7 +341,7 @@ inv_map<-ggplot(us_map %>%
 map_dat<-subst_tables_c %>% 
   filter(race_ethn=="AIAN") %>% 
   group_by(state) %>% 
-  summarize(c = mean(c_c)) 
+  summarize(c = mean(c)) 
 
 us_map<-us_map()
 
@@ -364,7 +366,7 @@ sub_map<-ggplot(us_map %>%
 map_dat<-fc_tables_c %>% 
   filter(race_ethn=="AIAN") %>% 
   group_by(state) %>% 
-  summarize(c = mean(c_c)) 
+  summarize(c = mean(c)) 
 
 us_map<-us_map()
 
@@ -389,7 +391,7 @@ fc_map<-ggplot(us_map %>%
 map_dat<-tpr_tables_c %>% 
   filter(race_ethn=="AIAN") %>% 
   group_by(state) %>% 
-  summarize(c = mean(c_c)) 
+  summarize(c = mean(c)) 
 
 us_map<-us_map()
 
@@ -457,13 +459,13 @@ map_dat<-investigation_tables_c %>%
 map_dat_wht<-map_dat %>% 
   ungroup() %>% 
   filter(race_ethn == "White") %>% 
-  rename(c_white = c_c) %>% 
+  rename(c_white = c) %>% 
   select(-race_ethn, -c, -race_ethn)
 
 map_dat<-map_dat %>% 
   filter(race_ethn=="AIAN") %>% 
   left_join(map_dat_wht) %>% 
-  mutate(disp = c_c / c_white) 
+  mutate(disp = c / c_white) 
 
 ### slice into quantiles
 ### xwalk state maps onto abbreviations

### 2021-01-01 visuals_cjlr.r
fast forward

@@ -454,13 +454,15 @@ map_dat<-investigation_tables_c %>%
               filter(c == min(c)) %>% 
               mutate(var = "TPR")) %>% 
   select(-.imp) %>% 
-  distinct()
+  distinct() %>% 
+  ungroup() %>% 
+  complete(state, race_ethn, var)
 
 map_dat_wht<-map_dat %>% 
   ungroup() %>% 
   filter(race_ethn == "White") %>% 
   rename(c_white = c) %>% 
-  select(-race_ethn, -c, -race_ethn)
+  select(-race_ethn)
 
 map_dat<-map_dat %>% 
   filter(race_ethn=="AIAN") %>% 
@@ -504,7 +506,6 @@ us_map<-us_map %>%
                                         "3-4",
                                         "4+")))
 
-
 ggplot(us_map,
        aes(x = x, y = y, group = group,
            fill = rate_ratio)) + 
@@ -512,6 +513,6 @@ ggplot(us_map,
   coord_fixed() +
   facet_wrap(~var) + 
   theme_void() + 
-  scale_fill_brewer(palette = "Purples") +
-  labs(fill = "AIAN / White\nRate Ratio") + 
+  scale_fill_brewer(palette = "Purples", na.value = "grey45") +
+  labs(fill = "AIAN / White\nRate Ratio") +
   ggsave("./vis/cjlr/5.png", width = 8, height = 4)

### 2021-08-22 state_ms.Rmd
add state ms

--- 2020-12-31 cfp_final_report.Rmd
@@
 ---
-title: "Mechanisms of AIAN Inequality in Child Welfare Across U.S. States."
+title: "American Indian and Alaska Native child welfare system contact across U.S. States: magnitudes and mechanisms"
 author: "Frank Edwards and Theresa Rocha Beardall"
-date: "December, 2020"
-bibliography: "casey_rpt.bib"
+bibliography: "lib.bib"
 link-citations: true
-csl: "vancouver-brackets-no-et-al.csl"
-abstract: "American Indian and Alaska Native children are separated from their families by state child welfare agencies at exceptionally high rates.
-This study connects contemporary trends in Native family separation to histories of Indian child removal, and provides insights into the geographies and institutional sites where inequalities emerge.
-We find that the total number of AIAN children in foster care or adoptive homes in states with large Native populations has increased since the passage of the Indian Child Welfare Act (1978).
+#csl: "vancouver-brackets-no-et-al.csl"
+abstract: "Following a long history of forced assimilation through family separation, American Indian and Alaska Native children continue to be separated from their families by state child welfare agencies at exceptionally high rates.
+This study provides insights into the geographies and institutional sites where inequalities emerge.
 We find that risks of child welfare system contact are highest for AIAN infants, and that risk is highly variable across states.
@@
 output: bookdown::pdf_document2
+# output: word_document
+toc: false
 ---
@@
 
-Despite growing attention to threats against the health, safety, and wellbeing of tribal communities in recent years, the child welfare system remains a confounding and persistently complex source of harm in the lives of Native families throughout the U.S.
+The child welfare system remains a confounding and persistently complex source of harm in the lives of Native families throughout the United States [@haight_scoping_2018; @dettlaff_racial_2021].
 These harms stem from institutional acts of violence waged against Native Peoples by the U.S.
-federal government.
+government [@glenn_settler_2015; @wolfe_settler_2006].
 This began with early Americans who were eager to physically remove Native Peoples from their homelands in order to establish commerce, statehood, and homes for non-Native families.
-Often, the intention of such violence was to break up Native families and assimilate them into the nation’s population.
-Boarding schools and adoption programs carried the U.S.
-mission of assimilating Native children well into the 20th century.
-
-Family separation continues to jeopardize and threaten the autonomy and wellbeing of Native families, and as a result, the very existence of tribal communities.
-These threats remain despite the passage of the landmark Indian Child Welfare Act (ICWA) of 1978, a law that was implemented to address and end generations of abuse, mistreatment, and forcible removal of Native children from their homes.
-
-In the following study, we describe the scale of the ongoing crisis of family separation by 1) evaluating the magnitude of change of family separation since ICWA, 2) identifying geographic variation in child welfare inequality, and 3) identifying the stages in case processing where inequality is accelerated.
-This study provides new empirical evidence about where and how Native families are being pushed into the child welfare system, and situates these findings in the long historical legacy of Native family separation.
+During the late 19th through the mid-20th century, federal, state, and religious organizations worked to aggressively break up Native families and assimilate them into the nation’s white population through boarding schools, fostering, and adoption [@adams_education_1995; @trennert_phoenix_1988; @jacobs_generation_2014; @cross_american_1999].
+Genocidal efforts to eradicate Native nations and cultures through the removal and separation of indigenous children persisted with explicit support from federal and state governments and leading social work organizations until the passage of the landmark Indian Child Welfare Act (ICWA) of 1978.
+ICWA provided tribal governments critical tools to exercise tribal sovereignty in US family courts and provides a critical framework for protecting Native families and respecting tribal sovereignty [@pevar_rights_2012; @jacobs_generation_2014].
+Despite these protections, family separation continues to jeopardize and threaten the autonomy and wellbeing of Native families, and as a result, the very existence of tribal communities [@beardall_abolition_2021].
+
+In this study, we describe the scale of the ongoing crisis of family separation by 1) identifying geographic patterns in the exposure of Native children and families to child welfare system contact and 2) identifying the stages in child welfare case processing where inequality is accelerated for Native children and families.
+This study provides new empirical evidence about where and how American Indian and Alaska Native families are being pushed into the child welfare system, and situates these findings in the long historical legacy of Native family separation.s
 
 # Data
-
-This study relies on a number of data sources.
-First, we compile and digitize statistics collected by the Association for American Indian Affairs during the late 1960s and early to mid 1970s in an effort to document the breadth and depth of the Indian family separation crisis @ninety-fifth_congress_indian_1977.
-To capture the contemporary exposure of AIAN children to the child welfare system, we rely on federal data systems collected by the U.S.
-Children's Bureau.
-We use the 2014 - 2018 National Child Abuse and Neglect Data System (NCANDS) child files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_national_2018, the 2014 - 2018 Adoption and Foster Care Analysis and Reporting System (AFCARS) foster care files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_adoption_2018 and the 2010 - 2018 AFCARS adoption files @u_s_department_of_health_and_human_services__administration_on_children_youth_and_families_adoption_2018-1.
-We rely on the U.S.
-Census Bureau's Population Estimates Program (PEP) for contemporary population estimates @us_census_bureau_age_2020.
-
-## Historical child welfare system data
-
-The Association on American Indian Affairs submitted a report titled "Indian Child Welfare Statistical Summary, July 1976" in support of the Indian Child Welfare Act (1978) during congressional hearings, included as Appendix G in hearings convened by the Select Committee on Indian Affairs on August 4, 1977.
-The results presented by AAIA were the result of multiple waves of surveys sent by AAIA to federal, state, local, and private service providers in 19 states with large AIAN populations (Alaska, Arizona, California, Idaho, Maine, Michigan, Minnesota, Montana, Nevada, New Mexico, New York, North Dakota, Oklahoma, Oregon, South Dakota, Utah, Washington, Wisconsin, and Wyoming).
-AAIA compiled statistics on the characteristics of both Native and non-Native children in foster care, in adoption, or in boarding programs for each of these states.
-
-These data consistently provide point-in-time estimates of the numbers of Native and non-Native children in foster care, adoptions, or in Bureau of Indian Affairs boarding schools.
-During this period, some states did not collect information on race/ethnicity for children in foster care.
-When this was the case, AAIA imputed foster care caseload totals for Native and non-Native children by multiplying the total volume of children in foster care by the proportion of the population that was Native and non-Native respectively.
-This approach (as AAIA acknowledges), likely produced conservative estimates of AIAN foster care caseloads in these states, as Native children were more likely than non-Native children to enter foster care in these states during this period.
-
-These landmark data provided a first close-to-comprehensive national picture of the scale of Indian family separation by combining federal, state, and local data sources, and data provided much of the quantitative empirical foundation for the landmark reforms implemented in the Indian Child Welfare Act.
-We use them here to measure change in the exposure of AIAN families to foster care and adoption since the passage of ICWA.
-
-## Contemporary child welfare system data
 
@@
 
-The AFCARS adoption file provides detailed information on all children newly adopted with state or local child welfare agency involvement in the U.S.
-each year.
-To provide point-in-time estimates of total numbers of children in adoptive homes, we aggregate data from 2010 - 2019, then evaluate the number of both Native and non-Native children in adoptive homes who would have been 21 years old or younger in 2019.
-We use the 21-year-old threshold for comparability with data collected by AAIA.
-
-## Population data
-
 To obtain valid estimates of the AIAN child population by age for U.S.
@@
 
-We are not able to leverage Census population data directly to make comparisons of the incidence rate of family separation between the 1970s and 2010s.
-Changes in the enumeration of the AIAN population between 1976 and 2019 make such comparisons impractical.
-For example, the U.S.
-Census counted approximately 1.4 million American Indians and Alaska Natives in the 1980 decennial count, a 70 percent increase over the 1970 decennial census.
-Such dramatic shifts are largely explained by the Census changing its methodology for determining race/ethnicity of a respondent.
-Until 1970, the Census worker reported a subject's race/ethnicity based on worker observations.
-In 1980, the Census changed to self-identification of race/ethnicity @lujan_american_2014.
-In 1990, the Census counted about 2 million American Indians and Alaska Natives, an approximately 40 percent increase from the prior Census.
-By 2000, the Census reported 4.2 million American Indians and Alaska Natives living in the United States, more than double the 2000 population count.
-Such changes are not driven by population processes alone, but instead are an artifact of shifting methods of enumerating the racial and ethnic identities of U.S.
-residents by the Census Bureau.
-This makes direct comparisons of historical AIAN population counts to contemporary AIAN population counts problematic.
-
 # Methods
-
-## Historical comparisons
-
-We directly compare counts of children experiencing family separation through adoption and foster care in the states surveyed by AAIA in 1976 to counts of children in foster care or adoption in 2019.
-Because historical and contemporary AIAN population estimates are incompatible, we do not provide estimates as per capita rates.
-AAIA reports point-in-time counts of children in foster care or adoption by state.
-For comparability, we compute point-in-time contemporary foster care caseloads by counting the number of children in foster care at the end of the reporting year in 2019, and adoption caseloads by using annual counts of new adoptions from 2010-2019 and estimating the number of children concurrently in adoptive homes under the age of 21 (for comparability with AAIA measures, which used 21 as the age threshold for inclusion).
-We compare counts of AIAN and non-AIAN children in each system for states included in the AAIA data at these two time points, then compare aggregate totals of children in some form of state-involved family separation (inclusive of Bureau of Indian Affairs boarding schools) across these two time periods.
 
@@
 # Findings
-
-## Change Since ICWA
-
-```{r icwa-setup}
-
-### ICWA DATA NAT
-
-icwa_nat<-icwa_fc %>%
-  group_by(race_ethn) %>% 
-  summarize(fc = sum(fc, na.rm=T),
-            adopted = sum(adopted, na.rm=T),
-            boarding = sum(boarding, na.rm=T))
-
-##
-
-afcars_caseload19<-read_csv("./data/afcars19_caseload.csv")
-afcars_caseload19<-afcars_caseload19 %>% 
-  mutate(race_ethn = ifelse(AMIAKN==1,
-                            "AIAN",
-                            "other")) %>% 
-  rename(state = St,
-         st_fips = STATE,
-         fc = caseload) %>% 
-  filter(st_fips!=72) %>% 
-  select(st_fips, race_ethn, fc) %>% 
-  filter(!is.na(race_ethn))
-
-### join with afcars adoption data
-afcars_caseload<-afcars_caseload19 %>% 
-  left_join(afcars_adopt_caseload_19 %>% 
-  mutate(race_ethn = ifelse(race_ethn=="AIAN",  
-                            race_ethn,
-                            "other")) %>% 
-  mutate(n_non_preferred = ifelse(race_ethn=="AIAN",
-                                  n - n_kin_or_aian,
-                                  n - n_kin),
-         adopt_non_preferred = n_non_preferred / n) %>% 
-    rename(st_fips = STATE,
-           state = St,
-           adopted = n) %>% 
-  select(state, race_ethn, adopted, adopt_non_preferred)) %>% 
-  mutate(state,  race_ethn, fc, adopted) %>% 
-  mutate(period = "2019")
-
-### state comparisons
-icwa_delta<-icwa_fc %>% 
-  mutate(race_ethn = ifelse(race_ethn=="White",
-                            "other",
-                            race_ethn)) %>% 
-  bind_rows(afcars_caseload %>% 
-              select(state, period, race_ethn, fc, adopted)) %>% 
-  filter(state%in%icwa_fc$state)
-
-### make comparable national totals
-### can't compare per cap rates
-### pop denominators aren't comparable
-
-icwa_delta_pct<-
-  icwa_delta %>% 
-  pivot_wider(id_cols = c(state, race_ethn),
-              names_from = period, 
-              values_from = c(fc, adopted)) %>% 
-  mutate(delta_fc = (fc_2019 / fc_1976 -1) * 100,
-         delta_adopt = (adopted_2019/adopted_1976 -1)*100) %>% 
-  select(state, race_ethn, delta_fc, delta_adopt) %>% 
-  pivot_longer(cols = delta_fc:delta_adopt) %>% 
-  mutate(name = ifelse(
-    name=="delta_fc",
-    "Foster care",
-    "Adoption"
-  ))
-
-### remove adoption states where we don't have AIAN ICWA numbers
-index<-icwa_delta_pct %>% 
-  filter(is.na(value))
-
-icwa_delta_pct <- icwa_delta_pct %>% 
-  mutate(value = ifelse(state%in%index$state & name == "Adoption",
-         NA,
-         value))
-
-icwa_nat<-icwa_delta %>% 
-  filter(!state%in%index$state) %>% 
-  group_by(period, race_ethn) %>% 
-  summarise(fc = sum(fc),
-            adopted = sum(adopted))
-
-### plus boarding
-total_inst<-icwa_delta %>% 
-  group_by(period, race_ethn) %>% 
-  summarise(fc = sum(fc),
-            adopted = sum(adopted, na.rm=T)) %>% 
-  mutate(total = fc + adopted)
-
-icwa_nat_na<-icwa_delta %>% 
-  group_by(period, race_ethn) %>% 
-  summarise(fc = sum(fc, na.rm=T),
-            adopted = sum(adopted, na.rm=T))
-```
-
-We compare data collected by the Association on American Indian Affairs between 1973 and 1976 on American Indian and Alaska Native children in foster care and adoption to contemporary data on AIAN children in foster care and adoption collected through the Adoption and Foster Care Analysis and Reporting System between 2010 and 2019.
-Figure \@ref(fig:icwa-fc-counts) shows the total numbers of children in foster care as reported by AAIA in the mid-1970s and in AFCARS in 2019.
-There were `r icwa_nat_na$fc[1]` Native children in foster care in the 19 states for which AAIA collected data in the mid-1970s.
-In 2019, there were `r icwa_nat_na$fc[3]` Native children in foster care in these states, an increase of `r round(((icwa_nat_na$fc[3] / icwa_nat_na$fc[1]) - 1) * 100,0)` percent.
-For non-Native children in these 19 states, caseloads increased by `r round(((icwa_nat_na$fc[4] / icwa_nat_na$fc[2]) - 1) * 100,0)` percent over the same period.
-
-```{r icwa-fc-counts, fig.cap = "Children in foster care on reporting date (caseload) by state and period. Data from AAIA surveys and AFCARS"}
-ggplot(icwa_delta %>% 
-         mutate(race_ethn = 
-                  ifelse(race_ethn=="other",
-                         "Non-AIAN",
-                         race_ethn)),
-       aes(y = state, x = fc, fill = period)) + 
-  geom_col(position = position_dodge2(reverse = T)) + 
-  facet_wrap(~race_ethn, scales = "free") + 
-  scale_fill_brewer(palette = "Dark2") + 
-  labs(y = "",
-       x = "Children in foster care",
-       fill = "") 
-```
-
-Figure \@ref(fig:icwa-adopt-counts) displays counts of children 21 years of age or younger in state-involved adoptions for both the mid-1970s and 2019.
-Note that 6 of the 19 included states lack complete data on AIAN children in adoption in the AAIA data (Idaho, Maine, Nevada, New Mexico, New York, and Wyoming).
-For the 13 states with complete adoption data, AAIA estimated that there were `r icwa_nat$adopted[1]` Native children in state-involved adoptions in the mid-1970s.
-In these same 13 states, there were `r icwa_nat$adopted[3]` AIAN children in adoptions in 2019, an increase of `r round(((icwa_nat_na$adopted[3] / icwa_nat_na$adopted[1]) - 1) * 100,0)` percent.
-For non-native children in these states, there was a `r round(abs(((icwa_nat_na$adopted[2] / icwa_nat_na$adopted[4]) - 1) * 100),0)` percent reduction in the numbers of adopted children over the same period.
-
-```{r icwa-adopt-counts, fig.cap = "Children 21 and under in state-involved adoptions on reporting date by state and period. Data from AAIA surveys and AFCARS"}
-  ggplot(icwa_delta %>% 
-           mutate(race_ethn = 
-                    ifelse(race_ethn=="other",
-                           "Non-AIAN",
-                           race_ethn)),
-         aes(y = state, x = adopted, fill = period)) + 
-    geom_col(position = position_dodge2(reverse = T)) + 
-    facet_wrap(~race_ethn, scales = "free") + 
-    scale_fill_brewer(palette = "Dark2") + 
-    labs(y = "",
-         x = "Children in adoption",
-         fill = "")
-
-#### make state-level proportional change over time 
-### for both outcomes
-```
-
-Figure \@ref(fig:icwa-pct) shows the relative change in adoption and foster care caseloads between the mid-1970s and 2019 for these states.
-Of the 19 states, 14 saw growth in the numbers of Native children in foster care since the passage of ICWA.
-Of the 13 states with complete adoption data, 8 had more Native children in foster care in 2019 than in the mid-1970s prior to the passage of ICWA.
-For non-Native children, foster care caseloads were higher in 16 of 19 states and adoption caseloads were higher in 6 of the 13 states with complete data.
-Oklahoma saw the largest absolute growth in Native foster care caseloads.
-In the mid-1970s, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 1976) %>% select(fc) ` Native children in foster care in Oklahoma.
-In 2019, there were `r icwa_delta %>% filter(state=="OK", race_ethn=="AIAN", period == 2019) %>% select(fc) ` Native children in foster care in Oklahoma, a growth of nearly `r round(icwa_delta_pct %>% filter(state=="OK", name == "Foster care", race_ethn == "AIAN") %>% select(value), 0)` percent.
-
-```{r icwa-pct, fig.cap = "Percent change in adoption and foster care caseloads between 1976 and 2019. Data from AAIA surveys and AFCARS"}
-
-ggplot(icwa_delta_pct %>% 
-         mutate(race_ethn =
-                  ifelse(race_ethn == "AIAN",
-                         race_ethn,
-                         "Non-AIAN")),
-       aes(x = value,
-           y = state,
-           fill = name)) + 
-  geom_col(position = position_dodge2(reverse = T)) + 
-  labs(y = "") +
-  geom_vline(xintercept = 0, lty = 2) +
-  scale_fill_brewer(palette = "Dark2") + 
-  facet_wrap(~race_ethn, scales = "free") + 
-  labs(fill = "", 
-       x = "Percent increase, 1976 to 2019")
-```
-
-```{r icwa-growth}
-#COMPUTE GROWTH IN AIAN - GROWTH IN NONAIAN FOR WHICH GREW FASTER.
-icwa_delta_aian<-icwa_delta_pct %>% 
-  filter(race_ethn=="AIAN") %>% 
-  select(-race_ethn) %>% 
-  rename(aian_value = value)
-icwa_delta_non<-icwa_delta_pct %>% 
-  filter(race_ethn!="AIAN") %>% 
-  select(-race_ethn) %>% 
-  rename(non_aian_value = value)
-
-icwa_pct_diff<-icwa_delta_aian %>% 
-  left_join(icwa_delta_non) %>% 
-  mutate(diff = aian_value - non_aian_value)
-
-tab_higher<-icwa_pct_diff %>% 
-  group_by(name) %>% 
-  summarise(aian_higher = sum(diff>0, na.rm=T))
-```
-
-Foster care caseloads grew in many states for both non-AIAN and AIAN children.
-Of the 19 states included here, 12 saw higher growth in AIAN foster care caseloads than they saw in non-AIAN foster care caseloads.
-Adoption caseloads also grew faster for AIAN children in these states than they did for non-AIAN children.
-In 7 of 13 states with complete data, AIAN adoption caseloads grew faster between the mid-1970s and 2019 than did non-AIAN adoption caseloads.
-
-Of course, these data do not provide a complete comparative picture of Indian family separation over time.
-Until the late 20th century, American Indian and Alaska Native children were routinely removed from their families and sent to residential boarding schools.
-Approximately 25000 Native children were in Bureau of Indian Affairs boarding schools in 1973.
-This total is larger than the total number of Native children in foster care or adoption recorded by AAIA in 1976.
-There were more AIAN children in boarding schools in 1973 than were in foster care in 2019.
-While the scale of separation of Native children from their families through foster care and adoption has increased over time in these states, the absolute scale of the separation of Native children from their families has declined nationally with the end of the boarding school era.
 
@@
 
+ggsave("./vis/cjlr/fig2.png")
+
 ```
@@
   geom_line() + 
-  geom_ribbon(alpha = 0.5, color = NA) + 
+  #geom_ribbon(alpha = 0.5, color = NA) + 
   facet_geo(~state) + 
@@
   geom_line() + 
-  geom_ribbon(alpha = 0.5,
-              color = NA) + 
+  # geom_ribbon(alpha = 0.5,
+  #             color = NA) + 
   labs(fill = "", color = "",
@@
   geom_line() + 
-  geom_ribbon(alpha = 0.5,
-              color = NA) + 
+  #geom_ribbon(alpha = 0.5,
+  #            color = NA) + 
   labs(fill = "", color = "",
@@
   geom_line() +
-  geom_ribbon(alpha = 0.5, color = NA) + 
+  #geom_ribbon(alpha = 0.5, color = NA) + 
   facet_geo(~state) + 
@@
 
-This report provides two sets of empirical findings.
-
-First, we examine data on the breadth of the separation of AIAN children prior to the passage of the Indian Child Welfare Act, and compare these data to contemporary AIAN family separation through the child welfare system.
-We find that, while the overall scale of AIAN family separation has decreased since ICWA due to the closing of BIA boarding schools, the scale of AIAN family separation through fostering and adoption has increased in the states for which we have comparable data.
-In some states, like Oklahoma, Washington, California, and Alaska, the frequency of AIAN foster care and adoption has increased dramatically since the mid-1970s.
-
-However, these results should be interpreted with several limitations in mind.
-Because of changes in the ways that the U.S.
-Census tabulates AIAN populations, and because of cultural shifts in how AIAN people identify themselves when asked questions about racial identity on surveys like the Census @lujan_american_2014, estimates of the size of the AIAN child population are not comparable between 1976 and 2014 - 2018.
-As such, we cannot directly compare population adjusted incidence rates, only raw counts.
-Nonetheless, the magnitude of growth in the frequency of AIAN foster care and adoption in many states has outpaced the kinds of growth we would expect from demographic processes alone, and suggests that the scope of AIAN fostering and adoption has indeed grown over this period.
-Future research should explore methods for normalizing population counts across these two periods.
-
-We should also consider AAIA's counts of Native children in foster care and adoption in the pre-ICWA period to be conservative.
-AAIA relied on agency surveys to count AIAN children that had been separated from their families, and limited survey non-response negatively biases these counts.
-Additionally, many state and local agencies did not collect data on a child's racial or ethnic identity during this period.
-In some states, AAIA imputed the total number of AIAN children in foster care and adoption by applying the state or local population composition to the total count of children in foster care or adoption.
-Assuming any disparity in placement (a reasonable assumption), these imputations would be negatively biased, as AAIA suggests in their report.
-As such, the pre-ICWA placement numbers are likely conservative estimates of the scale of AIAN family separation in 1976.
-Because there is likely no such bias in contemporary numbers, estimates of proportional change (as displayed in Figure \@ref(fig:icwa-pct)) may overstate the magnitude of change since ICWA.
-
-Second, we examine data on the contemporary contact of AIAN children with the child welfare system at five distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights.
+We examine data on the contemporary contact of AIAN children with the child welfare system at five distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights.
+Our findings reveal similar levels of national risk for AIAN children documented in other studies [@wildeman_cumulative_2020; @yi_cumulative_2020].
 We find that AIAN children are more likely than white children to be placed into foster care, even when rates of investigation and substantiation of cases are similar to those for white children.

### 2021-08-22 state_ms.Rmd
add revised state pdf

@@
 We find that risks of child welfare system contact are highest for AIAN infants, and that risk is highly variable across states.
-Our estimates show that in high risk states at 2014 - 2018 levels of risk, more than half of AIAN children will ever be investigated by a child welfare agency, more than one in five will experience a substantiated maltreatment case, and more than one in five will ever enter foster care.
-We further find that child welfare agency case processing exacerbates inequality.
-AIAN children are more likely than white children to enter foster care, conditional on experiencing a substantiated maltreatment case, than are white children.
 These exceptionally high levels of risk indicate that the crisis of Indian family separation is ongoing.
@@
   facet_wrap(~var)
-
-ggsave("./vis/cjlr/fig2.png")
 
